### 2018-08-27 README.Rmd
Adding project compendium files Take 2

---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# klgo.cws.assessment

This repository contains the data and code for our paper:

Reynolds, Joel H., Ward, Madeleine (20XX). _An assessment of the Coastal Waterbird monitoring project at Klondike Goldrush National Historical Park_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

Reynolds, Joel H., Ward, Madeleine (20XX). _An assessment of the Coastal Waterbird monitoring project at Klondike Goldrush National Historical Park_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

Reynolds, Joel H., Ward, Madeleine (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for 'An assessment of the Coastal Waterbird monitoring project at Klondike Goldrush National Historical Park'_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: </archive/master.zip>

Or you can install this compendium as an R package, klgo.cws.assessment, from GitHub with:
### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2018-08-27 analysis/paper/paper.Rmd
Adding project compendium files Take 2

---
title: "Title Goes Here"
author:
  - Author One
  - Author Two
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

library(klgo.cws.assessment) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

# Introduction

Here is a citation [@Marwick2017]

# Background

```{r demo, eval = FALSE}
1 + 1
```

# Methods

# Results

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- readr::read_csv("../data/raw_data/my_csv_file.csv")
```

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository("../..")
```

### 2018-09-11 analysis/paper/paper.Rmd
Begin developing the report markdown

@@
 ---
-title: "Title Goes Here"
+title: "Assessment of the KLGO Coastal Waterbird Survey Monitoring Program"
 author:
-  - Author One
-  - Author Two
+  - name: Joel H. Reynolds^[U.S. National Park Service, Alaska Region, Natural Resources Science Team]
+  - name: Madeleine Ward^[Swathmore College, Dept. of Mathematics and Statistics]
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: 
-    bookdown::html_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+output: word_document
+
 abstract: |
-  Text of abstract
+The Coastal Waterbird Survey has been conducted, with varying levels of effort and in-season frequency, since 2003 (CITE).
+The area of interest has been divided into eight contiguous survey units; on each survey, at each unit the number, age, composition (sex), breeding status, and noteworthy behavioral observations are recorded for each species (KLGO 2009b).
+
+Given the constraints on personnel, time, and budgets, it is imperative that surveys be conducted efficiently in terms of both choice of sampling design and choice of sampling effort.
+Now that KLGO has multiple years of experience implementing the protocol, it is timely to step back and assess how things are going.
+We review the current survey program’s objectives,  design, identify statistical issues, and recommend potential solutions.
+Major topics include ….
+The study concludes with a series of recommended tasks that the park should undertake to improve the coastal waterbird survey’s efficiency and effectiveness.
+
 keywords: |
@@
 
-```{r, setup, echo = FALSE}
+```{r setup, include=FALSE}
 knitr::opts_chunk$set(
@@
 # Introduction
+Since 2003, staff at Klondike Gold Rush National Historical Park (KLGO) have surveyed coastal waterbird species distribution and abundance multiple times each year from late April through late September (Coastal Waterbird Survey or ‘CWS’; [@Surdyk2018]).
+This is sufficient history to warrant stepping back and assessing the monitoring program.
+Such assessments are key to an adaptive monitoring program ([@Lindenmeyer 2009]) and are an important and distinct phase to promote program, and institutional, learning and evolution ([@Reynolds2016]).
 
-Here is a citation [@Marwick2017]
+Assessing a monitoring program focuses on four broad topics  following the major Phases of a monitoring program (ibid): 
+(i) why - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?
+(ii) how - [Data Collection Design] Does the monitoring data collection design effectively addresses the information objectives?
+Given interim learning regarding system characteristics and improvements in measurement technologies, are the information goals (including accuracy and precision goals), actually feasible?
+If so, can the data collection be made more efficient, reducing the observation burden on staff resources?
+(iii) how - [Data Analysis Design] Do the data analysis methods effectively address the information objectives? Are more informative analyses available given interim improvements in methods? 
+(iv) how - [Workflows & Reporting] Are workflows and related software systems in place to allow timely, and efficient, data management, analysis updating, reporting, and distribution of findings to target audiences of information users?
+
+This assessment addressed all four topics, but focused predominantly on topics (i) Problem Framing & Information Objectives, (ii) Data Collection Design, and (iii) Data Analysis Design.
+Since the former topic (Problem Framing & Information Objectives) determines the relevancy of many of the potential assessment actions related to the latter topics (Data Collection and Analysis Designs, Workflows & Reporting), therefore  the assessment was drafted as a first stage of an iterative effort, providing high level feedback, including listings of potential detailed syntheses and workflow enhancement activities (such as improved database interface, automation of annual analyses and updating of syntheses, etc.).
+With this foundation, further dialogue with KLGO staff regarding relevant Problem Framing & Information Objectives and other priorities will determine which follow-on activities to pursue.
+
+<!-- relation to existing NPS I&M methods & formats? -->
+
+<!-- Key Findings / Next Steps -->
 
@@
 # Methods
+This assessment compared the CWS monitoring program against a recommended checklist of monitoring program components (Appendix 1).
+The checklist was developed from published literature (Reynolds 2012; Reynolds et al.
+2016).
+It identifies multiple component decisions associated with each of the major Phases of a monitoring program (briefly overviewed above).
+For each component, the relevant aspects of the current CWS were reviewed, potential issues identified, potentially informative diagnostic analyses suggested, and recommendations provided.
+The characterization of the CWS was based on KLGO protocols and reports (Table 1), a site visit (6/4-6/2018) and personal communications with the KLGO Natural Resources Program Manager (Jami Belt, winter and spring 2018), KLGO Resources Manager (Anne Matsov, summer 2018) and long-standing survey participants (Elaine Furbish, Skagway Bird Club, spring and summer 2018).
 
-# Results
+# Assessment Results
+## Phase 1 - Problem Framing and Information Objectives
+*WHY* monitor? What potential decisions and issues should the observations inform?
+
+The management problem or question motivating the need for information about coastal waterbirds provides the foundation for the survey’s data collection and analysis designs (Fancy et al.
+2009).
+It clarifies the underlying goal of the monitoring program and answers “Why is this an important problem or question?”.
+
+A robust problem definition should include, among other aspects, the temporal and geographic scope of the problem; who has the authority to make decisions that could resolve or address the problem; what information about the resource is needed to improve the decision making (the information needs); and who are the stakeholders that are impacted by the decision and/or interested in the monitoring information.
+Stakeholder identification should include interested parties and their information needs in both near-term - for example, park interpretation and visitor education activities, as well as long-term - for example, regional-scale changes in migration patterns or assessment and improvement of projected climate change impacts on avifauna (e.g., Wu et al.
+2018).
+A good understanding of multiple information needs and pathways to influence decision making will guide the required data analyses and reporting and outreach.
+
+The information needs are what the monitoring effort aims to address (Reynolds 2012). 
+Even when the focus is simply the status and trend of a resource, there is an unstated expectation that someone will endeavor to achieve or maintain a desirable state or reverse an undesirable trend (Reynolds et al.
+2016).
+Whomever has that responsibility and authority (person, office, or organization) is a priority stakeholder.
+They should not only receive the reports but also allowed an opportunity to clarify their information needs, including the desired precision and quality of information, the timing and format of reporting, and better understand what information is feasible to obtain (Averill et al.
+2010).
+
+### Assessment findings:
+The 2003 coastal waterbird inventory and survey design were developed to address a top priority bird inventory need of the Southeast Alaska Inventory and Monitoring Network (Sharman et al.
+2000) - namely, an inventory of waterfowl/shorebirds at KLGO (Hahr and Trapp 2004).
+This was deemed a priority due to the area’s expected avifaunal richness (NABCI 2000) and the limited availability of reliable observation records - mainly consisting of systematic surveys conducted by NPS staff in the early 1980s (Hahr and Trapp 2004).
+
+The goal of the inventory was to document the occurrence of 90% of the species of waterbirds likely to occur in KLGO (ibid).
+The principal objectives were “ to develop scientifically valid sampling designs for inventorying waterbirds and breeding landbirds in KLGO, and to collect baseline information on bird species occurrence, distribution, abundance, and habitat associations in the park…[to] ...provide park managers with timely information that facilitates land management planning, helps track wildlife population and ecosystem change over time, and keeps resource managers apprised of changes that may require action before a crisis occurs” (ibid).
+
+The coastal waterbird surveys have been continued since 2003 (Surdyk and XXXX 2018)
+
+Hahr and Trapp (2004) Quotes:
+upper Taiya Inlet
+The park contains excellent breeding and migration habitat for landbirds and waterbirds, most notably the Taiya River riparian Sitka spruce-cottonwood forests and the Taiya River estuary.
+
+We recommend that bird species distributions, habitat associations, and abundances continue to be monitored in the park’s regionally and internationally important bird habitats, and that KLGO continue to contribute to the ALMS effort.
+
+The park should formalize a waterbird migration monitoring protocol for the Taiya River estuary and seek assistance from local area birders to expand this effort beyond the park boundary in order to encompass more of the upper Taiya Inlet.
+Many of the undocumented bird species on the park’s expected species list are waterbirds, most notably shorebirds.
+Monitoring of spring waterbird migration, especially during the brief period when shorebirds are present, is highly
+recommended.
+Adding additional bird species to the park’s list will be difficult because the remaining undocumented species are apparently uncommon, may occur in low numbers, or appear irregularly or accidentally in the park.
+
+There are no federally listed Threatened or Endangered bird species in KLGO; however, nine Alaska WatchList species have been documented in the park (Table 4).
+The WatchList identifies at-risk Alaskan bird populations that may be facing population decline, limited geographic range, habitat loss and other threats but are not already protected under the federal Endangered Species Act.
+
+Investigations into the breeding ecology, phenology, reproductive success, and population age structure of KLGO breeding birds would also yield valuable information on the status of these species, especially those identified on the Alaska WatchList.
+We intend to use the results of these inventories to create a revised bird checklist and an atlas database for the Taiya and Skagway River watersheds in collaboration with the Skagway Bird Club.
+The forthcoming Vital Signs Monitoring Program will be a great opportunity for KLGO natural resource managers to review the park’s nascent landbird inventory and monitoring program and determine its future direction.
+
+2014 St. Pierre (2015?) 
++ new inventory additions
++ spp composition
+
+2015 Surdyk (2016?)
+Long-term monitoring and baseline inventories help managers to make decisions that meet the National Park Service goal to maintain intact ecosystems in their natural state (Hahr and Trapp 2004).
+
+Timing of peak spring waterbird migration to coincide with eulochon run.
+
+2016 Surdyk & ??? (2017)
+better understand the relationship between waterbird migration and spring eulachon runs.
+
+This protocol addresses one monitoring question with three objectives: (1) what is the annual nest density and reproductive (fledgling) success in known Hawaiian petrel colonies, (2) what are the long-term trends in colony distribution and density monitored in approximate 5-year intervals, and (3) are these affected by predator control?
+The first goal of monitoring is to obtain unbiased estimates of Hawaiian petrel nest density and reproductive (fledging) success from known colonies in HAVO and HALE in order to detect changes in colony growth or decline.
+The second goal of monitoring is to periodically (approximately every five years) obtain unbiased estimates of nest density from potential Hawaiian petrel habitat.
+This information can be used to assess changes in density and distribution of subcolonies across the landscape.
+The third goal is to estimate nest density and fledging success in areas undergoing different management regimes to assess effectiveness of management.
+Benchmark levels of these estimates could serve as warnings of the need for modified management or further investigation of these colonies.
+
+Though not explicitly characterized in the protocol or reports reviewed for this assessment (Table 1), the following target audiences and information users were identified through conversations with KLGO staff, partners, and other waterbird researchers and stakeholders in Alaska.
+
+Recommendations:
+
+Potential Follow Up Activities:
+
+Define the Target Audiences for the information and their Objectives in using the information
+Decision Makers
+Who are target audiences for this information?
+Who uses it or may use it?
+Local contact: C. Elaine Furbish, Skagway Bird Club, Google Site
+Predominantly historical park; limited resource staff and capacity
+Visible natural resource of interest to local community and visitors; interp and educational opportunity; topic of information requests; provides ‘community service’ given absence of any other agency with natural resource staff;
+Annual report (time lag?); National Resource Condition Assessment; other? State of the Park?
+Potential for contribution to regional, flyway, and WASO programs (e.g., CCRP & Audubon climate watch, etc.)
+Current sightings passed to local bird club & used by avian-interested visitors
+Used in quarterly regional ‘significant bird observations’, which in turn is used to develop the Alaska-wide summary for the journal North American Birds (published by the American Birding Association http://www.aba.org/nab ) and includes ‘...sightings of birds that are out of range or out of season or occur regularly in small numbers, noteworthy breeding records, unusually large or small numbers of a particular species, unusual migration dates, etc….’ (S.
+C.
+Heinl, 2017, Summary of SE AK Bird Observations, June-July 2017, accessed 2018 June 6).
+Queries to Heather Renner & Robin Corcoran re: PSG-type users, etc.
+Opportunity for community based monitoring to track projected spp impacts of climate change, change in phenology, occurrence, abundance, etc.?
+
+Decisions
+What types of decisions is it / will it be used for?
+(specify for each target audience/ end user)
+No specific decisions
+Current:
+KLGO - spp lists, phenology of occurrence (guidance to visitors), trends in richness (NRCA), abundance?
+Skagway bird club - spp list, current occurrences (activity re: rare & unusual)
+SE AK / Regional - spp lists; unusual sightings, etc. (spp occurrences & broad seasonal timing).
+AKR -
+WASO/NPS - ??CCRP??
+Question to Gregor & Mel (AK Audubon).
+Could be used for garnering community engagement in Climate Watch & other national efforts associated with Audubon’s Climate Report.
+Not used in continental analyses.
+“One key message that I want to emphasize is that the focus of this work was purely on climate suitability.
+What this means, as we indicate in the briefs’ “Important” box up front and “Caveats” section further back, is that significant changes in climate suitability, as measured here, will not always result in a species response, and all projections should be interpreted as potential trends.
+Multiple other factors mediate responses to climate change, including habitat availability, ecological processes and biotic interactions, dispersal capacity, species' evolutionary adaptive capacity, and phenotypic plasticity (e.g., behavioral adjustments).
+Ultimately, models can tell us where to focus our concern and which species are most likely to be affected, but monitoring is the only way to validate these projections and should inform any on-the-ground conservation action.” Gregor Schuurman, email to DENA
+
+Potential:
+KLGO: 
+(easy) - changes in occurrence (spp, season scale); changes in phenology (migration timing, by spp; breeding timing, by spp); broad (order of magnitude) changes in abundance (index); broad (?) changes in composition (by spp & timing of survey)?; species occurrence x unit (~habitat?) - refine expected lists per survey and develop targeted SOP training materials;
+(harder) - finer changes in abundance; finer changes in composition and age; ???
+
+What Objectives are the decision makers hoping to achieve by using this information?
+
+Over what spatial scales is the information being used? 
+e.g., at the park, in the community, regional scale, flyway scale, nationally, internationally
+For each spatial scale of interest, what are the associated time scales of interest for detecting changes?
+E.g., annual, w/in season, spring vs fall, etc.
+
+From another report but quotes from Olsen et al. (1999):
+Olsen (et al.
+1999) noted that “Most of the thought that goes into a monitoring program should occur at this preliminary planning stage.
+The objectives guide, if not completely determine the scope of inference of the study and the data collected, both of which are crucial for attaining the stated objectives.” Olsen goes on to say that a “clear and concise statement of monitoring objectives is essential to realize the necessary compromises, select appropriate locations for inclusion in the study, take relevant and meaningful measurements at these locations, and perform analyses that will provide a basis for the conclusions necessary for meeting the stated objectives.”
+
+Fancy, Gross, and Carter (2009):
+The development of monitoring objectives, which provide additional focus about the purpose or desired outcome of the monitoring effort, was an iterative process that sometimes required several years to refine.
+Early in the design process, monitoring objectives were stated in more general terms, such as “Determine trends in the incidence of disease and infestation in selected plant communities and populations”, whereas the final monitoring plan and protocols provided monitoring objectives that met the test of being realistic, specific, and measurable (e.g., “Estimate trends in the proportion, severity, and survivorship of limber pine trees infected with white pine blister rust at Craters of the Moon National Monument”; Garrett et al.
+2007).
+
+Structure from Northeast Coastal and Barrier Network
+https://irma.nps.gov/DataStore/Reference/Profile/2195238
+Justification => Vital Sign => Monitoring Goal => Monitoring Questions => Monitoring Objectives => Measures (eg pg 55 & 56)
+
+Clarify Objectives of the Decision Makers 
+Among the decision maker objectives identified in 1.b.ii, clarify 
+the Fundamental Objective(s) from 
+the Means Objectives, then
+diagram the Objectives Hierarchy.
+Identify the Attributes that need to be monitored to assess achievement of each of the objectives. (‘Concepts by Theory’, Ford 2001)
+Identify the Measurements selected to characterize each attribute. (‘Concepts by Data’, Ford 2001) [WHAT]
+Identify any Threshold values associated w/ each Measurement
+
+Q: compiling and assessing regional trends in waterbirds in SE AK?
+Asked of Heather Renner (Ak Maritime), Robin Corcoran (Kodiak NWR)
+Answers:
+Not really; regular discussion at PSG but no one has taken it on. 
+Kathy Kuletz works on statewide trends in at sea seabird observations;  Piatt GLBA work (& Martin Renner); 
+Robin C: “There's always talk at the PSG meeting about the North Pacific seabird colony database, or some variation of that database for marine birds, but it never gets anywhere.
+Basically there is no money for a database manager.
+I've been submitting my data to Rob Kaler in Migratory Bird Management, but I think that's just for storage until the money for a database manager materializes.
+More recently Rob Suryan w/NOAA has been interested - for the North Pacific Pelagic Seabird Database.
+You might want to contact him, he's leading the Gulf Watch Alaska Long-Term Ecosystem Monitoring Program, and maybe interested in this work.”
+
+NPS/Audubon:
+Wu et al.
+2018.
+Projected avifaunal responses to climate change across the U.S.
+NPS.
+PLOS ONE.
+Appendices include projected park-specific turnover, colonization, and extirpations projections - potential to test with this survey.
+Would require discussions w/ CCRP & ARO - ideally as a coordinated effort with, say, AK Audubon and ACF(?) or LCCs as tracking metric.
+Appendices include “Potential management goals and activities for parks, organized by trend group.”
+
+Melanie AK Audubon: IBA requires demonstrating higher than ‘usual’ densities (relative to surrounding area).
+
+Nicole Michel & Brooke Batemen Audubon National Science Team: of interest and possible use in national scale quantitative analyses or climate species distribution models.
+“Scaling up and integrating datasets across regions, including data from different goals, is something that we’re starting to do in other regions.
+“
+
+Occurrence, Richness
+Phenology (migration, breeding)
+Abundance - Even without the structure needed to estimate detection probability, if the same protocol is used annually this data could be used to produce long-term trend indices for the study area.
+(competing staff priorities - peak (?) usually aligns with hooligan run. Possibly reduce summer frequency?)
+
+Diagram the Conceptual Model of the system
+E.g., create an influence diagram, being sure to denote the
+Fundamental & Means Objective(s)
+Factors influencing their achievement (aka ‘system drivers’)
+Associated management or policy decisions and/or potential actions.
+Include a written summary of the literature upon which the conceptual model is based.
+
+Summarize the management or policy decisions the monitoring is to inform (if any) based on 3.c.
+
+Phase 2 - Monitoring & Analysis Design
+HOW monitor? Clarify the various design choices.
+
+Clarify rationale for monitoring versus some other means of addressing the information objectives (e.g., literature review, controlled experiment, etc.), and type of monitoring being undertaken (status & trends or surveillance, threshold, effectiveness, or adaptive management).
+
+Questions: Why? What are information goals & objectives?
+What sort of observations would trigger action? What actions?
+
+Survey Design 
+Species: 
+Targets: loons, grebes, cormorants, herons, ducks, geese, plovers, sandpipers, gull, alcids, [kingfishers]
+Questions: clarify role of kingfishers.
+
+Incidental: 48 other spp
+Questions: are these all migratory? Any year-round residents? Any of special concern? 
+Any especially problematic for detection on water or identification of spp?
+
+Target Frame - space, time, scale, target populations
+Spatial: Upper Lynn Canal visible from coastline from Skagway to Dyea, including Skagway & Taiya river mouths and Nahku Bay (between Skagway and Dyea Point)
+
+Hahr & Trapp (2004), pg 13: Areas of coastline that were visible from public access points
+between Skagway and the Taiya River delta were divided into eight census units
+and surveyed once per week in late April and May and bi-weekly during late
+summer and fall (August – mid September) (Figure 5). Surveys were conducted
+as weather and scheduling permitted and dates/time were not selected at
+random. Standardized observation points were established for each census unit
+to ensure the most complete coverage of the study area, and to minimize
+observer bias.
+Observers counted and identified all birds visible within each unit and
+documented birds opportunistically observed while moving between units.
+Counts were made by scanning the census units with Bushnell Legend 10 x 42
+binoculars and a Bausch & Lomb 20-60 x 70 spotting scope until the observer
+was able to quantify all visible birds in the unit. Surveys were only conducted
+under optimal weather conditions; however, occasionally conditions deteriorated
+during the course of a survey making species identification difficult (especially in
+the case of low clouds, fog and wind). Data were recorded on standardized data
+forms developed after Collins et al. (2001) (Appendix B). ArcView shapefiles of
+the observation points and census units were created for use in the park’s GIS.
+All data, original data forms, and field notes were archived at KLGO.
+
+Temporal: ??how define ‘season’??
+Questions: what is cue to start of survey season?
+Diagnostic analysis => check initial counts x spp
+
+Sample Frame - 
+Spatial: same coverage as Target Frame. Area split into eight survey areas (see Hahr & Trapp 2004).
+Questions: Access to all units? Ease of observation of different units? Ease of distinguishing boundaries?
+Diagnostic analysis => value of all units - only some spp in unit X?; “unit by species occurrence” relationships?
+Sample Selection process - any strata or clusters? Two-stage processes? Probabilistic selection? [get in to enumerative vs analytical distinction of Deming?]
+Currently survey all of frame (‘census’ of units); no spatial sampling uncertainty. 
+All spatial uncertainty comes from Observation Process and detection issues.
+Temporal sampling uncertainty remains.
+
+Questions: feasible to reduce # of units? See Qs above about ‘unique contributions’ of each unit. Logistic considerations? What effort level is feasible?
+Diagnostic analysis => any ‘representative’ units? How would that be defined?
+
+Survey timing 
+Per season -
+
+Questions: how define when to begin surveys? Zeros? Triggering events?
+Potential confounding factors? (changes in phenology, ClCh); important objective or confounder? Design options for addressing?
+Diagnostic analysis=>  Signs of confounding?
+
+W/in season - (e.g., LEAU St. George); potential refinement?
+Temporal: 
+(migration) 1 per wk, Apr 27 - May 20; (breeding) 1 per 2 wks, Jun 6 - Jul 18; (migration) 1 per wk, Aug 1 - Sept 28.
+2016: 17 surveys total
+ Questions: How define beginning & end of breeding period?
+Info goals from migration periods? From breeding periods? 
+Design options to reduce frequency?
+Diagnostic analysis=> 
+.assessment of impact on priority info goals of different frequencies?
+.spp specific occurrency by survey date across years - capturing onset and end of migration?
+.other assessments of survey timing relative to phenology of spp
+.mine FB, Eagle Chat, others for date of first occurrences (see 2016 report); Elaine. Requires clarifying information goals of survey vrs ‘first occurrence’. E.g., not tail behavior.
+
+Per survey
+route sequencing and tidal influence; 
+SOP: 1,2,3,4,5,6,8,7 or reverse, depending on tide timing (determines when do unit 7).
+
+Questions: design options for confounding factors?
+Duration across eight units: feasible to pull off? 
+issue of bird movement between units? Anyway to bound potential bias?
+Design options?
+
+Monitoring Design
+Membership Design (panels, etc.)
+Single Panel
+
+Revisitation Design
+W/in Year; varying effort levels (strata in time across seasons)
+Between Years
+Always Repeat (pure panel), every year
+Questions: feasbility of not doing every year? What are information goals and triggers?
+
+Response Design 
+Attributes & Measurements
+Focal targets vrs all species
+Priority Spp for region?
+PSG?
+At Risk Spp (CCRP?)
+Audubon AK spp of concern? See Audubon Climate Report. Relevant spp (intersection with spp list)
+
+Observation Process
+Question: 
+How determine boundaries? Visual markers.
+How handle movement of birds into and out of unit?
+Check SOPs but basically ‘until things stop’.
+Problem is that that is an accumulating approach that will bias high areas with lots of productive habitat and/or well timed relative to changing tidal conditions and sequencing of survey.
+How handle tendency to record everything of note: e.g., recording observations in neighboring units when really supposed to be focused on this unit.
+E.g., variation in time observing and effort observing across units & sampling events.
+This fails to track actual observation activity (time) devoted to each unit as well as preferential bias to overestimate use in areas.
+Think about Forsell & Zweifelhofer?
+Chat w/ Emily & Mark Otto & Andy?
+Covariates - captured & not captured. Value, use in analyses; more reliable alternative data sources; alignment of spatial scales of interest and time scales of interest.
+Lack of control over time at each site; portion spent traveling between observation points, variation in number of observation points for each site.
+Variation in survey effort across sites, across per unit area, across surveys (days), etc.
+Major weather conditions (Northerlies, Southerlies) - simplify conditions to these
+Unit 1 - recording and impact of presence & # of cruise ships in dock, helicopter tours, etc., etc., other human disturbances.
+Distinguish value of covariates for act of (i) presence/absence assessment, (ii) abundance count, (iii) composition & breeding state.
+Check SOP to clarify categorization of conditions.
+Temp and wind greatly burden data collection process (slowing down greatly).
+Variation in observer #, training, and skill across surveys, across years
+
+Diagnostic analyses =>
+.distribution of time duration at each site by date and year
+.correlation across units?
+.average surface area per unit
+.something about sequencing of observation sites (e.g., confounding with tidal state)?
+.distribution of observers by survey date by year
+.distribution of observer experience(?)  by survey date by year?
+
+Detection issues associated with survey conditions
+.species specific counts w/in season by weather conditions and wave conditions
+.species occurrence by wave conditions & weather conditions
+.species age and composition by wave conditions & weather conditions 
+.total # species by weather conditions, wave conditions; internally consistent contrasts to demonstrate value, or lack thereof, of current covariates.
+.richness by major weather conditions (Northerlies, Southerlies, etc.) and for each spp
+.distribution of weather conditions
+.distribution of survey time by weather conditions
+.diagnostic to ID key survey times to ensure high quality observers?
+.consistency in ‘breeding periods’ across years (as captured in data)
+.quick diagnostics to ID covariates seldom recorded, not consistently recorded, consistently valued (e.g., no basis for contrast), etc.
+
+Value of site-specific covariate recordings vs available ‘local’ weather (Skagway weather observations from airport - wind direction & speed, temp, etc.)
+
+Ppt Codes: 0 - none, 1 - fog, 2 - drizzle, 3 - showers, 4 - light rain, 5 - moderate rain (steady), 6 - heavy rain, 7 - sleet, 8 - light snow, 9 - mod snow, 10 - heavy snow
+Beaufort scale codes: 0 - < 1 mph (air calm); 1 - 1-3 mph (smoke drift show direction of wind); 2 - 4-7 mph wind felt on face, leaves rustle; 3 - 8-12 mph leaves & small twigs in constant motion; 4 - 13-18 raises dust, loose paper, moves small branches; 5 - 19-24 small trees in leaf sway, crested wavelets form on inland waters;
+Wave height: 0 - calm; 1 - scaly ripples, no foam crests; 2 - small wavelets, crests glassy, no breaking; 3 - large wavelets, crests begin to break, scattered whitecaps; 4 - small waves, becoming longer, numerous whitecaps; 5 - moderate waves, many whitecaps, some spray
+SPP
+#
+Age (Adult, imm, juv, mixed, unknown)
+Composition - M, F, Fe/juve, Mixed, Unknown)
+Breeding Status - pairs, etcc; 24 potential codings see SOP #3.
+
+Recommendations => 
+Clearly record standardized observation locations on map and include in SOP to ensure consistency across observers and personnel.
+Issue of veg growth reducing visible portion of unit from each viewpoint - long-term confounding factor, etc.
+Value of covariates, recommendations on what to drop, what to include
+Observation quality notes to support assessment of extreme observations, etc.
+
+Detection issues (seeing and recording available birds)
+Design options for improving detection & measurement
+Hmm? Lit? Emily, Aaron C., Kathi, Mark Otto, Andy Royle?
+Feasible to do some sort of double observer process - coordinate immediately after survey?
+Only conduct intermittently per season rather than every time? E.g., like sightability functions?
+Major drivers of sightability?
+Capturing uncertainty on spp id
+Capturing uncertainty on spp count
+UAVs?
+Observer training & behavior classification
+Training materials from Kodiak or Homer or Cordova or Haines or Juneau or Yakutat or…?
+
+Species identification issues
+Priority ID concerns:
+Commonly encountered species that can be difficult
+Work with Elaine & Heather & Laura to develop list of commonly confused species based on those commonly encountered (database or Skagway Bird Club sightings)
+E.g., scoters, scaup, murrelets, etc.)
+Target Spp of interest
+Training materials or other improvements
+Db error checking?
+Other?
+Other expected sources of bias
+Species ID
+Behavior classification
+Changes in equipment
+Summary Metrics
+Confounding factors & Covariates
+
+Analysis Design
+Priority 
+Summaries of interest 
+(tie to Conceptual Model and Information Objectives)
+Associated analyses
+Reporting schedule
+Questions: Who does the analyses? Interest in setting up as Reproducible Research draft report? 
+General delay in reporting? When normally completed?
+
+Secondary
+Summaries of interest
+Associated analyses
+Reporting schedule
+
+Survey effort levels / Change detection targets / Power to detect changes
+Reiterate targets for ‘local’ inference vrs contribution to regional trends assessments
+
+1 Biotech conducts just this & amphi monitoring.
+
+Data Management & Reporting Structures / Software / Workflows
+Data Models
+Design Requirements / User reporting needs
+
+Elaine: Can put stuff in but can’t get it out of existing ACCESS db. Problematic. Bottleneck for analyses and summaries.
+Ebird - boundaries don’t include Skagway burough. Rather, formerly unincorporated area including Skagway, Haines, and Angoon - too broad, not useful.
+
+Recommendations: work with GIS shop to clean up db, improve reporting & query support; shiny app; reproducible research scripts to generate automatic reports? Shiny app to interface?
 

### 2018-09-12 analysis/paper/paper.Rmd
Just put an R in Swarthmore :)...is it supposed to be able to knit? There's an error but not sure what it means.
tiny edits to start clarifying section headers, ID initial diagnostics for Madeleine to pursue

@@
   - name: Joel H. Reynolds^[U.S. National Park Service, Alaska Region, Natural Resources Science Team]
-  - name: Madeleine Ward^[Swathmore College, Dept. of Mathematics and Statistics]
+  - name: Madeleine Ward^[Swarthmore College, Dept. of Mathematics and Statistics]
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
-Recommendations:
+### Recommendations:
 
@@
 
-Define the Target Audiences for the information and their Objectives in using the information
-Decision Makers
+Define the _Target Audiences_ for the information and their Objectives in using the information
+
+_Decision Makers_
 Who are target audiences for this information?
@@
 
-Decisions
+_Decisions_
 What types of decisions is it / will it be used for?
@@
 
-Phase 2 - Monitoring & Analysis Design
+## Phase 2 - Monitoring & Analysis Design
 HOW monitor? Clarify the various design choices.
@@
 
-Survey Design 
+### Survey Design 
 Species: 
@@
 
-Target Frame - space, time, scale, target populations
-Spatial: Upper Lynn Canal visible from coastline from Skagway to Dyea, including Skagway & Taiya river mouths and Nahku Bay (between Skagway and Dyea Point)
+*Target Frame* - space, time, scale, target populations
+*Spatial* Upper Lynn Canal visible from coastline from Skagway to Dyea, including Skagway & Taiya river mouths and Nahku Bay (between Skagway and Dyea Point)
 
@@
 
-Temporal: ??how define ‘season’??
+*Temporal* ??how define ‘season’??
+
 Questions: what is cue to start of survey season?
-Diagnostic analysis => check initial counts x spp
+*Madeleine later* Diagnostic analysis => check dates & unit of first non-zero counts each year for each spp (relative to dates of initiating surveys)
 
-Sample Frame - 
+*Sample Frame* 
 Spatial: same coverage as Target Frame. Area split into eight survey areas (see Hahr & Trapp 2004).
 Questions: Access to all units? Ease of observation of different units? Ease of distinguishing boundaries?
-Diagnostic analysis => value of all units - only some spp in unit X?; “unit by species occurrence” relationships?
-Sample Selection process - any strata or clusters? Two-stage processes? Probabilistic selection? [get in to enumerative vs analytical distinction of Deming?]
+*Madeleine later* Diagnostic analysis => value of all units - only some spp in unit X?; “unit by species occurrence” relationships?
+
+*Sample Selection process* - any strata or clusters? Two-stage processes? Probabilistic selection? [get in to enumerative vs analytical distinction of Deming?]
 Currently survey all of frame (‘census’ of units); no spatial sampling uncertainty. 
@@
 
-Survey timing 
+*Survey timing* 
 Per season -
@@
 
-Observation Process
+###Observation Process
 Question: 
@@
 Chat w/ Emily & Mark Otto & Andy?
-Covariates - captured & not captured. Value, use in analyses; more reliable alternative data sources; alignment of spatial scales of interest and time scales of interest.
+#### Covariates 
+- captured & not captured. Value, use in analyses; more reliable alternative data sources; alignment of spatial scales of interest and time scales of interest.
+
 Lack of control over time at each site; portion spent traveling between observation points, variation in number of observation points for each site.
+
 Variation in survey effort across sites, across per unit area, across surveys (days), etc.
+
 Major weather conditions (Northerlies, Southerlies) - simplify conditions to these
 Unit 1 - recording and impact of presence & # of cruise ships in dock, helicopter tours, etc., etc., other human disturbances.
+
 Distinguish value of covariates for act of (i) presence/absence assessment, (ii) abundance count, (iii) composition & breeding state.
+
 Check SOP to clarify categorization of conditions.
+
 Temp and wind greatly burden data collection process (slowing down greatly).
@@
 
-Diagnostic analyses =>
+*Madeleine early* Diagnostic analyses =>
 .distribution of time duration at each site by date and year
@@
 
-Detection issues associated with survey conditions
+#### Detection issues associated with survey conditions
 .species specific counts w/in season by weather conditions and wave conditions
@@
 
-Species identification issues
+####Species identification issues
 Priority ID concerns:
@@
 
-Analysis Design
+###Analysis Design
 Priority 
@@
 
-Data Management & Reporting Structures / Software / Workflows
+### Data Management & Reporting Structures / Software / Workflows
 Data Models

### 2018-09-25 analysis/paper/paper.Rmd
Added *.bib file; working on YAML header so paper.Rmd will knit; added analysis/scripts folder.

@@
 date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: word_document
+output: 
+ word_document:
+  toc: TRUE
+#  number_sections: TRUE
+bibliography: KLGOCWS_References.bib
+link-citations: yes
 
-abstract: |
-The Coastal Waterbird Survey has been conducted, with varying levels of effort and in-season frequency, since 2003 (CITE).
-The area of interest has been divided into eight contiguous survey units; on each survey, at each unit the number, age, composition (sex), breeding status, and noteworthy behavioral observations are recorded for each species (KLGO 2009b).
-
+abstract: The Coastal Waterbird Survey has been conducted, with varying levels of effort and in-season frequency, since 2003 [@Surdyk2018].
+The area of interest has been divided into eight contiguous survey units; on each survey, at each unit the number, age, composition (sex), breeding status, and noteworthy behavioral observations are recorded for each species [@Anonymous2009].
 Given the constraints on personnel, time, and budgets, it is imperative that surveys be conducted efficiently in terms of both choice of sampling design and choice of sampling effort.
@@
 
-keywords: |
+keywords:
  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
 ---
@@
 
-library(klgo.cws.assessment) # Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
 ```
@@
 This is sufficient history to warrant stepping back and assessing the monitoring program.
-Such assessments are key to an adaptive monitoring program ([@Lindenmeyer 2009]) and are an important and distinct phase to promote program, and institutional, learning and evolution ([@Reynolds2016]).
+Such assessments are key to an adaptive monitoring program [@Lindenmeyer 2009] and are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016].
 
 Assessing a monitoring program focuses on four broad topics  following the major Phases of a monitoring program (ibid):  
-(i) why - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?
-(ii) how - [Data Collection Design] Does the monitoring data collection design effectively addresses the information objectives?
+1. _why_ - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?  
+2.
+_how_ - [Data Collection Design] Does the monitoring data collection design effectively addresses the information objectives?
 Given interim learning regarding system characteristics and improvements in measurement technologies, are the information goals (including accuracy and precision goals), actually feasible?
 If so, can the data collection be made more efficient, reducing the observation burden on staff resources?
-(iii) how - [Data Analysis Design] Do the data analysis methods effectively address the information objectives? Are more informative analyses available given interim improvements in methods? 
-(iv) how - [Workflows & Reporting] Are workflows and related software systems in place to allow timely, and efficient, data management, analysis updating, reporting, and distribution of findings to target audiences of information users?
+3. _how_ - [Data Analysis Design] Do the data analysis methods effectively address the information objectives? Are more informative analyses available given interim improvements in methods?  
+4.
+_how_ - [Workflows & Reporting] Are workflows and related software systems in place to allow timely, and efficient, data management, analysis updating, reporting, and distribution of findings to target audiences of information users?
 
@@
 
-```{r demo, eval = FALSE}
-1 + 1
-```
-
 # Methods
 This assessment compared the CWS monitoring program against a recommended checklist of monitoring program components (Appendix 1).
-The checklist was developed from published literature (Reynolds 2012; Reynolds et al.
-2016).
+The checklist was developed from published literature [@Reynolds2012; @Reynolds2016].
 It identifies multiple component decisions associated with each of the major Phases of a monitoring program (briefly overviewed above).
@@
 
-The management problem or question motivating the need for information about coastal waterbirds provides the foundation for the survey’s data collection and analysis designs (Fancy et al.
-2009).
+The management problem or question motivating the need for information about coastal waterbirds provides the foundation for the survey’s data collection and analysis designs [@Fancy2009].
 It clarifies the underlying goal of the monitoring program and answers “Why is this an important problem or question?”.
@@
 A robust problem definition should include, among other aspects, the temporal and geographic scope of the problem; who has the authority to make decisions that could resolve or address the problem; what information about the resource is needed to improve the decision making (the information needs); and who are the stakeholders that are impacted by the decision and/or interested in the monitoring information.
-Stakeholder identification should include interested parties and their information needs in both near-term - for example, park interpretation and visitor education activities, as well as long-term - for example, regional-scale changes in migration patterns or assessment and improvement of projected climate change impacts on avifauna (e.g., Wu et al.
-2018).
+Stakeholder identification should include interested parties and their information needs in both near-term - for example, park interpretation and visitor education activities, as well as long-term - for example, regional-scale changes in migration patterns or assessment and improvement of projected climate change impacts on avifauna [e.g., @Wu2018].
 A good understanding of multiple information needs and pathways to influence decision making will guide the required data analyses and reporting and outreach.
 
-The information needs are what the monitoring effort aims to address (Reynolds 2012). 
-Even when the focus is simply the status and trend of a resource, there is an unstated expectation that someone will endeavor to achieve or maintain a desirable state or reverse an undesirable trend (Reynolds et al.
-2016).
+The information needs are what the monitoring effort aims to address [@Reynolds2012]. 
+Even when the focus is simply the status and trend of a resource, there is an unstated expectation that someone will endeavor to achieve or maintain a desirable state or reverse an undesirable trend [@Reynolds2016].
 Whomever has that responsibility and authority (person, office, or organization) is a priority stakeholder.
@@
 The 2003 coastal waterbird inventory and survey design were developed to address a top priority bird inventory need of the Southeast Alaska Inventory and Monitoring Network (Sharman et al.
-2000) - namely, an inventory of waterfowl/shorebirds at KLGO (Hahr and Trapp 2004).
-This was deemed a priority due to the area’s expected avifaunal richness (NABCI 2000) and the limited availability of reliable observation records - mainly consisting of systematic surveys conducted by NPS staff in the early 1980s (Hahr and Trapp 2004).
+2000) - namely, an inventory of waterfowl/shorebirds at KLGO [@Hahr2004].
+This was deemed a priority due to the area’s expected avifaunal richness (NABCI 2000) and the limited availability of reliable observation records - mainly consisting of systematic surveys conducted by NPS staff in the early 1980s [@Hahr2004].
 
@@
 
-The coastal waterbird surveys have been continued since 2003 (Surdyk and XXXX 2018)
+The coastal waterbird surveys have been continued since 2003 [@Surdyk2018].
 
@@
 
-Hahr & Trapp (2004), pg 13: Areas of coastline that were visible from public access points
+[@Hahr2004], pg 13: Areas of coastline that were visible from public access points
 between Skagway and the Taiya River delta were divided into eight census units
@@
 *Sample Frame* 
-Spatial: same coverage as Target Frame. Area split into eight survey areas (see Hahr & Trapp 2004).
+Spatial: same coverage as Target Frame. Area split into eight survey areas [see @Hahr2004].
 Questions: Access to all units? Ease of observation of different units? Ease of distinguishing boundaries?

### 2018-09-25 analysis/scripts/Example.R

@@ -1,0 +1,1 @@
+# blah generic script file

### 2018-09-26 analysis/scripts/practice_copy.Rmd
compiled scripts that should have led to summary

---
title: "Practice"
author: "Madeleine"
date: "9/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#libraries
library(readr)
library(dbplyr)
library(dplyr)
library(odbc)
#path is local
#specify odbc as backend
#might also need to specify driver?
data_base <- DBI::dbConnect(odbc::odbc(), path = "~/klgo_bio_monitoring/analysis/data/raw_data")
```

```{r}
#name tables
dbListTables(data_base)
#all code like this gets blanks
test <- dbGetQuery(data_base, SELECT * FROM "Field_Data")

#creates data frame - can use tidyverse instead?
#here's an example
field_data <- dbReadTable(data_base, "Field_Data")
new_data <- field_data %>%
  group_by(Species, Breeding_Status) %>%
  summarize(count = n())
```

### 2018-10-03 analysis/scripts/data_retrieval.R
i think this was the branch dataRetrieval. Deleted "practice" RMarkdown, included script with UCanAccess instructions (changed specific path to general).

@@ -1,0 +1,11 @@
+#data retrieval code using JDBC
+#library: download "UCanAccess" from this site: http://ucanaccess.sourceforge.net/site.html#home
+  #upload zipped file to RStudio
+library("RJDBC")
+#create JDBC driver
+drv <- JDBC("net.ucanaccess.jdbc.UcanaccessDriver",c("~/UCanAccess-4.0.4-bin/ucanaccess-4.0.4.jar","~/UCanAccess-4.0.4-bin/lib/jackcess-2.1.11.jar","~/UCanAccess-4.0.4-bin/lib/commons-lang-2.6.jar","~/UCanAccess-4.0.4-bin/lib/commons-logging-1.1.3.jar","~/UCanAccess-4.0.4-bin/lib/hsqldb.jar"))
+#create connection
+conn <- dbConnect(drv,"jdbc:ucanaccess:///"path to file"")
+
+#check that connection has worked
+dbListTables(conn)

### 2018-10-05 analysis/paper/paper.Rmd
Renamed data_retrieval script for clarity; introduced switch for data access commands based on who is knitting markdown file; after investigation, decided better to input data from CSV files; added recommendation text to update data management workflow.

@@
 )
+
+library(dply);library(ggplot2);library(dbplyr);
 
@@
 ### Survey Design
+
+*Target Frame* - space, time, scale, target populations
+*Spatial* Upper Lynn Canal visible from coastline from Skagway to Dyea, including Skagway & Taiya river mouths and Nahku Bay (between Skagway and Dyea Point)
+
+```{r DataInput, eval=TRUE, echo=FALSE, cache=TRUE}
+if (system("git config user.email")=="mward2@swarthmore.edu") source("./scripts/data_retrievalSwarthmore.R")
+else source("./scripts/data_retrievalNPS.R")
+
+```
+
 Species: 
@@
 Any especially problematic for detection on water or identification of spp?
-
-*Target Frame* - space, time, scale, target populations
-*Spatial* Upper Lynn Canal visible from coastline from Skagway to Dyea, including Skagway & Taiya river mouths and Nahku Bay (between Skagway and Dyea Point)
 

### 2018-10-05 analysis/scripts/data_retrievalNPS.R
Renamed data_retrieval script for clarity; introduced switch for data access commands based on who is knitting markdown file; initiation of script for access db on Windows machines.

@@ -1,0 +1,28 @@
+# Rectrieve Data when working on local windows machine
+############
+# This script uses library("RJDBC") to access the local MS ACCESS mdb file
+# in the ../data/raw_data folder.
+# ASSUMES you are working on Windows machine using locally installed R and data files.
+########
+# Joel H. Reynolds, joel_reynolds@nps.gov
+# Last Edit: 2018 Oct 5
+########
+
+########## WARNING #################
+# THIS SCRIPT IS INCOMPLETE        #
+########## WARNING #################
+
+
+# add package
+library("RODBC","odbc")
+
+#create connection to the MDB file
+conn <- odbcConnectAccess("../data/raw_data/Coastal_Bird_Survey_Data.mdb")
+
+#check that connection has worked
+sqlTables(conn)
+
+
+### Using dbplyr package & odbc DBI backend
+conn <-DBI::dbConnect(odbc::odbc(), path="../data/raw_data/Coastal_Bird_Survey_Data.mdb")
+

### 2018-10-05 analysis/scripts/data_retrievalSwarthmore.R

@@ -1,11 +1,26 @@
-#data retrieval code using JDBC
+# Rectrieve Data when working on swarthmore R server
+############
+# This script uses library("RJDBC") to access the MS ACCESS mdb file
+# on Madeleine's Swarthmore account using the Rserver software.
+########
+# Madeleine Ward
+# Last Edit: 2018 Oct 4
+########
+
 #library: download "UCanAccess" from this site: http://ucanaccess.sourceforge.net/site.html#home
-  #upload zipped file to RStudio
+
+#ASSUMES YOU HAVE ALREADY uploaded zipped database file to RStudio
+
+# add package
 library("RJDBC")
+
 #create JDBC driver
 drv <- JDBC("net.ucanaccess.jdbc.UcanaccessDriver",c("~/UCanAccess-4.0.4-bin/ucanaccess-4.0.4.jar","~/UCanAccess-4.0.4-bin/lib/jackcess-2.1.11.jar","~/UCanAccess-4.0.4-bin/lib/commons-lang-2.6.jar","~/UCanAccess-4.0.4-bin/lib/commons-logging-1.1.3.jar","~/UCanAccess-4.0.4-bin/lib/hsqldb.jar"))
-#create connection
+
+#create connection to the MDB file
 conn <- dbConnect(drv,"jdbc:ucanaccess:///"path to file"")
 
 #check that connection has worked
 dbListTables(conn)
+
+

### 2018-10-09 analysis/paper/paper.Rmd
Moved CSVs from derived_data to raw_data/CSVs

@@
 ```{r DataInput, eval=TRUE, echo=FALSE, cache=TRUE}
-if (system("git config user.email")=="mward2@swarthmore.edu") source("./scripts/data_retrievalSwarthmore.R")
-else source("./scripts/data_retrievalNPS.R")
+# if (system("git config user.email")=="mward2@swarthmore.edu") source("./scripts/data_retrievalSwarthmore.R")
+# else source("./scripts/data_retrievalNPS.R")
 
+###########
+# MADELEINE WILL REPLACE WITH CODE TO VARIOUS TABLES FROM CSV FILES
+###########
 ```
@@
 
-Recommendations: work with GIS shop to clean up db, improve reporting & query support; shiny app; reproducible research scripts to generate automatic reports? Shiny app to interface?
+The current database is written in Microsoft Access version XX for 32-bit operating systems.
+Software and machine operating systems have advanced enough for this to cause unnecessary burdens for accessing and importing the data for analysis and reporting in common statistical analysis environments (e.g., R).
+Recommend the Park request the ARO GIS shop to convert the database to an uptodate software.
+
+*JOEL: Talk to Angie about what this would entail.*
+
+_Recommendations_ work with GIS shop to clean up db, improve reporting & query support; shiny app; reproducible research scripts to generate automatic reports? Shiny app to interface?
 

### 2018-10-09 analysis/scripts/data_retrievalNPS.R

@@ -1,23 +1,57 @@
 # Rectrieve Data when working on local windows machine
 ############
-# This script uses library("RJDBC") to access the local MS ACCESS mdb file
+# This script uses library("odbc") to access the local MS ACCESS mdb file
 # in the ../data/raw_data folder.
-# ASSUMES you are working on Windows machine using locally installed R and data files.
+# ASSUMES
+# - you are working on Windows machine using locally installed R and data files.
+# - you are working on 32-bit version of R (so can use 32-bit drivers for odbc)
 ########
 # Joel H. Reynolds, joel_reynolds@nps.gov
 # Last Edit: 2018 Oct 5
 ########
+# Misc. Notes on Background:
+# on PC: Control Panel -> Admin -> ODBC Data Sources (64 bit)
+# -> User DSN
+#    only shows 32 bit drivers (dBASE files, Excel Files, MS Access Database [*mdb]).
+# -> Drivers
+#    shows 'SQL Server' with ref to a 32-bit dll.
+# ==> suggests have to work in 32 bit version of R & Rstudio, etc.
+# So in Rstudio, reset Global Options to 32-bit R.
+# Added check (R.Version call below) that will stop if run on 64-bit version.
+##########################################################
 
 ########## WARNING #################
 # THIS SCRIPT IS INCOMPLETE        #
 ########## WARNING #################
 
+# Check that using 32-bit R; if not, stop!
+if (R.Version()$arch!="i386") {warning("Need to run on 32-bit R to access database");stop()}
+### Modify to extract bit from sessionInfo()$platform via grep to get at XX in "(XX-bit)"
+
 
 # add package
-library("RODBC","odbc")
+library("odbc");library("DBI");
+
+
+# likely unuseful 32-bit solution since "odbcConnectAccess is only usable with 32-bit Windows"
+library(RODBC)
+db <- RODBC::odbcDriverConnect("Driver={Microsoft Access Driver (*.mdb, *.accdb)};
+                        DBQ=..\\data\\raw_data\\Coastal_Bird_Survey_Data.mdb")
+
+# Get data
+data <- as_tibble(sqlFetch (db , "Table or Query Name", rownames=TRUE))
+
+###############
+conn <- odbc::dbConnect(odbc(),
+                 Driver = "SQL Server",
+                 Server = "localhost\\SQLEXPRESS",  #way to set to local machine
[…]

### 2018-10-18 analysis/scripts/data_exploration.R
Made pathways reproducible.

@@ -1,0 +1,9 @@
+library(readr)
+species_data <- read_csv("./analysis/data/raw_data/CSVs/copy_of_species_data.csv")
+
+library(dplyr)
+library(tidyr)
+tlu_species_codes <- read_csv("./analysis/data/raw_data/CSVs/tlu_species_codes.csv")
+species_data_names <- left_join(species_data, tlu_species_codes, by = c("Species" = "Code")) %>%
+  separate(Date, c("Date", "Time"), sep = " ") %>%
+  select(-Time)

### 2018-10-18 analysis/scripts/data_retrievalCSVs.R
Ran into some errors with the CSV general retrial script; will look into it

@@ -1,0 +1,15 @@
+# Retrieve CSV Data when working on swarthmore R server
+############
+# This script uses library("readr") to access the comma-separated files associated with species data.
+########
+# Madeleine Ward
+# Last Edit: 2018 Oct 17
+########
+
+path = "./analysis/data/raw_data/CSVs/"
+file_names <- dir(path)
+library(readr)
+for(i in 1:length(file_names)){
+  print(file_names[i])
+  read_csv(file_names[i]) #this gets an error
+}

### 2018-10-20 analysis/scripts/data_retrievalCSVs.R
Finally fixed the error that was blocking the retrieval script. Script now collects all .csv files as data frames for future use.

@@ -7,9 +7,16 @@
 ########
 
 path = "./analysis/data/raw_data/CSVs/"
-file_names <- dir(path)
+
+#retrieve file names with full path included
+file_names <- dir(path, full.names = TRUE)
 library(readr)
 for(i in 1:length(file_names)){
-  print(file_names[i])
-  read_csv(file_names[i]) #this gets an error
+
+  #remove path and .csv from tibble name
+  title <- gsub(".*/", "", file_names[i])
+  title <- gsub(".csv", "", title)
+
+  #assign tibble names to data frames
+  assign(title, read_csv(file_names[i]))
 }

### 2018-10-22 analysis/scripts/data_exploration.R
basic scripts for the introductory analysis

@@ -1,9 +1,15 @@
-library(readr)
-species_data <- read_csv("./analysis/data/raw_data/CSVs/copy_of_species_data.csv")
+# check dates and unit of first non-zero counts each yea for each spp (relative to first survey date)
+############
+# This script uses library("dplyr") to wrangle species data.
+########
+# Madeleine Ward
+# Last Edit: 2018 Oct 22
+########
 
 library(dplyr)
 library(tidyr)
-tlu_species_codes <- read_csv("./analysis/data/raw_data/CSVs/tlu_species_codes.csv")
-species_data_names <- left_join(species_data, tlu_species_codes, by = c("Species" = "Code")) %>%
-  separate(Date, c("Date", "Time"), sep = " ") %>%
-  select(-Time)
+library(lubridate)
+species_data_names <- left_join(species_datafrom_Excel, tlu_species_codes, by = c("Species" = "Code")) %>%
+  separate(Date, c("Date1", "Time"), sep = " ") %>%
+  mutate(Date = mdy(Date1))
+  #group_by()

### 2018-10-22 analysis/scripts/observers.R

@@ -1,0 +1,8 @@
+#Distribution of observers by survey date by year
+#Distribution of observer experience(?) by survey date by year?
+############
+#
+########
+# Madeleine Ward
+# Last Edit: 2018 Oct 22
+########

### 2018-10-22 analysis/scripts/surface_area.R

@@ -1,0 +1,7 @@
+# Find the average surface area per unit
+############
+#
+########
+# Madeleine Ward
+# Last Edit: 2018 Oct 22
+########

### 2018-10-22 analysis/scripts/tide_states.R

@@ -1,0 +1,12 @@
+# title
+############
+# This script examines the interaction between tidal state and sequencing of observation sites
+########
+# Madeleine Ward
+# Last Edit: 2018 Oct 22
+########
+
+library(dplyr)
+tide_states <- tbl_Events %>%
+  full_join(tlu_tide_state, by = c("Tide_State" = "code")) %>%
+  full_join(tlu_locations, by = "Location_ID")

### 2018-10-22 analysis/scripts/units_species.R

@@ -1,0 +1,7 @@
+#Value of all units - only some spp in unit X?; “unit by species occurrence” relationships?
+############
+#
+########
+# Madeleine Ward
+# Last Edit: 2018 Oct 22
+########

### 2018-11-27 analysis/scripts/data_exploration.R
some of the graph data from over break.

@@ -11,24 +11,65 @@ library(dplyr)
 library(lubridate)
 library(tidyr)
 
-species_data_names <-
+## @knitr data_before_2007
+
+#species_before_2007 <-
   #combine species observation with species names
-  left_join(species_datafrom_Excel, tlu_species_codes,
-            by = c("Species" = "Code")) %>%
+  #left_join(species_datafrom_Excel, tlu_species_codes,
+            #by = c("Species" = "Code")) %>%
 
   #separate date and time
   #use lubridate pkg to create Date variable
-  separate(Date, c("Date1", "Time"), sep = " ") %>%
-  mutate(Date = mdy(Date1), Year = as.numeric(format(Date, "%Y"))) %>%
+  #separate(Date, c("Date1", "Time"), sep = " ") %>%
+  #mutate(Date = mdy(Date1), Year = as.numeric(format(Date, "%Y"))) %>%
 
-  select(Year, Date, `Survey Unit`, Species, `# Observed`,
-         Category, Family, Common_Name) %>%
-  filter(`# Observed`!=0) %>%
-  group_by(Species) %>%
-  arrange(Year) %>%
+  #select(Year, Date, `Survey Unit`, Species, `# Observed`,
+         #Category, Family, Common_Name) %>%
+  #filter(`# Observed`!=0) %>%
+  #group_by(Species) %>%
+  #arrange(Year) %>%
+
   #take first observation from each year
-  filter(row_number()==1)
+  #filter(row_number()==1)
 
 #to do: make row for # of days past first survey date?
 
-#question: is
+
+## @knitr data2007
+species_data_2007 <- left_join(tbl_Field_Data, tbl_Events,
+                               by = "Event_ID") %>%
+  left_join(tlu_species_codes,
+            by = c("Species" = "Code")) %>%
+  separate(Start_Date, c("Date", "Time_0"), sep = " ") %>%
+  separate(Start_Time, c("Date_01", "Time_Start"), sep = " ") %>%
+  #how to fit all these into one argument??
+  separate(End_Time, c("Date_02", "Time_End"), sep = " ") %>%
+  mutate(Date = mdy(Date),
+         Day_Month = format(Date, "%m-%d")) %>%
+  filter(`Num_ Observed` != 0, Year >= 2003) %>%
+  group_by(Year) %>%
+  arrange(Date) %>%
+  mutate(Days_since_first = Date - Date[row_number()==1]) %>%
[…]

### 2018-11-28 analysis/scripts/data_exploration.R
Added some to paper.rmd. Also, part 2 of script changes, although it won't knit as is.

@@ -11,8 +11,6 @@ library(dplyr)
 library(lubridate)
 library(tidyr)
 
-## @knitr data_before_2007
-
 #species_before_2007 <-
   #combine species observation with species names
   #left_join(species_datafrom_Excel, tlu_species_codes,
@@ -34,8 +32,6 @@ library(tidyr)
 
 #to do: make row for # of days past first survey date?
 
-
-## @knitr data2007
 species_data_2007 <- left_join(tbl_Field_Data, tbl_Events,
                                by = "Event_ID") %>%
   left_join(tlu_species_codes,
@@ -73,3 +69,7 @@ plot <- ggplot(species_data_2007, aes(x=as.factor(Days_since_first), y=`Num_ Obs
 for (i in seq(1, length(unique(species_data_2007$Species)), 6)) {
   print(plot)
 }
+
+
+#time series
+#

### 2018-11-30 analysis/scripts/data_exploration.R
Added necessary library calls to "paper.Rmd" so that they are all in one place.

@@ -1,15 +1,12 @@
 # check dates and unit of first non-zero counts each year
 # for each spp (relative to first survey date)
 ############
-# This script uses library("dplyr") to wrangle species data.
+# This script uses libraries "dplyr", "lubridate", and "tidyr" to wrangle species data;
+# and uses library "ggplot2" to visualize the data.
 ########
 # Madeleine Ward
 # Last Edit: 2018 Oct 22
 ########
-
-library(dplyr)
-library(lubridate)
-library(tidyr)
 
 #species_before_2007 <-
   #combine species observation with species names
@@ -72,4 +69,17 @@ for (i in seq(1, length(unique(species_data_2007$Species)), 6)) {
 
 
 #time series
-#
+#dot plot
+ggplot(species_data_2007, aes(as.factor(Year), `Num_ Observed`)) +
+  geom_boxplot() +
+  geom_dotplot(#binaxis='y',
+               #stackdir='center',
+               #dotsize = .5,
+               #fill="red"
+    ) +
+  theme(axis.text.x = element_text(angle=65, vjust=0.6)) +
+  labs(title="Box plot + Dot plot",
+       subtitle="City Mileage vs Class: Each dot represents 1 row in source data",
+       caption="Source: mpg",
+       x="Class of Vehicle",
+       y="City Mileage")

### 2018-11-30 analysis/scripts/data_retrievalCSVs.R

@@ -10,7 +10,7 @@ path = "./analysis/data/raw_data/CSVs/"
 
 #retrieve file names with full path included
 file_names <- dir(path, full.names = TRUE)
-library(readr)
+
 for(i in 1:length(file_names)){
 
   #remove path and .csv from tibble name

### 2018-11-30 analysis/scripts/surface_area.R

@@ -1,7 +1,0 @@
-# Find the average surface area per unit
-############
-#
-########
-# Madeleine Ward
-# Last Edit: 2018 Oct 22
-########

### 2018-11-30 analysis/scripts/units_species.R

@@ -6,9 +6,6 @@
 # Last Edit: 2018 Oct 22
 ########
 
-library(stringr)
-
-
 tbl_Locations$GIS_Location_ID <- str_sub(tbl_Locations$GIS_Location_ID,start=-1)
 
 species_units <-

### 2018-12-03 analysis/scripts/data_exploration.R
Added libraries to paper.RMD. Also made all script files run internally (look around line 260 in paper for calls to both data_exploration and units_species, the problematic ones). However, I still can't knit the doc because of some sort of working directory problem, so run into an error at the CSV script. If the directory isn't giving you a problem it should knit.

@@ -51,24 +51,36 @@ species_data_2007 <- left_join(tbl_Field_Data, tbl_Events,
   #take first observation from each year
   #filter(row_number()==1)
 
-species_data_2007$type <-
-ifelse(species_data_2007$`Num_ Observed` < mean(species_data_2007$`Num_ Observed`), "below", "above")
+#species_data_2007$type <-
+#ifelse(species_data_2007$`Num_ Observed` < mean(species_data_2007$`Num_ Observed`), "below", "above")
 
-#unfinished
-plot <- ggplot(species_data_2007, aes(x=as.factor(Days_since_first), y=`Num_ Observed`)) +
+#faceted by species?
+ggplot(species_data_2007,
+       aes(as.factor(Days_since_first),
+           `Num_ Observed`)) +
   geom_bar(stat='identity', aes(fill = as.factor(Year))) +
-  labs(title="Ordered Bar Chart",
-       subtitle="Make Vs Avg. Mileage",
-       caption="source: mpg") +
-  theme(axis.text.x = element_text(angle=65, vjust=0.6)) +
-  facet_wrap(~Species)
+  #theme(axis.text.x = element_text(angle=65, vjust=0.6)) +
+  #paginate
+  facet_wrap_paginate(~Species, ncol = 1)
 
-for (i in seq(1, length(unique(species_data_2007$Species)), 6)) {
-  print(plot)
-}
+#another way to paginate
+#for (i in seq(1, length(unique(species_data_2007$Species)), 6)) {
+ # print(plot)
+#}
 
+#first count every year?
+ggplot(species_data_2007 %>%
+         ungroup() %>%
+         group_by(Species, Year) %>%
+         arrange(Days_since_first) %>%
+         top_n(1), aes(as.numeric(Days_since_first),
+               `Num_ Observed`,
+               color = as.factor(Year))) +
+  #using facet_wrap_paginate() from ggforce
+  #to distribute over pages
+  facet_wrap_paginate(~Species, ncol = 1)
 
-#time series
+#time series?
 #dot plot
 ggplot(species_data_2007, aes(as.factor(Year), `Num_ Observed`)) +
   geom_boxplot() +
@@ -76,10 +88,4 @@ ggplot(species_data_2007, aes(as.factor(Year), `Num_ Observed`)) +
                #stackdir='center',
                #dotsize = .5,
                #fill="red"
-    ) +
-  theme(axis.text.x = element_text(angle=65, vjust=0.6)) +
[…]

### 2018-12-03 analysis/scripts/units_species.R

@@ -9,15 +9,47 @@
 tbl_Locations$GIS_Location_ID <- str_sub(tbl_Locations$GIS_Location_ID,start=-1)
 
 species_units <-
-  left_join((species_datafrom_Excel <-
-               mutate(species_datafrom_Excel,
-                      survey_unit = as.character(`Survey Unit`))),
+  left_join(species_data <- left_join(tbl_Field_Data, tbl_Events,
+                                                                by = "Event_ID") %>%
+              left_join(tlu_species_codes,
+                        by = c("Species" = "Code")) %>%
+              separate(Start_Date, c("Date", "Time_0"), sep = " ") %>%
+              separate(Start_Time, c("Date_01", "Time_Start"), sep = " ") %>%
+              #how to fit all these into one argument??
+              separate(End_Time, c("Date_02", "Time_End"), sep = " ") %>%
+              mutate(Date = mdy(Date)) %>%
+              filter(Year >= 2003) %>%
+              group_by(Year) %>%
+              arrange(Date) %>%
+              mutate(Days_since_first = Date - Date[row_number()==1]),
             tbl_Locations,
-            by = c("survey_unit" = "GIS_Location_ID")) %>%
-  group_by(survey_unit, Species) %>%
-  summarize(total_Observed = sum(`# Observed`)) %>%
-  left_join(tlu_species_codes, by = c("Species" = "Code")) %>%
-  spread(survey_unit, total_Observed)
+            by = "Location_ID") %>%
+  group_by(GIS_Location_ID, Year, Days_since_first, Species) %>%
+  summarize(total_Observed = sum(`Num_ Observed`)) %>%
+  left_join(tlu_species_codes, by = c("Species" = "Code"))# %>%
+#optional: spread command
+  #spread(GIS_Location_ID, total_Observed)
 
-#to do: change to naming conventions
-#to do: make plot?
+#to do: make plot
+ggplot(species_units, aes(as.numeric(Days_since_first),
+                          total_Observed,
+                          color = as.factor(Year))) +
+  geom_jitter(height = .3, aes(fill = GIS_Location_ID)) +
+  #in ggforce package to supposedly distribute facets over pages
+  facet_wrap_paginate(~Species, ncol = 1, scales = "free") +
+  ylim(c(0, 2000))
+
+#how to
+  #1. order bars with a y aesthetic,
+  #2. have species not take up so much space
+  #3. get rid of 0/NA values to take up less space?? ("free_x" not working)
[…]

### 2018-12-10 analysis/paper/paper.tex
i have changed the knitting settings to put the paper in the project directory instead of the /paper folder. Now the codes for scripts run!

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{eurosym}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Assessment of the KLGO Coastal Waterbird Survey Monitoring Program},
            pdfkeywords={keyword 1; keyword 2; keyword 3},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Assessment of the KLGO Coastal Waterbird Survey Monitoring Program}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{true \\ true}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{10 December, 2018}

\begin{document}
\maketitle
\begin{abstract}
The Coastal Waterbird Survey has been conducted, with varying levels of
effort and in-season frequency, since 2003 (Surdyk and Evans
\protect\hyperlink{ref-Surdyk2018}{2018}). The area of interest has been
divided into eight contiguous survey units; on each survey, at each unit
the number, age, composition (sex), breeding status, and noteworthy
behavioral observations are recorded for each species (Anonymous
\protect\hyperlink{ref-Anonymous2009}{2009}). Given the constraints on
personnel, time, and budgets, it is imperative that surveys be conducted
efficiently in terms of both choice of sampling design and choice of
sampling effort. Now that KLGO has multiple years of experience
implementing the protocol, it is timely to step back and assess how
things are going. We review the current survey programâ\euro{}™s
objectives, design, identify statistical issues, and recommend potential
solutions. Major topics include â\euro{}¦. The study concludes with a
series of recommended tasks that the park should undertake to improve
the coastal waterbird surveyâ\euro{}™s efficiency and effectiveness.
\end{abstract}

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\section{Things To Discuss}\label{things-to-discuss}

\begin{itemize}
\tightlist
\item
  add all libraries used in scripts to `setup' code chunk (just above)
  rather than just loading in script so that it is easy to see what
  needs to be loaded before you can successfully knit.
\end{itemize}

\section{Introduction}\label{introduction}

Since 2003, staff at Klondike Gold Rush National Historical Park (KLGO)
have surveyed coastal waterbird species distribution and abundance
multiple times each year from late April through late September (Coastal
Waterbird Survey or `CWS'; (Surdyk and Evans
\protect\hyperlink{ref-Surdyk2018}{2018})). This is sufficient history
to warrant stepping back and assessing the monitoring program. Such
assessments are key to an adaptive monitoring program ({\textbf{???}}),
are an important and distinct phase to promote program, and
institutional, learning and evolution (Reynolds et al.
\protect\hyperlink{ref-Reynolds2016}{2016}), and align with guidance
from NPS' Inventory and Monitoring Division ({\textbf{???}}, Mitchell et
al. (\protect\hyperlink{ref-Mitchell2018}{2018})).

Assessing a monitoring program focuses on four broad topics following
the major Phases of a monitoring program (ibid):\\
1. \emph{why} - {[}Problem Framing \& Information Objectives{]} Are the
management objective(s) motivating the monitoring well articulated? Do
they remain relevant and important?\\
2. \emph{how} - {[}Data Collection Design{]} Does the monitoring data
collection design effectively addresses the information objectives?
Given interim learning regarding system characteristics and improvements
in measurement technologies, are the information goals (including
accuracy and precision goals), actually feasible? If so, can the data
collection be made more efficient, reducing the observation burden on
staff resources?\\
3. \emph{how} - {[}Data Analysis Design{]} Do the data analysis methods
effectively address the information objectives? Do the data analysis
methods properly account for impactful features of the data collection
design? Are more informative analyses available given interim
improvements in methods?\\
4. \emph{how} - {[}Workflows \& Reporting{]} Are workflows and related
software systems in place to allow timely, and efficient, data
management, analysis updating, reporting, and distribution of findings
to target audiences of information users?

This assessment addressed all four topics, but focused predominantly on
topics (i) Problem Framing \& Information Objectives, (ii) Data
Collection Design, and (iii) Data Analysis Design. Since `Problem
Framing \& Information Objectives' determine the relevancy of many of
the potential assessment actions related to the other topics, the
assessment was drafted as a first stage of an iterative effort,
providing high level feedback, including listings of potential detailed
syntheses and workflow enhancement activities (such as improved database
interface, automation of annual analyses and updating of syntheses,
etc.). With this foundation, further dialogue with KLGO staff regarding
relevant `Problem Framing \& Information Objectives' and other
priorities will determine which follow-on activities to pursue.

\section{Background}\label{background}

\section{Methods}\label{methods}

The CWS monitoring program was compared against a recommended checklist
of monitoring program components developed from published literature
(Appendix 1). It identifies multiple component decisions associated with
each of the major Phases of a monitoring program (briefly overviewed
above). For each component, the relevant aspects of the current CWS were
reviewed, potential issues identified, potentially informative
diagnostic analyses suggested, and recommendations provided. The
characterization of the CWS was based on KLGO protocols and reports
(Table 1), a site visit (6/4-6/2018) and personal communications with
the KLGO Natural Resources Program Manager (Jami Belt, winter and spring
2018), KLGO Resources Manager (Anne Matsov, summer 2018) and
long-standing survey participants (Elaine Furbish, Skagway Bird Club,
spring and summer 2018).

\section{Assessment Results}\label{assessment-results}

\subsection{Phase 1 - Problem Framing and Information
Objectives}\label{phase-1---problem-framing-and-information-objectives}

\emph{WHY} monitor? What potential decisions and issues should the
observations inform?

The management problem or question motivating the need for information
about coastal waterbirds provides the foundation for the survey's data
collection and analysis designs (Fancy, Gross, and Carter
\protect\hyperlink{ref-Fancy2009}{2009}). It clarifies the underlying
goal of the monitoring program and answers ``Why is this an important
problem or question?''.

A robust problem definition should include, among other aspects, the
temporal and geographic scope of the problem; who has the authority to
make decisions that could resolve or address the problem; what
information about the resource is needed to improve the decision making
(the information needs); and who are the stakeholders that are impacted
by the decision and/or interested in the monitoring information.
Stakeholder identification should include interested parties and their
information needs in both near-term - for example, park interpretation
and visitor education activities, as well as long-term - for example,
regional-scale changes in migration patterns or assessment and
improvement of projected climate change impacts on avifauna (e.g., Wu et
al. \protect\hyperlink{ref-Wu2018}{2018}). A good understanding of
multiple information needs and pathways to influence decision making
will guide the required data analyses and reporting and outreach.

The information needs are what the monitoring effort aims to address
(Reynolds \protect\hyperlink{ref-Reynolds2012}{2012}). Even when the
focus is simply the status and trend of a resource, there is an unstated
expectation that someone will endeavor to achieve or maintain a
desirable state or reverse an undesirable trend (Reynolds et al.
\protect\hyperlink{ref-Reynolds2016}{2016}). Whomever has that
responsibility and authority (person, office, or organization) is a
priority stakeholder. They should not only receive the reports but also
allowed an opportunity to clarify their information needs, including the
desired precision and quality of information, the timing and format of
reporting, and better understand what information is feasible to obtain
(Averill et al. 2010).

\subsubsection{Assessment findings:}\label{assessment-findings}

The 2003 coastal waterbird inventory and survey design was developed to
address ``a top priority bird inventory (need)'' of the Southeast Alaska
Inventory and Monitoring Network (Sharman et al. 2000) - namely, an
inventory of waterfowl/shorebirds at KLGO (Hahr et al.
\protect\hyperlink{ref-Hahr2004}{2004}). This was deemed a priority due
to the area's expected avifaunal richness ((NABCI)
\protect\hyperlink{ref-NABCI2000}{2000}) and the limited availability of
reliable observation records - mainly consisting of systematic surveys
conducted by NPS staff in the early 1980s (Hahr et al.
\protect\hyperlink{ref-Hahr2004}{2004}).

The inventory's stated goal was to document the occurrence of 90\% of
the species of waterbirds likely to occur in KLGO (ibid). The principal
objectives were ``to develop scientifically valid sampling designs for
inventorying waterbirds and breeding landbirds in KLGO, and to collect
baseline information on bird species occurrence, distribution,
abundance, and habitat associations in the park\ldots{}{[}to{]}
\ldots{}provide park managers with timely information that facilitates
land management planning, helps track wildlife population and ecosystem
change over time, and keeps resource managers apprised of changes that
may require action before a crisis occurs'' (ibid). While no Threatened
or Endangered bird species were known to occur in KLGO, as of 2004, nine
species on the Alaska Watchlist had been documented in the park (ibid),
thus suggesting potential priority information needs associated with
monitoring for changes in those species occurrence and abundance.
Potential secondary objectives mentioned include ``investigations into
the breeding ecology, phenology, reproductive success and population age
structure of KLGO breeding birds'' (ibid). Later reports mention
additional objectives of estimating changes in species occurrence and
composition (St. Pierre \protect\hyperlink{ref-St.Pierre2015}{2015}),
characterizing timing of peak spring waterbird migration and its
relation to spring eulachon runs (Surdyk
\protect\hyperlink{ref-Surdyk2017}{2017}).

Stakeholders mentioned or implied in park reports include park
superintendent and related managers, those charged with monitoring
regional and internationally important bird habitats, the Skagway Bird
Club, the Alaska Landbird Monitoring Survey and the Boreal Partners in
Flight.

The coastal waterbird surveys have been continued since 2003 (Surdyk and
Evans \protect\hyperlink{ref-Surdyk2018}{2018}).

QUOTES Hahr and Trapp (2004): We intend to use the results of these
inventories to create a revised bird checklist and an atlas database for
the Taiya and Skagway River watersheds in collaboration with the Skagway
Bird Club.

This protocol addresses one monitoring question with three objectives:
(1) what is the annual nest density and reproductive (fledgling) success
in known Hawaiian petrel colonies, (2) what are the long-term trends in
colony distribution and density monitored in approximate 5-year
intervals, and (3) are these affected by predator control? The first
goal of monitoring is to obtain unbiased estimates of Hawaiian petrel
nest density and reproductive (fledging) success from known colonies in
HAVO and HALE in order to detect changes in colony growth or decline.
The second goal of monitoring is to periodically (approximately every
five years) obtain unbiased estimates of nest density from potential
Hawaiian petrel habitat. This information can be used to assess changes
in density and distribution of subcolonies across the landscape. The
third goal is to estimate nest density and fledging success in areas
undergoing different management regimes to assess effectiveness of
management. Benchmark levels of these estimates could serve as warnings
of the need for modified management or further investigation of these
colonies.

Though not explicitly characterized in the protocol or reports reviewed
for this assessment (Table 1), the following target audiences and
information users were identified through conversations with KLGO staff,
partners, and other waterbird researchers and stakeholders in Alaska.
(current) - park - Skagway Bird Club; compiles for addition to
`significant bird observations in SE AK', which gets compiled in to AK
sumamries for `North American Birds' (mainly spp lists) (potential) -
BPIF - PSG - Audubon Alaska (for IDing Important Bird Areas - though
require accurate density estimates) - Audubon North America - cliamte
species distribution models (Nicole Michel, Brooke Bateman, etc.); esp
if associated w/ repeat sites \& effort data (date, time of day,
duration, distance/area covered) - see Nicole email 6/7/2018. Watchlist
of projected changes - (Wu et al. \protect\hyperlink{ref-Wu2018}{2018})

\subsubsection{Recommendations:}\label{recommendations}

Potential Follow Up Activities:

Define the \emph{Target Audiences} for the information and their
Objectives in using the information

\emph{Decision Makers} Who are target audiences for this information?
Who uses it or may use it? Local contact: C. Elaine Furbish, Skagway
Bird Club, Google Site Predominantly historical park; limited resource
staff and capacity Visible natural resource of interest to local
community and visitors; interp and educational opportunity; topic of
information requests; provides `community service' given absence of any
other agency with natural resource staff; Annual report (time lag?);
National Resource Condition Assessment; other? State of the Park?
Potential for contribution to regional, flyway, and WASO programs (e.g.,
CCRP \& Audubon climate watch, etc.) Current sightings passed to local
bird club \& used by avian-interested visitors Used in quarterly
regional `significant bird observations', which in turn is used to
develop the Alaska-wide summary for the journal North American Birds
(published by the American Birding Association
\url{http://www.aba.org/nab} ) and includes `\ldots{}sightings of birds
that are out of range or out of season or occur regularly in small
numbers, noteworthy breeding records, unusually large or small numbers
of a particular species, unusual migration dates, etc\ldots{}.' (S. C.
Heinl, 2017, Summary of SE AK Bird Observations, June-July 2017,
accessed 2018 June 6). Queries to Heather Renner \& Robin Corcoran re:
PSG-type users, etc. Opportunity for community based monitoring to track
projected spp impacts of climate change, change in phenology,
occurrence, abundance, etc.?

\emph{Decisions} What types of decisions is it / will it be used for?
(specify for each target audience/ end user) No specific decisions
Current: KLGO - spp lists, phenology of occurrence (guidance to
visitors), trends in richness (NRCA), abundance? Skagway bird club - spp
list, current occurrences (activity re: rare \& unusual) SE AK /
Regional - spp lists; unusual sightings, etc. (spp occurrences \& broad
seasonal timing). AKR - WASO/NPS - ??CCRP?? Question to Gregor \& Mel
(AK Audubon). Could be used for garnering community engagement in
Climate Watch \& other national efforts associated with Audubon's
Climate Report. Not used in continental analyses. ``One key message that
I want to emphasize is that the focus of this work was purely on climate
suitability. What this means, as we indicate in the briefs'
``Important'' box up front and ``Caveats'' section further back, is that
significant changes in climate suitability, as measured here, will not
always result in a species response, and all projections should be
interpreted as potential trends. Multiple other factors mediate
responses to climate change, including habitat availability, ecological
processes and biotic interactions, dispersal capacity, species'
evolutionary adaptive capacity, and phenotypic plasticity (e.g.,
behavioral adjustments). Ultimately, models can tell us where to focus
our concern and which species are most likely to be affected, but
monitoring is the only way to validate these projections and should
inform any on-the-ground conservation action.'' Gregor Schuurman, email
to DENA

Potential: KLGO: (easy) - changes in occurrence (spp, season scale);
changes in phenology (migration timing, by spp; breeding timing, by
spp); broad (order of magnitude) changes in abundance (index); broad (?)
changes in composition (by spp \& timing of survey)?; species occurrence
x unit (\textasciitilde{}habitat?) - refine expected lists per survey
and develop targeted SOP training materials; (harder) - finer changes in
abundance; finer changes in composition and age; ???

What Objectives are the decision makers hoping to achieve by using this
information?

Over what spatial scales is the information being used? e.g., at the
park, in the community, regional scale, flyway scale, nationally,
internationally For each spatial scale of interest, what are the
associated time scales of interest for detecting changes? E.g., annual,
w/in season, spring vs fall, etc.

From another report but quotes from Olsen et al. (1999): Olsen (et al.
1999) noted that ``Most of the thought that goes into a monitoring
program should occur at this preliminary planning stage. The objectives
guide, if not completely determine the scope of inference of the study
and the data collected, both of which are crucial for attaining the
stated objectives.'' Olsen goes on to say that a ``clear and concise
statement of monitoring objectives is essential to realize the necessary
compromises, select appropriate locations for inclusion in the study,
take relevant and meaningful measurements at these locations, and
perform analyses that will provide a basis for the conclusions necessary
for meeting the stated objectives.''

Fancy, Gross, and Carter (2009): The development of monitoring
objectives, which provide additional focus about the purpose or desired
outcome of the monitoring effort, was an iterative process that
sometimes required several years to refine. Early in the design process,
monitoring objectives were stated in more general terms, such as
``Determine trends in the incidence of disease and infestation in
selected plant communities and populations'', whereas the final
monitoring plan and protocols provided monitoring objectives that met
the test of being realistic, specific, and measurable (e.g., ``Estimate
trends in the proportion, severity, and survivorship of limber pine
trees infected with white pine blister rust at Craters of the Moon
National Monument''; Garrett et al. 2007).

Structure from Northeast Coastal and Barrier Network
\url{https://irma.nps.gov/DataStore/Reference/Profile/2195238}
Justification =\textgreater{} Vital Sign =\textgreater{} Monitoring Goal
=\textgreater{} Monitoring Questions =\textgreater{} Monitoring
Objectives =\textgreater{} Measures (eg pg 55 \& 56)

Clarify Objectives of the Decision Makers Among the decision maker
objectives identified in 1.b.ii, clarify the Fundamental Objective(s)
from the Means Objectives, then diagram the Objectives Hierarchy.
Identify the Attributes that need to be monitored to assess achievement
of each of the objectives. (`Concepts by Theory', Ford 2001) Identify
the Measurements selected to characterize each attribute. (`Concepts by
Data', Ford 2001) {[}WHAT{]} Identify any Threshold values associated w/
each Measurement

Q: compiling and assessing regional trends in waterbirds in SE AK? Asked
of Heather Renner (Ak Maritime), Robin Corcoran (Kodiak NWR) Answers:
Not really; regular discussion at PSG but no one has taken it on. Kathy
Kuletz works on statewide trends in at sea seabird observations; Piatt
GLBA work (\& Martin Renner); Robin C: ``There's always talk at the PSG
meeting about the North Pacific seabird colony database, or some
variation of that database for marine birds, but it never gets anywhere.
Basically there is no money for a database manager. I've been submitting
my data to Rob Kaler in Migratory Bird Management, but I think that's
just for storage until the money for a database manager materializes.
More recently Rob Suryan w/NOAA has been interested - for the North
Pacific Pelagic Seabird Database. You might want to contact him, he's
leading the Gulf Watch Alaska Long-Term Ecosystem Monitoring Program,
and maybe interested in this work.''

NPS/Audubon: Wu et al. 2018. Projected avifaunal responses to climate
change across the U.S. NPS. PLOS ONE. Appendices include projected
park-specific turnover, colonization, and extirpations projections -
potential to test with this survey. Would require discussions w/ CCRP \&
ARO - ideally as a coordinated effort with, say, AK Audubon and ACF(?)
or LCCs as tracking metric.\\
Appendices include ``Potential management goals and activities for
parks, organized by trend group.''

Melanie AK Audubon: IBA requires demonstrating higher than `usual'
densities (relative to surrounding area).

Nicole Michel \& Brooke Batemen Audubon National Science Team: of
interest and possible use in national scale quantitative analyses or
climate species distribution models. ``Scaling up and integrating
datasets across regions, including data from different goals, is
something that we're starting to do in other regions. ``

Occurrence, Richness Phenology (migration, breeding) Abundance - Even
without the structure needed to estimate detection probability, if the
same protocol is used annually this data could be used to produce
long-term trend indices for the study area. (competing staff priorities
- peak (?) usually aligns with hooligan run. Possibly reduce summer
frequency?)

Diagram the Conceptual Model of the system E.g., create an influence
diagram, being sure to denote the Fundamental \& Means Objective(s)
Factors influencing their achievement (aka `system drivers') Associated
management or policy decisions and/or potential actions. Include a
written summary of the literature upon which the conceptual model is
based.

Summarize the management or policy decisions the monitoring is to inform
(if any) based on 3.c.

\subsection{Phase 2 - Monitoring \& Analysis
Design}\label{phase-2---monitoring-analysis-design}

HOW monitor? Clarify the various design choices.

Clarify rationale for monitoring versus some other means of addressing
the information objectives (e.g., literature review, controlled
experiment, etc.), and type of monitoring being undertaken (status \&
trends or surveillance, threshold, effectiveness, or adaptive
management).

Questions: Why? What are information goals \& objectives? What sort of
observations would trigger action? What actions?

\subsubsection{Survey Design}\label{survey-design}

\emph{Target Frame} - space, time, scale, target populations
\emph{Spatial} Upper Lynn Canal visible from coastline from Skagway to
Dyea, including Skagway \& Taiya river mouths and Nahku Bay (between
Skagway and Dyea Point)

\begin{verbatim}
#> [1] "/home/mward2/klgo_bio_monitoring"
\end{verbatim}

Species: Targets: loons, grebes, cormorants, herons, ducks, geese,
plovers, sandpipers, gull, alcids, {[}kingfishers{]} Questions: clarify
role of kingfishers.

Incidental: 48 other spp Questions: are these all migratory? Any
year-round residents? Any of special concern? Any especially problematic
for detection on water or identification of spp?

(Hahr et al. \protect\hyperlink{ref-Hahr2004}{2004}), pg 13: Areas of
coastline that were visible from public access points between Skagway
and the Taiya River delta were divided into eight census units and
surveyed once per week in late April and May and bi-weekly during late
summer and fall (August -- mid September) (Figure 5). Surveys were
conducted as weather and scheduling permitted and dates/time were not
selected at random. Standardized observation points were established for
each census unit to ensure the most complete coverage of the study area,
and to minimize observer bias. Observers counted and identified all
birds visible within each unit and documented birds opportunistically
observed while moving between units. Counts were made by scanning the
census units with Bushnell Legend 10 x 42 binoculars and a Bausch \&
Lomb 20-60 x 70 spotting scope until the observer was able to quantify
all visible birds in the unit. Surveys were only conducted under optimal
weather conditions; however, occasionally conditions deteriorated during
the course of a survey making species identification difficult
(especially in the case of low clouds, fog and wind). Data were recorded
on standardized data forms developed after Collins et al. (2001)
(Appendix B). ArcView shapefiles of the observation points and census
units were created for use in the park's GIS. All data, original data
forms, and field notes were archived at KLGO.

\emph{Temporal} ??how define `season'??

Questions: what is cue to start of survey season? \emph{Madeleine later}
Diagnostic analysis =\textgreater{} check dates \& unit of first
non-zero counts each year for each spp (relative to dates of initiating
surveys)

\emph{Sample Frame} Spatial: same coverage as Target Frame. Area split
into eight survey areas (see Hahr et al.
\protect\hyperlink{ref-Hahr2004}{2004}). Questions: Access to all units?
Ease of observation of different units? Ease of distinguishing
boundaries? \emph{Madeleine later} Diagnostic analysis =\textgreater{}
value of all units - only some spp in unit X?; ``unit by species
occurrence'' relationships?

\emph{Sample Selection process} - any strata or clusters? Two-stage
processes? Probabilistic selection? {[}get in to enumerative vs
analytical distinction of Deming?{]} Currently survey all of frame
(`census' of units); no spatial sampling uncertainty. All spatial
uncertainty comes from Observation Process and detection issues.
Temporal sampling uncertainty remains.

Questions: feasible to reduce \# of units? See Qs above about `unique
contributions' of each unit. Logistic considerations? What effort level
is feasible? Diagnostic analysis =\textgreater{} any `representative'
units? How would that be defined?

\emph{Survey timing} Per season -

Questions: how define when to begin surveys? Zeros? Triggering events?
Potential confounding factors? (changes in phenology, ClCh); important
objective or confounder? Design options for addressing? Diagnostic
analysis=\textgreater{} Signs of confounding?

W/in season - (e.g., LEAU St.~George); potential refinement? Temporal:
(migration) 1 per wk, Apr 27 - May 20; (breeding) 1 per 2 wks, Jun 6 -
Jul 18; (migration) 1 per wk, Aug 1 - Sept 28. 2016: 17 surveys total
Questions: How define beginning \& end of breeding period? Info goals
from migration periods? From breeding periods? Design options to reduce
frequency? Diagnostic analysis=\textgreater{} .assessment of impact on
priority info goals of different frequencies? .spp specific occurrency
by survey date across years - capturing onset and end of migration?
.other assessments of survey timing relative to phenology of spp .mine
FB, Eagle Chat, others for date of first occurrences (see 2016 report);
Elaine. Requires clarifying information goals of survey vrs `first
occurrence'. E.g., not tail behavior.

Per survey route sequencing and tidal influence; SOP: 1,2,3,4,5,6,8,7 or
reverse, depending on tide timing (determines when do unit 7).

Questions: design options for confounding factors? Duration across eight
units: feasible to pull off? issue of bird movement between units?
Anyway to bound potential bias? Design options?

Monitoring Design Membership Design (panels, etc.) Single Panel

Revisitation Design W/in Year; varying effort levels (strata in time
across seasons) Between Years Always Repeat (pure panel), every year
Questions: feasbility of not doing every year? What are information
goals and triggers?

Response Design Attributes \& Measurements Focal targets vrs all species
Priority Spp for region? PSG? At Risk Spp (CCRP?) Audubon AK spp of
concern? See Audubon Climate Report. Relevant spp (intersection with spp
list)

\subsubsection{Observation Process}\label{observation-process}

Question: How determine boundaries? Visual markers. How handle movement
of birds into and out of unit? Check SOPs but basically `until things
stop'. Problem is that that is an accumulating approach that will bias
high areas with lots of productive habitat and/or well timed relative to
changing tidal conditions and sequencing of survey. How handle tendency
to record everything of note: e.g., recording observations in
neighboring units when really supposed to be focused on this unit. E.g.,
variation in time observing and effort observing across units \&
sampling events. This fails to track actual observation activity (time)
devoted to each unit as well as preferential bias to overestimate use in
areas. Think about Forsell \& Zweifelhofer? Chat w/ Emily \& Mark Otto
\& Andy? \#\#\#\# Covariates - captured \& not captured. Value, use in
analyses; more reliable alternative data sources; alignment of spatial
scales of interest and time scales of interest.

Lack of control over time at each site; portion spent traveling between
observation points, variation in number of observation points for each
site.

Variation in survey effort across sites, across per unit area, across
surveys (days), etc.

Major weather conditions (Northerlies, Southerlies) - simplify
conditions to these Unit 1 - recording and impact of presence \& \# of
cruise ships in dock, helicopter tours, etc., etc., other human
disturbances.

Distinguish value of covariates for act of (i) presence/absence
assessment, (ii) abundance count, (iii) composition \& breeding state.

Check SOP to clarify categorization of conditions.

Temp and wind greatly burden data collection process (slowing down
greatly). Variation in observer \#, training, and skill across surveys,
across years

\emph{Madeleine early} Diagnostic analyses =\textgreater{} .distribution
of time duration at each site by date and year .correlation across
units? .average surface area per unit .something about sequencing of
observation sites (e.g., confounding with tidal state)? .distribution of
observers by survey date by year .distribution of observer experience(?)
by survey date by year?

\paragraph{Detection issues associated with survey
conditions}\label{detection-issues-associated-with-survey-conditions}

.species specific counts w/in season by weather conditions and wave
conditions .species occurrence by wave conditions \& weather conditions
.species age and composition by wave conditions \& weather conditions
.total \# species by weather conditions, wave conditions; internally
consistent contrasts to demonstrate value, or lack thereof, of current
covariates. .richness by major weather conditions (Northerlies,
Southerlies, etc.) and for each spp .distribution of weather conditions
.distribution of survey time by weather conditions .diagnostic to ID key
survey times to ensure high quality observers? .consistency in `breeding
periods' across years (as captured in data) .quick diagnostics to ID
covariates seldom recorded, not consistently recorded, consistently
valued (e.g., no basis for contrast), etc.

Value of site-specific covariate recordings vs available `local' weather
(Skagway weather observations from airport - wind direction \& speed,
temp, etc.)

Ppt Codes: 0 - none, 1 - fog, 2 - drizzle, 3 - showers, 4 - light rain,
5 - moderate rain (steady), 6 - heavy rain, 7 - sleet, 8 - light snow, 9
- mod snow, 10 - heavy snow Beaufort scale codes: 0 - \textless{} 1 mph
(air calm); 1 - 1-3 mph (smoke drift show direction of wind); 2 - 4-7
mph wind felt on face, leaves rustle; 3 - 8-12 mph leaves \& small twigs
in constant motion; 4 - 13-18 raises dust, loose paper, moves small
branches; 5 - 19-24 small trees in leaf sway, crested wavelets form on
inland waters; Wave height: 0 - calm; 1 - scaly ripples, no foam crests;
2 - small wavelets, crests glassy, no breaking; 3 - large wavelets,
crests begin to break, scattered whitecaps; 4 - small waves, becoming
longer, numerous whitecaps; 5 - moderate waves, many whitecaps, some
spray SPP \# Age (Adult, imm, juv, mixed, unknown) Composition - M, F,
Fe/juve, Mixed, Unknown) Breeding Status - pairs, etcc; 24 potential
codings see SOP \#3.

Recommendations =\textgreater{} Clearly record standardized observation
locations on map and include in SOP to ensure consistency across
observers and personnel. Issue of veg growth reducing visible portion of
unit from each viewpoint - long-term confounding factor, etc. Value of
covariates, recommendations on what to drop, what to include Observation
quality notes to support assessment of extreme observations, etc.

Detection issues (seeing and recording available birds) Design options
for improving detection \& measurement Hmm? Lit? Emily, Aaron C., Kathi,
Mark Otto, Andy Royle? Feasible to do some sort of double observer
process - coordinate immediately after survey? Only conduct
intermittently per season rather than every time? E.g., like
sightability functions? Major drivers of sightability? Capturing
uncertainty on spp id Capturing uncertainty on spp count UAVs? Observer
training \& behavior classification Training materials from Kodiak or
Homer or Cordova or Haines or Juneau or Yakutat or\ldots{}?

\paragraph{Species identification
issues}\label{species-identification-issues}

Priority ID concerns: Commonly encountered species that can be difficult
Work with Elaine \& Heather \& Laura to develop list of commonly
confused species based on those commonly encountered (database or
Skagway Bird Club sightings) E.g., scoters, scaup, murrelets, etc.)
Target Spp of interest Training materials or other improvements Db error
checking? Other? Other expected sources of bias Species ID Behavior
classification Changes in equipment Summary Metrics Confounding factors
\& Covariates

\subsubsection{Analysis Design}\label{analysis-design}

Priority Summaries of interest (tie to Conceptual Model and Information
Objectives) Associated analyses:

Number of species sighted per year: 2003-2016 (source 2 found 2003-2009)
.Waterbird and non-waterbird species .Positive association between
species richness and year of survey, when excluding year 2014, which
started later than usual. R-squared = .54. .Species richness within
year: 2016 .Individual and max. numbers of observed counts for each
species for year 2016, with first and last sightings noted. .Peak
abundance/species richness identified with observed species counts and
individual sightings across surveys during 2016 season .Peak abundance
(May 4) observed to be in tandem with spring euchalon (April 21-)
.Species richness peaked May 4-11

Observer bias: variation in number of surveys and length of monitoring
season .2010: No seasonal technician; only sporadic surveys .2014:
Surveys begun 1 month late (May 22) .Across years: observer turnover
rate, varying experience

Number of surveys across census units per year (2003-2009) \&
distribution of survey dates

Percent of expected waterbird and breeding landbird species (174)
confirmed (158; 91\%)

Diversity .Simpson's Diversity Index (1/D) (2003-2009): heavily
sensitive to species richness .Q statistic (2003-2009) .Simpson evenness
measure: (1/D)/\# species, where values approaching 1 signal more equal
distribution (not sensitive to species richness)

Comparisons across census units: .Unit characteristics .Area of unit
(square km) .Shoreline habitat (meters) .Waterbird characteristics:
.Average number of waterbirds per survey per square km .Diversity
measures Analysis across dates each year: .Number of species; number of
birds .Diversity measures

Reporting schedule Questions: Who does the analyses? Interest in setting
up as Reproducible Research draft report? General delay in reporting?
When normally completed?

Secondary Summaries of interest Associated analyses Reporting schedule

Survey effort levels / Change detection targets / Power to detect
changes Reiterate targets for `local' inference vrs contribution to
regional trends assessments

1 Biotech conducts just this \& amphi monitoring.

\subsubsection{Data Management \& Reporting Structures / Software /
Workflows}\label{data-management-reporting-structures-software-workflows}

Data Models Design Requirements / User reporting needs

Elaine: Can put stuff in but can't get it out of existing ACCESS db.
Problematic. Bottleneck for analyses and summaries. Ebird - boundaries
don't include Skagway burough. Rather, formerly unincorporated area
including Skagway, Haines, and Angoon - too broad, not useful.

The current database is written in Microsoft Access version XX for
32-bit operating systems. Software and machine operating systems have
advanced enough for this to cause unnecessary burdens for accessing and
importing the data for analysis and reporting in common statistical
analysis environments (e.g., R). Recommend the Park request the ARO GIS
shop to convert the database to an uptodate software.

\emph{JOEL: Talk to Angie about what this would entail.}

\emph{Recommendations} work with GIS shop to clean up db, improve
reporting \& query support; shiny app; reproducible research scripts to
generate automatic reports? Shiny app to interface?

\section{Discussion}\label{discussion}

\section{Conclusion}\label{conclusion}

\section{Acknowledgements}\label{acknowledgements}

\subparagraph{pagebreak}\label{pagebreak}

\section{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-Anonymous2009}{}
Anonymous, (KLGO NHP). 2009. ``Waterbird Monitoring Protocol for
Klondike Gold Rush National Historical Park , Skagway , Alaska Standard
Operating Procedure ( SOP ) \# 3 Conducting the Waterbird Survey
Transect Revision History Log :'' April 9, 2009.

\hypertarget{ref-Fancy2009}{}
Fancy, Steven G., J. E. Gross, and S. L. Carter. 2009. ``Monitoring the
condition of natural resources in US national parks.''
\emph{Environmental Monitoring and Assessment} 151 (1-4): 161--74.
doi:\href{https://doi.org/10.1007/s10661-008-0257-y}{10.1007/s10661-008-0257-y}.

\hypertarget{ref-Hahr2004}{}
Hahr, Meg, Todd W Trapp, Meg Hahr, and Todd W Trapp. 2004. ``Waterbird
and Breeding Landbird Inventories in Klondike Gold Rush National
Historical Park Waterbird and Breeding Landbird Inventories in Klondike
Gold Rush National Historical Park Final Report,'' no. May.

\hypertarget{ref-Mitchell2018}{}
Mitchell, Brian, Alice Chung-MacCoubrey, Jim Comiskey, Lisa Garrett,
Maggie MacCluskie, Bill Moore, Tom Philippie, Geoff Sanders, and John
Paul Schmit. 2018. ``Inventory and Monitoring Division Protocol Review
Guidance.'' Fort Collins, CO.

\hypertarget{ref-NABCI2000}{}
(NABCI), North American Bird Conservation Initiative. 2000. ``North
American Bird Conservation Initiative Bird Conservation Region Map and
Descriptions.'' Arlington, VA.
\url{http://nabci-us.org/resources/bird-conservation-regions/}.

\hypertarget{ref-Reynolds2012}{}
Reynolds, Joel H. 2012. ``An overview of statistical considerations in
long-term monitoring.'' In \emph{Design and Analysis of Long-Term
Ecological Monitoring Studies}, 23--53.
doi:\href{https://doi.org/10.1017/CBO9781139022422.005}{10.1017/CBO9781139022422.005}.

\hypertarget{ref-Reynolds2016}{}
Reynolds, Joel H., Melinda G. Knutson, Ken B. Newman, Emily D.
Silverman, and William L. Thompson. 2016. ``A road map for designing and
implementing a biological monitoring program.'' \emph{Environmental
Monitoring and Assessment} 188 (7). Environmental Monitoring;
Assessment: 1--25.
doi:\href{https://doi.org/10.1007/s10661-016-5397-x}{10.1007/s10661-016-5397-x}.

\hypertarget{ref-St.Pierre2015}{}
St. Pierre, Mallory (KLGO NHP). 2015. ``2014 Bird Surveys at Klondike
Gold Rush National Historical Park.'' Fort Collins: U.S. Department of
Interior, National Park Service, NRSS.

\hypertarget{ref-Surdyk2017}{}
Surdyk, Shelby L. (KLGO NHP). 2017. ``Bird Surveys at Klondike Gold Rush
NHP: 2015 summary.''

\hypertarget{ref-Surdyk2018}{}
Surdyk, Shelby L. (KLGO NHP), and S. A. (KLGO NHP) Evans. 2018. ``Bird
Surveys at Klondike Gold Rush NHP Bird Surveys at Klondike Gold Rush
NHP: 2016 summary.'' Fort Collins: National Park Service.

\hypertarget{ref-Wu2018}{}
Wu, J X, C B Wilsey, L Taylor, and G W Schuurman. 2018. ``Projected
avifaunal responses to climate change across the U.S. National Park
System.'' \emph{PLoS ONE}.
doi:\href{https://doi.org/10.1371/journal.pone.0190557}{10.1371/journal.pone.0190557}.

\subparagraph{pagebreak}\label{pagebreak-1}

\subsubsection{Colophon}\label{colophon}

This report was generated on 2018-12-10 13:33:40 using the following
computational environment and dependencies:

\begin{verbatim}
#>  setting  value                       
#>  version  R version 3.5.1 (2018-07-02)
#>  system   x86_64, linux-gnu           
#>  ui       X11                         
#>  language (EN)                        
#>  collate  en_US.UTF-8                 
#>  tz       EST5EDT                     
#>  date     2018-12-10                  
#> 
#>  package    * version date       source        
#>  assertthat   0.2.0   2017-04-11 CRAN (R 3.3.1)
#>  backports    1.1.2   2017-12-13 CRAN (R 3.5.0)
#>  base       * 3.5.1   2018-09-10 local         
#>  bindr        0.1.1   2018-03-13 CRAN (R 3.5.0)
#>  bindrcpp   * 0.2.2   2018-03-29 CRAN (R 3.5.0)
#>  colorspace   1.3-2   2016-12-14 CRAN (R 3.5.0)
#>  compiler     3.5.1   2018-09-10 local         
#>  crayon       1.3.4   2017-09-16 CRAN (R 3.3.1)
#>  datasets   * 3.5.1   2018-09-10 local         
#>  DBI          1.0.0   2018-05-02 CRAN (R 3.5.0)
#>  dbplyr     * 1.2.2   2018-07-25 CRAN (R 3.5.0)
#>  devtools     1.13.6  2018-06-27 CRAN (R 3.5.0)
#>  digest       0.6.17  2018-09-12 CRAN (R 3.5.0)
#>  dplyr      * 0.7.6   2018-06-29 CRAN (R 3.5.0)
#>  evaluate     0.11    2018-07-17 CRAN (R 3.5.0)
#>  farver       1.1.0   2018-11-20 CRAN (R 3.5.1)
#>  ggforce    * 0.1.3   2018-07-07 CRAN (R 3.5.1)
#>  ggplot2    * 3.0.0   2018-07-03 CRAN (R 3.5.0)
#>  ggthemes   * 4.0.1   2018-08-24 CRAN (R 3.5.1)
#>  glue         1.3.0   2018-07-17 CRAN (R 3.5.0)
#>  graphics   * 3.5.1   2018-09-10 local         
#>  grDevices  * 3.5.1   2018-09-10 local         
#>  grid         3.5.1   2018-09-10 local         
#>  gtable       0.2.0   2016-02-26 CRAN (R 3.5.0)
#>  hms          0.4.2   2018-03-10 CRAN (R 3.5.0)
#>  htmltools    0.3.6   2017-04-28 CRAN (R 3.5.0)
#>  knitr        1.20    2018-02-20 CRAN (R 3.5.0)
#>  lazyeval     0.2.1   2017-10-29 CRAN (R 3.5.0)
#>  lubridate  * 1.7.4   2018-04-11 CRAN (R 3.5.0)
#>  magrittr     1.5     2014-11-22 CRAN (R 3.1.2)
#>  MASS         7.3-50  2018-04-30 CRAN (R 3.5.1)
#>  memoise      1.1.0   2017-04-21 CRAN (R 3.4.3)
#>  methods    * 3.5.1   2018-09-10 local         
#>  munsell      0.5.0   2018-06-12 CRAN (R 3.5.0)
#>  pillar       1.3.0   2018-07-14 CRAN (R 3.5.0)
#>  pkgconfig    2.0.2   2018-08-16 CRAN (R 3.5.0)
#>  plyr         1.8.4   2016-06-08 CRAN (R 3.5.0)
#>  purrr        0.2.5   2018-05-29 CRAN (R 3.5.0)
#>  R6           2.2.2   2017-06-17 CRAN (R 3.3.1)
#>  Rcpp         0.12.18 2018-07-23 CRAN (R 3.5.0)
#>  readr      * 1.1.1   2017-05-16 CRAN (R 3.5.0)
#>  rlang        0.2.2   2018-08-16 CRAN (R 3.5.0)
#>  rmarkdown    1.10    2018-06-11 CRAN (R 3.5.0)
#>  rprojroot    1.3-2   2018-01-03 CRAN (R 3.5.0)
#>  scales       1.0.0   2018-08-09 CRAN (R 3.5.0)
#>  stats      * 3.5.1   2018-09-10 local         
#>  stringi      1.2.4   2018-07-20 CRAN (R 3.5.0)
#>  stringr    * 1.3.1   2018-05-10 CRAN (R 3.5.0)
#>  tibble       1.4.2   2018-01-22 CRAN (R 3.5.0)
#>  tidyr      * 0.8.1   2018-05-18 CRAN (R 3.5.0)
#>  tidyselect   0.2.4   2018-02-26 CRAN (R 3.5.0)
#>  tools        3.5.1   2018-09-10 local         
#>  tweenr       1.0.0   2018-09-27 CRAN (R 3.5.1)
#>  units        0.6-0   2018-06-09 CRAN (R 3.5.0)
#>  utils      * 3.5.1   2018-09-10 local         
#>  withr        2.1.2   2018-03-15 CRAN (R 3.5.0)
#>  yaml         2.2.0   2018-07-25 CRAN (R 3.5.0)
\end{verbatim}

The current Git commit details are:

\end{document}

### 2019-01-05 analysis/paper/paper.tex
changed files for readability

@@
   \postdate{\par}
-    \date{10 December, 2018}
+    \date{11 December, 2018}
 
@@
 
-This report was generated on 2018-12-10 13:33:40 using the following
+This report was generated on 2018-12-11 18:15:51 using the following
 computational environment and dependencies:
@@
 #>  tz       EST5EDT                     
-#>  date     2018-12-10                  
+#>  date     2018-12-11                  
 #> 

### 2019-01-05 analysis/scripts/data_exploration.R

@@ -55,7 +55,7 @@ species_data_2007 <- left_join(tbl_Field_Data, tbl_Events,
 #ifelse(species_data_2007$`Num_ Observed` < mean(species_data_2007$`Num_ Observed`), "below", "above")
 
 #faceted by species?
-ggplot(species_data_2007,
+plot1 <- ggplot(species_data_2007 %>% ungroup() %>% filter(Species < "AMPI"),
        aes(as.factor(Days_since_first),
            `Num_ Observed`)) +
   geom_bar(stat='identity', aes(fill = as.factor(Year))) +
@@ -63,13 +63,15 @@ ggplot(species_data_2007,
   #paginate
   facet_wrap_paginate(~Species, ncol = 1)
 
+print(plot1)
+
 #another way to paginate
 #for (i in seq(1, length(unique(species_data_2007$Species)), 6)) {
  # print(plot)
 #}
 
 #first count every year?
-ggplot(species_data_2007 %>%
+plot2 <- ggplot(species_data_2007 %>%
          ungroup() %>%
          group_by(Species, Year) %>%
          arrange(Days_since_first) %>%
@@ -79,13 +81,15 @@ ggplot(species_data_2007 %>%
   #using facet_wrap_paginate() from ggforce
   #to distribute over pages
   facet_wrap_paginate(~Species, ncol = 1)
+print(plot2)
 
 #time series?
 #dot plot
-ggplot(species_data_2007, aes(as.factor(Year), `Num_ Observed`)) +
+plot3 <- ggplot(species_data_2007, aes(as.factor(Year), `Num_ Observed`)) +
   geom_boxplot() +
   geom_dotplot(#binaxis='y',
                #stackdir='center',
                #dotsize = .5,
                #fill="red"
     )
+print(plot3)

### 2019-01-05 analysis/scripts/units_species.R

@@ -31,23 +31,25 @@ species_units <-
   #spread(GIS_Location_ID, total_Observed)
 
 #to do: make plot
-ggplot(species_units, aes(as.numeric(Days_since_first),
+plot4 <- ggplot(species_units, aes(as.numeric(Days_since_first),
                           total_Observed,
                           color = as.factor(Year))) +
   geom_jitter(height = .3, aes(fill = GIS_Location_ID)) +
   #in ggforce package to supposedly distribute facets over pages
   facet_wrap_paginate(~Species, ncol = 1, scales = "free") +
   ylim(c(0, 2000))
+print(plot4)
 
 #how to
   #1. order bars with a y aesthetic,
   #2. have species not take up so much space
   #3. get rid of 0/NA values to take up less space?? ("free_x" not working)
-ggplot(species_units, aes(Species, total_Observed, fill = as.factor(Year))) +
+plot5 <- ggplot(species_units, aes(Species, total_Observed, fill = as.factor(Year))) +
   geom_bar(stat = "identity") +
   coord_flip() +
   #in ggforce package to supposedly distribute facets over pages
   facet_wrap_paginate(~GIS_Location_ID, ncol = 2, scales = "free")
+print(plot5)
 #bar chart:
   #faceted by unit (8) and by non-zero species?
   #x = all species(?), could be separated by year

### 2019-02-06 analysis/scripts/data_retrievalCSV.R
Corrected relative paths to source'd files.

@@ -6,7 +6,11 @@
 # Last Edit: 2018 Oct 17
 ########
 
-path = "./analysis/data/raw_data/CSVs/"
+print(getwd()) # JR added 2019 Feb 5 to help traceback
+
+#path = "./analysis/data/raw_data/CSVs/"
+path = "../data/raw_data/CSVs/"  #JR corrected 2019 Feb 5
+
 
 #retrieve file names with full path included
 file_names <- dir(path, full.names = TRUE)

### 2019-02-07 analysis/scripts/data_exploration.R
Edit of 'data_exploration.R' to explicitly 'print' ggplot objects so they appear in knitr rendering.

@@ -39,7 +39,7 @@ species_data <- left_join(tbl_Field_Data, tbl_Events,
   ungroup() #%>%
 #arrange()
 
-p1 <- ggplot(species_data, aes(Common_Name, Month_Day)) + #store ggplot as plot object
+plot1<- ggplot(species_data, aes(Common_Name, Month_Day)) + #store ggplot as plot object
   geom_point(col="tomato2", size=3) +   # Draw points
   geom_segment(aes(x=Common_Name,
                    y=min(Month_Day),
@@ -50,6 +50,7 @@ p1 <- ggplot(species_data, aes(Common_Name, Month_Day)) + #store ggplot as plot 
        subtitle="Common Name vs. Sighting") +
   coord_flip()
 
+print(plot1) # JR explicitly render
 
 
 species_data_2007 <- left_join(tbl_Field_Data, tbl_Events,

### 2019-02-09 analysis/paper/paper.Rmd
Ran into some errors with the CSV general retrial script; will look into it
Finally fixed the error that was blocking the retrieval script. Script now collects all .csv files as data frames for future use.
Separated functionality in 'data_exploration.R' into a 'data_wrangling' script and a 'survey_timing' data visualization script. Modified 'data_retrievalCSV' to suppress messages. Modifications to r code chunks to get knitr to compile to html, word, or pdf.

@@
 author:
- - name: Joel H. Reynolds^[U.S. National Park Service, Alaska Region, Natural Resources Science Team]
+- name: Joel H. Reynolds^[U.S. National Park Service, Alaska Region, Natural Resources
+    Science Team]
 - name: Madeleine Ward^[Swarthmore College, Dept. of Mathematics and Statistics]
@@
   word_document:
-  toc: TRUE
-#  number_sections: TRUE
+    toc: yes
+  html_document:
+    df_print: paged
+    toc: yes
+  pdf_document:
+    toc: yes
+keywords: keyword 1; keyword 2; keyword 3
+link-citations: yes
 bibliography: KLGOCWS_References.bib
-link-citations: yes
-
-abstract: The Coastal Waterbird Survey has been conducted, with varying levels of effort and in-season frequency, since 2003 [@Surdyk2018].
-The area of interest has been divided into eight contiguous survey units; on each survey, at each unit the number, age, composition (sex), breeding status, and noteworthy behavioral observations are recorded for each species [@Anonymous2009].
-Given the constraints on personnel, time, and budgets, it is imperative that surveys be conducted efficiently in terms of both choice of sampling design and choice of sampling effort.
-Now that KLGO has multiple years of experience implementing the protocol, it is timely to step back and assess how things are going.
-We review the current survey program’s objectives,  design, identify statistical issues, and recommend potential solutions.
-Major topics include ….
-The study concludes with a series of recommended tasks that the park should undertake to improve the coastal waterbird survey’s efficiency and effectiveness.
-
-keywords:
- keyword 1; keyword 2; keyword 3
+abstract: The Coastal Waterbird Survey has been conducted, with varying levels of
+  effort and in-season frequency, since 2003 [@Surdyk2018]. The area of interest has
+  been divided into eight contiguous survey units; on each survey, at each unit the
+  number, age, composition (sex), breeding status, and noteworthy behavioral observations
+  are recorded for each species [@Anonymous2009]. Given the constraints on personnel,
+  time, and budgets, it is imperative that surveys be conducted efficiently in terms
+  of both choice of sampling design and choice of sampling effort. Now that KLGO has
+  multiple years of experience implementing the protocol, it is timely to step back
+  and assess how things are going. We review the current program's objectives and survey design,
+  identify statistical issues, and recommend potential solutions. Major topics include... blah blah. The study concludes with a series of recommended tasks that the park should
+  undertake to improve the coastal waterbird survey's efficiency and effectiveness.
 ---
@@
 
-```{r setup, include=FALSE}
+```{r setup, include=FALSE, cache = FALSE}
 knitr::opts_chunk$set(
   collapse = TRUE,
-  warning = FALSE,
-  message = FALSE,
+  warning = TRUE, # set to FALSE when done
+  message = TRUE, # set to FALSE when doen
   echo = FALSE,
   comment = "#>",
-  fig.path = "../figures/"
+  dev = c('png','pdf'),
+  pdf.options(encoding = "ISOLatin9.enc"),
+  fig.path = "../figures/",
+  error = FALSE
 )
 
-library(dply);library(ggplot2);library(dbplyr);
+library(dplyr);library(ggplot2);library(dbplyr);library(readr);library(stringr);library(tidyr);library(lubridate);library(lattice);library(ggthemes);library(ggforce)
 
 ```
+# Things To Discuss with Madeleine
+- add all libraries used in scripts to 'setup' code chunk (just above) rather than just loading in script so that it is easy to see what needs to be loaded before you can successfully knit.
+- documentation in scripts
+
+# Issues to Resolve
+- why isn't code chunk option warning= FALSE suppressing warnings() in Data_Retrieval? (from read_csv)
+- How do we use latex figure environment in rmarkdown (for inclusion of ggplot graphics; captioning figures and labelling for cross-references, etc)?
+- conversion from paper.md to paper.pdf (pandoc issue?) (i) update MikTex package, set dev=c('pdf') in setup
+- github permissions for Madeleine on local machine
+- What is up with 2008 surveys? why such distinct timing?
 
@@
 This is sufficient history to warrant stepping back and assessing the monitoring program.
-Such assessments are key to an adaptive monitoring program [@Lindenmeyer 2009] and are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016].
+Such assessments are key to an adaptive monitoring program [@Lindenmeyer 2009], are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016], and align with guidance from NPS' Inventory and Monitoring Division [@Gallo2018a, @Mitchell2018].
 
@@
 If so, can the data collection be made more efficient, reducing the observation burden on staff resources?
-3. _how_ - [Data Analysis Design] Do the data analysis methods effectively address the information objectives? Are more informative analyses available given interim improvements in methods?  
+3.
+_how_ - [Data Analysis Design] Do the data analysis methods effectively address the information objectives?
+Do the data analysis methods properly account for impactful features of the data collection design?
+Are more informative analyses available given interim improvements in methods?
 4.
@@
 This assessment addressed all four topics, but focused predominantly on topics (i) Problem Framing & Information Objectives, (ii) Data Collection Design, and (iii) Data Analysis Design.
-Since the former topic (Problem Framing & Information Objectives) determines the relevancy of many of the potential assessment actions related to the latter topics (Data Collection and Analysis Designs, Workflows & Reporting), therefore  the assessment was drafted as a first stage of an iterative effort, providing high level feedback, including listings of potential detailed syntheses and workflow enhancement activities (such as improved database interface, automation of annual analyses and updating of syntheses, etc.).
-With this foundation, further dialogue with KLGO staff regarding relevant Problem Framing & Information Objectives and other priorities will determine which follow-on activities to pursue.
-
-<!-- relation to existing NPS I&M methods & formats? -->
+Since 'Problem Framing & Information Objectives' determine the relevancy of many of the potential assessment actions related to the other topics,  the assessment was drafted as a first stage of an iterative effort, providing high level feedback, including listings of potential detailed syntheses and workflow enhancement activities (such as improved database interface, automation of annual analyses and updating of syntheses, etc.).
+With this foundation, further dialogue with KLGO staff regarding relevant 'Problem Framing & Information Objectives' and other priorities will determine which follow-on activities to pursue.
 
@@
 # Methods
-This assessment compared the CWS monitoring program against a recommended checklist of monitoring program components (Appendix 1).
-The checklist was developed from published literature [@Reynolds2012; @Reynolds2016].
+The CWS monitoring program was compared against a recommended checklist of monitoring program components developed from published literature (Appendix 1).
 It identifies multiple component decisions associated with each of the major Phases of a monitoring program (briefly overviewed above).
@@
 ### Assessment findings:
-The 2003 coastal waterbird inventory and survey design were developed to address a top priority bird inventory need of the Southeast Alaska Inventory and Monitoring Network (Sharman et al.
+The 2003 coastal waterbird inventory and survey design was developed to address "a top priority bird inventory (need)" of the Southeast Alaska Inventory and Monitoring Network (Sharman et al.
 2000) - namely, an inventory of waterfowl/shorebirds at KLGO [@Hahr2004].
-This was deemed a priority due to the area’s expected avifaunal richness (NABCI 2000) and the limited availability of reliable observation records - mainly consisting of systematic surveys conducted by NPS staff in the early 1980s [@Hahr2004].
+This was deemed a priority due to the area’s expected avifaunal richness [@NABCI2000] and the limited availability of reliable observation records - mainly consisting of systematic surveys conducted by NPS staff in the early 1980s [@Hahr2004].
 
-The goal of the inventory was to document the occurrence of 90% of the species of waterbirds likely to occur in KLGO (ibid).
-The principal objectives were “ to develop scientifically valid sampling designs for inventorying waterbirds and breeding landbirds in KLGO, and to collect baseline information on bird species occurrence, distribution, abundance, and habitat associations in the park…[to] ...provide park managers with timely information that facilitates land management planning, helps track wildlife population and ecosystem change over time, and keeps resource managers apprised of changes that may require action before a crisis occurs” (ibid).
+The inventory's stated goal was to document the occurrence of 90% of the species of waterbirds likely to occur in KLGO (ibid).
+The principal objectives were “to develop scientifically valid sampling designs for inventorying waterbirds and breeding landbirds in KLGO, and to collect baseline information on bird species occurrence, distribution, abundance, and habitat associations in the park…[to] ...provide park managers with timely information that facilitates land management planning, helps track wildlife population and ecosystem change over time, and keeps resource managers apprised of changes that may require action before a crisis occurs” (ibid).
+While no Threatened or Endangered bird species were known to occur in KLGO, as of 2004, nine species on the Alaska Watchlist had been documented in the park (ibid), thus suggesting potential priority information needs associated with monitoring for changes in those species occurrence and abundance.
+Potential secondary objectives mentioned include "investigations into the breeding ecology, phenology, reproductive success and population age structure of KLGO breeding birds" (ibid).
+Later reports mention additional objectives of estimating changes in species occurrence and composition [@St.Pierre2015], characterizing timing of peak spring waterbird migration and its relation to spring eulachon runs [@Surdyk2017].
+
+Stakeholders mentioned or implied in park reports include park superintendent and related managers, those charged with monitoring regional and internationally important bird habitats, the Skagway Bird Club, the Alaska Landbird Monitoring Survey and the Boreal Partners in Flight.
 
@@
 
-Hahr and Trapp (2004) Quotes:
-upper Taiya Inlet
-The park contains excellent breeding and migration habitat for landbirds and waterbirds, most notably the Taiya River riparian Sitka spruce-cottonwood forests and the Taiya River estuary.
-
-We recommend that bird species distributions, habitat associations, and abundances continue to be monitored in the park’s regionally and internationally important bird habitats, and that KLGO continue to contribute to the ALMS effort.
-
-The park should formalize a waterbird migration monitoring protocol for the Taiya River estuary and seek assistance from local area birders to expand this effort beyond the park boundary in order to encompass more of the upper Taiya Inlet.
-Many of the undocumented bird species on the park’s expected species list are waterbirds, most notably shorebirds.
-Monitoring of spring waterbird migration, especially during the brief period when shorebirds are present, is highly
-recommended.
-Adding additional bird species to the park’s list will be difficult because the remaining undocumented species are apparently uncommon, may occur in low numbers, or appear irregularly or accidentally in the park.
-
-There are no federally listed Threatened or Endangered bird species in KLGO; however, nine Alaska WatchList species have been documented in the park (Table 4).
-The WatchList identifies at-risk Alaskan bird populations that may be facing population decline, limited geographic range, habitat loss and other threats but are not already protected under the federal Endangered Species Act.
-
-Investigations into the breeding ecology, phenology, reproductive success, and population age structure of KLGO breeding birds would also yield valuable information on the status of these species, especially those identified on the Alaska WatchList.
+QUOTES
+Hahr and Trapp (2004):
 We intend to use the results of these inventories to create a revised bird checklist and an atlas database for the Taiya and Skagway River watersheds in collaboration with the Skagway Bird Club.
-The forthcoming Vital Signs Monitoring Program will be a great opportunity for KLGO natural resource managers to review the park’s nascent landbird inventory and monitoring program and determine its future direction.
-
-2014 St. Pierre (2015?) 
-+ new inventory additions
-+ spp composition
-
-2015 Surdyk (2016?)
-Long-term monitoring and baseline inventories help managers to make decisions that meet the National Park Service goal to maintain intact ecosystems in their natural state (Hahr and Trapp 2004).
-
-Timing of peak spring waterbird migration to coincide with eulochon run.
-
-2016 Surdyk & ??? (2017)
-better understand the relationship between waterbird migration and spring eulachon runs.
 
@@
 Though not explicitly characterized in the protocol or reports reviewed for this assessment (Table 1), the following target audiences and information users were identified through conversations with KLGO staff, partners, and other waterbird researchers and stakeholders in Alaska.
+(current)
+- park
+- Skagway Bird Club; compiles for addition to 'significant bird observations in SE AK', which gets compiled in to AK sumamries for 'North American Birds' (mainly spp lists)
+(potential)
+- BPIF
+- PSG
+- Audubon Alaska (for IDing Important Bird Areas - though require accurate density estimates)
+- Audubon North America - cliamte species distribution models (Nicole Michel, Brooke Bateman, etc.); esp if associated w/ repeat sites & effort data (date, time of day, duration, distance/area covered) - see Nicole email 6/7/2018.
+Watchlist of projected changes - [@Wu2018]
 
@@
 
-```{r DataInput, eval=TRUE, echo=FALSE, cache=TRUE}
-# if (system("git config user.email")=="mward2@swarthmore.edu") source("./scripts/data_retrievalSwarthmore.R")
+```{r Data_Retrieval, eval=TRUE, echo=FALSE, warning = FALSE}
+# if (system("git config user.email")=="mward2@swarthmore.edu") 
+#      source("./scripts/data_retrievalSwarthmore.R")
 # else source("./scripts/data_retrievalNPS.R")
 
-###########
-# MADELEINE WILL REPLACE WITH CODE TO VARIOUS TABLES FROM CSV FILES
-###########
+#NOTE: working directory when knitting markup is /analysis/paper. So relative pathway needs to be ../scripts etc.
+source("../scripts/data_retrievalCSV.R")
+
 ```
@@
 
+```{r Data_Wrangling, eval=TRUE, echo=FALSE, warning = TRUE}
+# reconstructs species_data tibble from tables read in by DataInput
+#
+source("../scripts/data_wrangling.R")
+
+```
+
 *Temporal* ??how define ‘season’??
@@
 *Madeleine later* Diagnostic analysis => check dates & unit of first non-zero counts each year for each spp (relative to dates of initiating surveys)
+
+```{r EDA_SurveyTiming, eval=TRUE, cache = FALSE, fig.show= 'asis'}
+ source("../scripts/survey_timing_EDA.R", echo=FALSE) 
+print(Plot1)
+
+```
 
@@
 (tie to Conceptual Model and Information Objectives)
-Associated analyses
+Associated analyses:
+
+Number of species sighted per year: 2003-2016 (source 2 found 2003-2009)
+.Waterbird and non-waterbird species
+.Positive association between species richness and year of survey, when excluding year 2014, which started later than usual. R-squared = .54.
+.Species richness within year: 2016
+.Individual and max. numbers of observed counts for each species for year 2016, with first and last sightings noted.
+.Peak abundance/species richness identified with observed species counts and individual sightings across surveys during 2016 season
+.Peak abundance (May 4) observed to be in tandem with spring euchalon (April 21-)
+.Species richness peaked May 4-11
+
+Observer bias: variation in number of surveys and length of monitoring season
+.2010: No seasonal technician; only sporadic surveys
+.2014: Surveys begun 1 month late (May 22)
+.Across years: observer turnover rate, varying experience
+
+Number of surveys across census units per year (2003-2009) & distribution of survey dates
+
+Percent of expected waterbird and breeding landbird species (174) confirmed (158; 91%)
+
+Diversity
+.Simpson’s Diversity Index (1/D) (2003-2009): heavily sensitive to species richness
+.Q statistic (2003-2009)
+.Simpson evenness measure: (1/D)/# species, where values approaching 1 signal more equal distribution (not sensitive to species richness)
+
+Comparisons across census units:
+.Unit characteristics
+.Area of unit (square km)
+.Shoreline habitat (meters)
+.Waterbird characteristics:
+.Average number of waterbirds per survey per square km
+.Diversity measures
+Analysis across dates each year:
+.Number of species; number of birds
+.Diversity measures
+
 Reporting schedule
@@
 # if your Rmd is not in analysis/paper/
-git2r::repository("../..")
+#git2r::repository("../..")
 ```

### 2019-02-09 analysis/scripts/survey_timing_EDA.R

@@ -1,0 +1,47 @@
+############
+# Basic data visualization of survey timing (from species_data object),
+# ultimately leading to further filtering of that object for later analyses.
+########
+# Libraries used: ggplot2,
+########
+# Joel Reynolds (repurposing code by Madeleine Ward)
+# Last Edit: 2019 Feb 8
+########
+
+# Display the dates surveys were conducted across each year
+Plot1<- species_data %>%
+  ggplot(aes(YearDay,Year)) +
+  geom_point(col="blue", size=2) +   # Draw points
+  geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
+             linetype=2, colour="grey") +
+  labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)",
+       caption="The annual dates of the first and last survey have both varied by over a month across the years.")
+
+####### Ideas to pursue later
+#######
+
+# another view showing frequency across years
+# first need to collapse across species, locations, etc....
+if (0) {
+  species_data %>%
+    ggplot(aes(YearDay)) +
+    geom_dotplot(col="blue") +   # Draw points
+    geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
+               linetype=2, colour="grey") +
+    labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)",
+         caption="The annual dates of the first and last survey have both varied by over a month across the years.")
+
+}
+
+if (0) {
+# figure out which dates observed any of the key species - maybe facet on each?
+# then plot an open circle over that survey date (so we see which surveys were bust)
+# build toward species specific phenology plots.
+  species_data %>%
+    ggplot(aes(YearDay,Year)) +
+    geom_point(col="blue", size=2) +   # Draw points
+    geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
+               linetype=2, colour="grey") +
+    labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)",
+         caption="The annual dates of the first and last survey have both varied by over a month across the years.")
+}

### 2019-02-09 analysis/scripts/UNUSED_data_exploration.R
Separated functionality in 'data_exploration.R' into a 'data_wrangling' script and a 'survey_timing' data visualization script. Modifications to r code chunks to get knitr to compile to html, word, or pdf.

@@ -40,17 +40,14 @@ species_data <- left_join(tbl_Field_Data, tbl_Events,
 #arrange()
 
 plot1<- ggplot(species_data, aes(Common_Name, Month_Day)) + #store ggplot as plot object
-  geom_point(col="tomato2", size=3) +   # Draw points
-  geom_segment(aes(x=Common_Name,
+  geom_point(col="tomato2", size=2) +   # Draw points
+  geom_segment(aes(x=Common_Name, xend=Common_Name,  #2019 Feb 7 JR added xend
                    y=min(Month_Day),
                    yend=max(Month_Day)),
                linetype="dashed",
                size=0.1) +   # Draw dashed lines
-  labs(title="Dot Plot",
-       subtitle="Common Name vs. Sighting") +
+  labs(title="Sightings by Species") +
   coord_flip()
-
-print(plot1) # JR explicitly render
 
 
 species_data_2007 <- left_join(tbl_Field_Data, tbl_Events,

### 2019-02-09 analysis/scripts/UNUSED_data_explorationJR.R

@@ -1,0 +1,83 @@
+############
+# This script uses libraries "dplyr", "lubridate", and "tidyr" to wrangle species data
+########
+# Madeleine Ward, Joel Reynolds
+# Last Edit: 2018 Oct 22; Updated: 2019 Feb 7
+########
+
+
+# re-build the necessary data frame from the various CSV files
+# (basically reconstructing the query if the database was functioning).
+
+# 1st left_join - attaches relevant Event data to each event in tbl_Field_Data (using Event_ID)
+species_data <- left_join(tbl_Field_Data, tbl_Events,
+                          by = "Event_ID") %>%
+  # 2nd lef_join - attaches relevant Species names (Common, Latin, etc.) (using Species Code)
+    left_join(tlu_species_codes,
+            by = c("Species" = "Code")) %>%
+  #subset out just nonzero observations after 2003
+  filter(`Num_ Observed` > 0, Year >= 2003) %>%
+  #create new Month_Day column & Julian Date column
+  mutate(Date = floor_date(mdy_hm(Start_Date), unit = "day"),
+         Month_Day = format(Date, "%m-%d"),
+         JulDate = julian(Date)-julian(floor_date(Date, unit = "year"))+1) %>%
+  #define grouping by Species
+  group_by(Species, Year, Month_Day) %>%
+  #add total observations column (by species by observation date)
+  mutate(total_obs = sum(`Num_ Observed`)) %>%
+  ungroup() %>%
+  group_by(Species, Year) %>%
+  filter(row_number()==1) %>%
+  ungroup() %>%
+  mutate(species_num = as.integer(as.factor(Species)))
+
+
+# TO DO:
+# filter to core field season (not all year)
+# use species code, winnow to core spp
+# use dotplot() directly
+
+plot1<- ggplot(species_data, aes(Common_Name, JulDate)) + #store ggplot as plot object
+  geom_point(col="tomato2", size=2) +   # Draw points
+  geom_segment(aes(x=Common_Name, xend=Common_Name,  #2019 Feb 7 JR added xend
+                   y=min(JulDate),
+                   yend=max(JulDate)),
+               linetype="dashed",
+               size=0.1) +   # Draw dashed lines
+  labs(title="Sightings by Species") +
+  coord_flip()
+
+print(plot1) # saves graphics file in analysis/figures/
+library(lattice)
[…]

### 2019-02-09 analysis/scripts/data_retrievalCSV.R

@@ -1,12 +1,12 @@
 # Retrieve CSV Data when working on swarthmore R server
 ############
-# This script uses library("readr") to access the comma-separated files associated with species data.
+# Read in the comma-separated files associated with species data.
+########
+# Libraries used: readr
 ########
 # Madeleine Ward
 # Last Edit: 2018 Oct 17
 ########
-
-print(getwd()) # JR added 2019 Feb 5 to help traceback
 
 #path = "./analysis/data/raw_data/CSVs/"
 path = "../data/raw_data/CSVs/"  #JR corrected 2019 Feb 5

### 2019-02-09 analysis/scripts/data_wrangling.R

@@ -1,0 +1,40 @@
+############
+# Create the survey & species data object "species_data".
+########
+# Libraries used: dplyr, lubridate, tidyr
+########
+# Joel Reynolds (repurposing code by Madeleine Ward)
+# Last Edit: 2019 Feb 8
+########
+
+
+# re-build the necessary data frame from the various CSV files
+# (basically reconstructing the query as if the database was functioning).
+
+# 1st left_join - attaches relevant Event data to each event in tbl_Field_Data (using Event_ID)
+species_data <- left_join(tbl_Field_Data, tbl_Events,
+                          by = "Event_ID") %>%
+  # 2nd lef_join - attaches relevant Species names (Common, Latin, etc.) (using Species Code)
+  left_join(tlu_species_codes,
+            by = c("Species" = "Code")) %>%
+  #subset out just columns we care about, re-order by interest, rename where necessary
+  select(Start_Date, Year, Month, Day, Start_Time, End_Time, Duration,
+         Location_ID,
+         Species, ID, Category, Family, Common_Name, Latin_Name, Latin_Code,
+         Num_Obs = 'Num_ Observed',
+         Temp, Cloud_Cover, Precip_Code,
+         Wind_code, Wind_dir, Wave_ht, Tide, Tide_Level, Tide_State,
+         Event_Group_ID, Protocol_Name, Observer, Event_Notes,
+         Data_ID, Event_ID, Data_Location_ID) %>%
+  #subset out just nonzero observations after 2003 (using renamed 'Num_Obs')
+  filter(Num_Obs > 0, Year >= 2003) %>%
+  #strip HMS from Start_Date, create new Julian date column ('YearDay') & Fractional Month_Day
+  mutate(Date = floor_date(mdy_hm(Start_Date), unit = "day"),
+         YearDay = yday(Date)) %>%
+  #         Month_DayFrac = format(Date, "%m-%d"),
+  #define grouping by Species
+  group_by(Species, Year, YearDay) %>%
+  #add total observations column (by species by observation date)
+  mutate(total_obs = sum(Num_Obs)) %>%
+  ungroup()
+

### 2019-02-09 analysis/paper/paper.Rmd
Separated functionality in 'data_exploration.R' into a 'data_wrangling' script and a 'survey_timing' data visualization script. Modifications to r code chunks to get knitr to compile to html, word, or pdf.

@@
 # Issues to Resolve
-- why isn't code chunk option warning= FALSE suppressing warnings() in Data_Retrieval? (from read_csv)
 - How do we use latex figure environment in rmarkdown (for inclusion of ggplot graphics; captioning figures and labelling for cross-references, etc)?
-- conversion from paper.md to paper.pdf (pandoc issue?) (i) update MikTex package, set dev=c('pdf') in setup
+- conversion from paper.md to paper.pdf (pandoc issue?) 
+  (i) update MikTex package, set dev=c('pdf') in setup
 - github permissions for Madeleine on local machine
+- adding files to .gitignore
 - What is up with 2008 surveys? why such distinct timing?

### 2019-02-13 analysis/scripts/time_duration.R
test

@@ -1,6 +1,6 @@
 # title
 ############
-# This script examines the time duration at each site by day and year
+# This script examines the time duration at each site by day and year.
 ########
 # Madeleine Ward
 # Last Edit: 2018 Oct 30

### 2019-02-13 analysis/paper/paper.Rmd
minor additions to *.bib file and .gitignore.

@@
 output:
-  word_document:
-    toc: yes
   html_document:
@@
     toc: yes
+  word_document:
+    toc: yes
+    # reference_docx: NRSS_template_2018July.docx get NRSS report template from Paul Burger
 keywords: keyword 1; keyword 2; keyword 3
@@
   undertake to improve the coastal waterbird survey's efficiency and effectiveness.
+editor_options: 
+  chunk_output_type: inline
 ---
@@
 - conversion from paper.md to paper.pdf (pandoc issue?) 
-  (i) update MikTex package, set dev=c('pdf') in setup
-- github permissions for Madeleine on local machine
 - adding files to .gitignore
@@
 This is sufficient history to warrant stepping back and assessing the monitoring program.
-Such assessments are key to an adaptive monitoring program [@Lindenmeyer 2009], are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016], and align with guidance from NPS' Inventory and Monitoring Division [@Gallo2018a, @Mitchell2018].
+Such assessments are key to an adaptive monitoring program [@Lindenmeyer2009], are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016], and align with guidance from NPS' Inventory and Monitoring Division [@Gallo2018a, @Mitchell2018].
 

### 2019-02-13 analysis/paper/paper.Rmd
Fix some biblio errors; initial tests of figure labels and captions.

--- 2019-02-09 analysis/paper/paper.Rmd
@@
     toc: yes
+    fig_caption: true
+  word_document:
+    toc: yes
+    fig_caption: true
+    reference_docx: NRR_Author_Template_V4.2.docx
   pdf_document:
     toc: yes
+    fig_caption: true
 keywords: keyword 1; keyword 2; keyword 3
@@
 This is sufficient history to warrant stepping back and assessing the monitoring program.
-Such assessments are key to an adaptive monitoring program [@Lindenmeyer 2009], are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016], and align with guidance from NPS' Inventory and Monitoring Division [@Gallo2018a, @Mitchell2018].
+Such assessments are key to an adaptive monitoring program [@Lindenmayer2009], are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016], and align with guidance from NPS' Inventory and Monitoring Division [@Gallo2018a, @Mitchell2018].
 
@@
 *Madeleine later* Diagnostic analysis => check dates & unit of first non-zero counts each year for each spp (relative to dates of initiating surveys)
+The annual dates of the first and last survey each year has varied by over a month (Figure \ref{fig:Figure_SurveyTiming}).
 
-```{r EDA_SurveyTiming, eval=TRUE, cache = FALSE, fig.show= 'asis'}
+```{r EDA_SurveyTiming, eval=TRUE, cache = TRUE, include = FALSE}
+# dotplot of survey times by year
 source("../scripts/survey_timing_EDA.R", echo=FALSE)
-print(Plot1)
 
@@
 
+```{r Figure_SurveyTiming, echo=FALSE, eval=TRUE, cache=TRUE, dependson= c("EDA_SurveyTiming"), fig.show= 'asis', fig.cap="\\label{fig:Figure_SurveyTiming}The annual dates of the first and last survey have both varied by over a month across the years."}
+
+print(Plot1)
+```
 *Sample Frame* 

### 2019-02-13 analysis/paper/paper.Rmd
Fix some biblio errors; initial tests of figure labels and captions.

--- 2019-02-09 analysis/paper/paper.Rmd
@@
 output:
-  word_document:
+  bookdown::word_document2:
+    fig_caption: yes
     toc: yes
@@
     df_print: paged
+    fig_caption: yes
+    number_sections: yes
     toc: yes
   pdf_document:
+    fig_caption: yes
     toc: yes
@@
   multiple years of experience implementing the protocol, it is timely to step back
-  and assess how things are going. We review the current program's objectives and survey design,
-  identify statistical issues, and recommend potential solutions. Major topics include... blah blah. The study concludes with a series of recommended tasks that the park should
-  undertake to improve the coastal waterbird survey's efficiency and effectiveness.
+  and assess how things are going. We review the current program's objectives and
+  survey design, identify statistical issues, and recommend potential solutions. Major
+  topics include... blah blah. The study concludes with a series of recommended tasks
+  that the park should undertake to improve the coastal waterbird survey's efficiency
+  and effectiveness.
 ---
@@
 
-# Issues to Resolve
-- How do we use latex figure environment in rmarkdown (for inclusion of ggplot graphics; captioning figures and labelling for cross-references, etc)?
+# Issues to Resolve with Rmarkdown, Knitr, Github, etc.
 - conversion from paper.md to paper.pdf (pandoc issue?) 
@@
 - adding files to .gitignore
+
+# Issues to Resolve with content (narrative, figures, analyses)
 - What is up with 2008 surveys? why such distinct timing?
@@
 
-```{r Data_Retrieval, eval=TRUE, echo=FALSE, warning = FALSE}
+```{r Data-Retrieval, eval=TRUE, echo=FALSE, warning = FALSE}
 # if (system("git config user.email")=="mward2@swarthmore.edu") 
@@
 
-```{r EDA_SurveyTiming, eval=TRUE, cache = FALSE, fig.show= 'asis'}
+The annual dates of the first and last survey each year has varied by over a month (Figure \@ref(fig:Fig-SurveyTiming)).
+
+```{r EDA_SurveyTiming, eval=TRUE, cache = TRUE, include = FALSE}
+# dotplot of survey times by year
 source("../scripts/survey_timing_EDA.R", echo=FALSE)
+
+```
+
+```{r Fig-SurveyTiming, echo=FALSE, eval=TRUE, cache=TRUE, dependson= c("EDA_SurveyTiming"), fig.show= 'asis', fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
+
 print(Plot1)
-
 ```

### 2019-02-16 analysis/paper/paper.Rmd

@@
     toc: yes
+  word_document:
+    toc: yes
+    fig_caption: true
+    reference_docx: NRR_Author_Template_V4.2.docx
   pdf_document:
@@
     toc: yes
+    fig_caption: true
 keywords: keyword 1; keyword 2; keyword 3

### 2019-02-16 analysis/paper/paper.Rmd

--- 2019-02-13 analysis/paper/paper.Rmd
@@
     toc: yes
+editor_options: 
+  chunk_output_type: inline
 keywords: keyword 1; keyword 2; keyword 3
@@
   multiple years of experience implementing the protocol, it is timely to step back
-  and assess how things are going. We review the current program's objectives and
-  survey design, identify statistical issues, and recommend potential solutions. Major
-  topics include... blah blah. The study concludes with a series of recommended tasks
-  that the park should undertake to improve the coastal waterbird survey's efficiency
-  and effectiveness.
+  and assess how things are going. We review the current program's objectives and survey design,
+  identify statistical issues, and recommend potential solutions. Major topics include... blah blah. The study concludes with a series of recommended tasks that the park should
+  undertake to improve the coastal waterbird survey's efficiency and effectiveness.
 ---

### 2019-02-16 analysis/paper/paper.Rmd
Removing ggplot caption in Survey Timing figure; correcting cache option in EDA_SurveyTiming chunk (don't cache since won't see any changes in externally sourced script).

@@
 
-```{r EDA_SurveyTiming, eval=TRUE, cache = TRUE, include = FALSE}
+```{r EDA_SurveyTiming, eval=TRUE, cache = FALSE, include = FALSE}
 # dotplot of survey times by year
@@
 
-```{r Fig-SurveyTiming, echo=FALSE, eval=TRUE, cache=TRUE, dependson= c("EDA_SurveyTiming"), fig.show= 'asis', fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
+```{r Fig-SurveyTiming, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
 

### 2019-02-20 analysis/scripts/UNUSED_data_exploration.R
Staged all my new inputs from local version. Effected scripts are the unused ones.

@@ -15,63 +15,35 @@ species_data <- left_join(tbl_Field_Data, tbl_Events,
          Month_Day = format(Date, "%m-%d")) %>%
   filter(`Num_ Observed` > 0, Year >= 2003) %>%
   group_by(Species, Year, Month_Day) %>%
-  mutate(total_obs = sum(`Num_ Observed`)) #%>%
-  #ungroup() %>%
-  #group_by(Species, Year) %>%
-  #filter(row_number()==1) %>%
-  #ungroup() #%>%
-  #arrange()
-
-species_data$species_num= as.integer(as.factor(species_data$Species))
-species_data <- left_join(tbl_Field_Data, tbl_Events,
-                          by = "Event_ID") %>%
-  left_join(tlu_species_codes,
-            by = c("Species" = "Code")) %>%
-  mutate(Date = substr(Start_Date, 1, 8),
-         Date = mdy(Date),
-         Month_Day = format(Date, "%m-%d")) %>%
-  filter(`Num_ Observed` > 0, Year >= 2003) %>%
-  group_by(Species, Year, Month_Day) %>%
   mutate(total_obs = sum(`Num_ Observed`)) %>%
-  ungroup() %>%
-  group_by(Species, Year) %>%
-  filter(row_number()==1) %>%
   ungroup() #%>%
+#ungroup() %>%
+#group_by(Species, Year) %>%
+#filter(row_number()==1) %>%
+#ungroup() #%>%
 #arrange()
 
-plot1<- ggplot(species_data, aes(Common_Name, Month_Day)) + #store ggplot as plot object
-  geom_point(col="tomato2", size=2) +   # Draw points
-  geom_segment(aes(x=Common_Name, xend=Common_Name,  #2019 Feb 7 JR added xend
-                   y=min(Month_Day),
-                   yend=max(Month_Day)),
-               linetype="dashed",
-               size=0.1) +   # Draw dashed lines
-  labs(title="Sightings by Species") +
-  coord_flip()
-
-
-species_data_2007 <- left_join(tbl_Field_Data, tbl_Events,
-                               by = "Event_ID") %>%
-  left_join(tlu_species_codes,
-            by = c("Species" = "Code")) %>%
-  separate(Start_Date, c("Date", "Time_0"), sep = " ") %>%
-  separate(Start_Time, c("Date_01", "Time_Start"), sep = " ") %>%
-  #how to fit all these into one argument??
[…]

### 2019-02-20 analysis/scripts/data_retrievalCSV.R

@@ -11,7 +11,6 @@
 #path = "./analysis/data/raw_data/CSVs/"
 path = "../data/raw_data/CSVs/"  #JR corrected 2019 Feb 5
 
-
 #retrieve file names with full path included
 file_names <- dir(path, full.names = TRUE)
 

### 2019-02-20 analysis/scripts/time_duration.R

@@ -6,22 +6,71 @@
 # Last Edit: 2018 Oct 30
 ########
 
+# title
+############
+# This script examines the time duration at each site by day and year
+########
+# Madeleine Ward
+# Last Edit: 2018 Oct 30
+########
+
 #TO DO: ACCOUNT FOR DECIMAL FORMATS -
 #if/else
 #ex. some problematic times below
-new <- species_datafrom_Excel[375:419,]
+new <- tbl_Field_Data[375:419,]
 
-times <- species_datafrom_Excel$Begin
+times <- tbl_Field_Data$Begin
 times[377]
 times_converted <- sub("(\\d+)(\\d{2})", "\\1:\\2", times)
 times_converted
 
-#this code finds duration time-wise for dates/units
-time_duration <- species_datafrom_Excel %>%
-  distinct(Date, `Survey Unit`, .keep_all = TRUE) %>%
-  mutate(start_time = sub("(\\d+)(\\d{2})", "\\1:\\2", Begin),
-         end_time = sub("(\\d+)(\\d{2})", "\\1:\\2", End))#,
-         #duration = difftime(end_time, start_time,
-          #                   units = c("mins")))
+library(chron)
+#x <- chron(times=time)
 
- # select(Date, `Survey Unit`, )
+#this code finds duration time-wise for dates/units
+time_duration <- tbl_Field_Data %>%
+  left_join(tbl_Events,
+            by = "Event_ID") %>%
+  left_join(tbl_Locations,
+            by = "Location_ID") %>%
+  mutate(time_start = substr(Start_Time, 10, 17),
+         time_end = substr(End_Time, 10, 17),
+         Date = substr(Start_Date, 1, 8),
+         Date = mdy(Date),
+         time_start = chron(times=time_start),
+         time_end = chron(times=time_end)) %>%
+  distinct(Date, GIS_Location_ID, .keep_all = TRUE) %>%
+  filter(GIS_Location_ID == "Site 1")
+
+library(ggalt)
+ggplot(time_duration, aes(x = time_start,
+                          xend = time_end,
+                          y = Date,
+                          group = Date)) +
+  geom_dumbbell(#color="#a3c4dc",
+    #size=0.75,
+    #point.colour.l="#0e668b"
+  ) #+
+scale_x_continuous(label=percent) +
+  labs(x=NULL,
+       y=NULL,
+       title="Dumbbell Chart",
+       subtitle="Pct Change: 2013 vs 2014",
[…]

### 2019-02-20 analysis/scripts/units_species.R

@@ -6,11 +6,13 @@
 # Last Edit: 2018 Oct 22
 ########
 
-tbl_Locations$GIS_Location_ID <- str_sub(tbl_Locations$GIS_Location_ID,start=-1)
+tbl_Locations$GIS_Location_ID <-
+  str_sub(tbl_Locations$GIS_Location_ID,start=-1)
+
 
 species_units <-
   left_join(species_data <- left_join(tbl_Field_Data, tbl_Events,
-                                                                by = "Event_ID") %>%
+                                      by = "Event_ID") %>%
               left_join(tlu_species_codes,
                         by = c("Species" = "Code")) %>%
               separate(Start_Date, c("Date", "Time_0"), sep = " ") %>%
@@ -28,12 +30,12 @@ species_units <-
   summarize(total_Observed = sum(`Num_ Observed`)) %>%
   left_join(tlu_species_codes, by = c("Species" = "Code"))# %>%
 #optional: spread command
-  #spread(GIS_Location_ID, total_Observed)
+#spread(GIS_Location_ID, total_Observed)
 
 #to do: make plot
 plot4 <- ggplot(species_units, aes(as.numeric(Days_since_first),
-                          total_Observed,
-                          color = as.factor(Year))) +
+                                   total_Observed,
+                                   color = as.factor(Year))) +
   geom_jitter(height = .3, aes(fill = GIS_Location_ID)) +
   #in ggforce package to supposedly distribute facets over pages
   facet_wrap_paginate(~Species, ncol = 1, scales = "free") +
@@ -41,9 +43,9 @@ plot4 <- ggplot(species_units, aes(as.numeric(Days_since_first),
 print(plot4)
 
 #how to
-  #1. order bars with a y aesthetic,
-  #2. have species not take up so much space
-  #3. get rid of 0/NA values to take up less space?? ("free_x" not working)
+#1. order bars with a y aesthetic,
+#2. have species not take up so much space
+#3. get rid of 0/NA values to take up less space?? ("free_x" not working)
 plot5 <- ggplot(species_units, aes(Species, total_Observed, fill = as.factor(Year))) +
   geom_bar(stat = "identity") +
   coord_flip() +
@@ -51,7 +53,7 @@ plot5 <- ggplot(species_units, aes(Species, total_Observed, fill = as.factor(Yea
   facet_wrap_paginate(~GIS_Location_ID, ncol = 2, scales = "free")
 print(plot5)
 #bar chart:
-  #faceted by unit (8) and by non-zero species?
[…]

### 2019-02-21 analysis/paper/paper.Rmd
Example of next code chunks to work on in paper.Rmd, and modified survey_timing to clarify what next steps are needed.

@@
 observer bias.
+
 Observers counted and identified all birds visible within each unit and
@@
 
-Questions: what is cue to start of survey season?
-*Madeleine later* Diagnostic analysis => check dates & unit of first non-zero counts each year for each spp (relative to dates of initiating surveys)
-
 The annual dates of the first and last survey each year has varied by over a month (Figure \@ref(fig:Fig-SurveyTiming)).
@@
 
+print(Plot1)
+```
+
+Questions: what is cue to start of survey season?
+*Madeleine later* Diagnostic analysis => check dates & unit of first non-zero counts each year for each spp (relative to dates of initiating surveys)
+
+Narrative on how well the current surveys are capturing the Spring arrival and Fall departure of each of the targeted species (\@ref(fig:Fig-SpeciesTiming)).
+Implications for informing Audubon Climate Change Watch or national phenology network, yadda yadd.
+
+```{r SpeciesTiming, eval=TRUE, cache = FALSE, include = FALSE}
+# Madeleine: write a script to extract, for just the target species listed at line 250, 
+# just the dates each year that each of the target species was observed (somewhere in the survey areas - don't care which, at the moment)
+# then add a layer to the ggplot 'Plot1' (horrid name, by the way) to visually distinguish those dates for that species.
+# will have to play around to see if it is feasible to do this by faceting on the species or just make a separate 
+# timing graph for each species - your call. It will be unreadable if try and put all of the target species on 
+# a single plot
+
+# tweak this call to source that new script, then un-comment, etc.
+# source("../scripts/species_timing_EDA.R", echo=FALSE)
+
+```
+
+```{r Fig-SpeciesTiming, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="For each of the target species, the dates on which they were observed in any of the sample areas."}
+
+#edit to use the name of the new plot
 print(Plot1)

### 2019-02-21 analysis/scripts/survey_timing_EDA.R

@@ -16,6 +16,23 @@ Plot1<- species_data %>%
              linetype=2, colour="grey") +
   labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)")
 
+
+#### STILL TO FINISH (then remove "if (0) {","}")
+if (0) {
+  # figure out which dates observed any of the key species - maybe facet on each?
+  # then plot an open circle over that survey date (so we see which surveys were bust)
+  # build toward species specific phenology plots.
+  species_data %>%
+    ggplot(aes(YearDay,Year)) +
+    geom_point(col="blue", size=2) +   # Draw points
+    geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
+               linetype=2, colour="grey") +
+    labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)")
+  # add new aes & geom_point to plot open circle on just those dates that species X was seen
+  # geom_point....
+}
+
+
 ####### Ideas to pursue later
 #######
 
@@ -31,14 +48,4 @@ if (0) {
 
 }
 
-if (0) {
-# figure out which dates observed any of the key species - maybe facet on each?
-# then plot an open circle over that survey date (so we see which surveys were bust)
-# build toward species specific phenology plots.
-  species_data %>%
-    ggplot(aes(YearDay,Year)) +
-    geom_point(col="blue", size=2) +   # Draw points
-    geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
-               linetype=2, colour="grey") +
-    labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)")
-}
+

### 2019-02-21 analysis/paper/paper.Rmd
Example of next code chunks to work on in paper.Rmd, and modified survey_timing to clarify what next steps are needed. (updated with some additions Feb 21)

@@
 *Target Frame* - space, time, scale, target populations
+
 *Spatial* Upper Lynn Canal visible from coastline from Skagway to Dyea, including Skagway & Taiya river mouths and Nahku Bay (between Skagway and Dyea Point)
+
+> INSERT FIGURE FROM NRSS report
 
@@
 Incidental: 48 other spp
+
+> Consider graphic of spp frequency of occurrence (maybe some panel-rich figure, one panel per spp, showing annual average percentage of daily observations in which spp was sighted - as line-connected points?- ordered from most frequently observed spp to least frequently.
+Think about appropriate estimator so annual value accounts for variation in survey effort across years.) likely needs to be a full page figure (e.g., we'll need to set  fig.width and fig.height)
+
+```{r EDA_AllSpp_Frequency, eval=TRUE, cache = FALSE, include = FALSE}
+# lets discuss what I'm thinking of. It is of secondary importance, so don't want this 
+# to consume tons of your time.
+
+#source("../scripts/survey_timing_EDA.R", echo=FALSE)
+
+```
+
+```{r Fig-AllSppFreq, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="blah blah about consistency in target species occurrence. Maybe something about new spp showing up?"}
+
+#print(Plot1)
+```
+
 Questions: are these all migratory? Any year-round residents? Any of special concern? 

### 2019-02-22 analysis/paper/paper.Rmd

@@
 
-> Consider graphic of spp frequency of occurrence (maybe some panel-rich figure, one panel per spp, showing annual average percentage of daily observations in which spp was sighted - as line-connected points?- ordered from most frequently observed spp to least frequently.
-Think about appropriate estimator so annual value accounts for variation in survey effort across years.) likely needs to be a full page figure (e.g., we'll need to set  fig.width and fig.height)
+> MADELEINE: add a graphic of spp frequency of occurrence (maybe some panel-rich figure, one panel per spp, showing annual average percentage of daily observations in which spp was sighted - as line-connected points?- ordered from most frequently observed spp to least frequently.
+(e.g., use facets).
+LET"S DISCUSS - Think about appropriate estimator so annual value accounts for variation in survey effort across years.) likely needs to be a full page figure (e.g., we'll need to set  fig.width and fig.height)
 
-```{r EDA_AllSpp_Frequency, eval=TRUE, cache = FALSE, include = FALSE}
+```{r EDA_AllSpp_Frequency, eval=TRUE, cache = FALSE, include = FALSE, fig.height=8, fig.width=6}
 # lets discuss what I'm thinking of. It is of secondary importance, so don't want this 
@@
 ```{r SpeciesTiming, eval=TRUE, cache = FALSE, include = FALSE}
-# Madeleine: write a script to extract, for just the target species listed at line 250, 
+# MADELEINE: write a script to extract, for just the target species listed at line 250, 
 # just the dates each year that each of the target species was observed (somewhere in the survey areas - don't care which, at the moment)

### 2019-02-23 analysis/scripts/target_species.R
This script creates a data frame of target species from the species_data frame from data_wrangling.R

@@ -1,0 +1,15 @@
+# Select data from target species from field dataset
+########
+# Libraries used: dplyr
+########
+# Madeleine Ward
+# Last Edit: 2019 Feb 23
+########
+
+#create vector of target species types
+targetList <- c("loon", "grebe", "cormorant", "eron", "duck", "goose", "plover", "sandpiper", "gull", "alcid", "kingfisher")
+
+#use data frame species_data from data_wrangling.R
+target_data <- species_data %>%
+  #filter for species in targetList
+  filter(str_detect(str_to_lower(Common_Name, locale = "en"), paste(targetList, collapse="|")))

### 2019-02-24 analysis/scripts/survey_timing_EDA.R
Made plot for target species sightings. In progress but runs w/o error.

@@ -11,26 +11,28 @@
 # Display the dates surveys were conducted across each year
 Plot1<- species_data %>%
   ggplot(aes(YearDay,Year)) +
-  geom_point(col="blue", size=2) +   # Draw points
+  geom_point(col="black", size=1.5) +   # Draw points
   geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
              linetype=2, colour="grey") +
-  labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)")
+  labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)") +
+  theme_bw()
 
+species_target_data <- species_data %>%
+  mutate(is_target = ifelse(Common_Name %in% target_data$Common_Name, 1, 0))
 
-#### STILL TO FINISH (then remove "if (0) {","}")
-if (0) {
+#### STILL TO FINISH
   # figure out which dates observed any of the key species - maybe facet on each?
   # then plot an open circle over that survey date (so we see which surveys were bust)
   # build toward species specific phenology plots.
-  species_data %>%
-    ggplot(aes(YearDay,Year)) +
-    geom_point(col="blue", size=2) +   # Draw points
-    geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
-               linetype=2, colour="grey") +
-    labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)")
-  # add new aes & geom_point to plot open circle on just those dates that species X was seen
-  # geom_point....
-}
+Plot2<- species_target_data %>%
+  ggplot(aes(YearDay,Year)) +
+  geom_point(col="black", size=1.5, aes(shape = as.factor(is_target))) +   # Draw points
+  geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
+             linetype=2, colour="grey") +
+  labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)")+
+  scale_shape_manual(values = c(16, 21)) +
+  theme_bw() +
+  facet_wrap(~Common_Name)
 
 
 ####### Ideas to pursue later

### 2019-02-27 analysis/paper/paper.Rmd
Paper updated with call for target data frame
Figure requests to ensure Madeleine doesn't get bored.

@@
 output:
-  bookdown::word_document2:
-    fig_caption: yes
-    toc: yes
   html_document:
@@
     number_sections: yes
+    toc: yes
+  bookdown::word_document2:
+    fig_caption: yes
     toc: yes
@@
 - What is up with 2008 surveys? why such distinct timing?
+- Think about Covariate EDA, if any...change in observers?
+- Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
+- clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.
 
@@
 ```
+*Site Accessibility*
+How consistently are they able to conduct surveys at each site? E.g., how often is Site 1 dropped due to cruise ships?
+
+> MADELEINE: With all these initial round of figures, don't get too caught up in refining fine details as we'll likely iterate a couple times and very well may end up deciding a totally different figure is more effective.
+For example, this initial round of requests is a lot of dotplots, more or less, but we may end up realizing it is better to look at a summary of '% success per week' or something else that is a more continuous outcome variable.
+Hopefully some of that makes sense.
+
+> MADELEINE: add two graphs (they will end up rather similar to ones for looking at timing of survey effort) to visualize how consistently they are able to complete planned surveys at each of the survey 'sites'.
+Probably okay to do both at dotplots or variants thereof.
+
+> FIRST graph: x axis is Day of Season (Julian Date; be nice to ultimately include vertical reference lines to denote beginnings of months), y axis is year; open dots (circles/points) for the days of planned surveys - e.g., every day there was any observation effort at any site; closed dots (filled in circles, pch=19?) for days where the survey was successfully completed, facet by site.
+So we are seeing whethere there are consistently problematic sites (like site 1, half of which consists of cruise ship docking areas).
+
+> SECOND graph: [THIS IS SECONDARY PRIORITY] x axis is Time of Day, perhaps broken in to 1 or 2 hour blocks , y axis is ....this is a tough one so totally open to your ideas.
+How about y axis is % [0, 1.0] and for each year there is a different curve showing the % of planned surveys (as defined above) that were successfully completed that season within that time block; facet by site.
+So we are seeing whethere there are consistently problematic times of day at each site.
+
+*Survey Effort*
+Would like to get a sense of the consistency of (alternatively, variation in) survey effort (e.g., duration at each site) across Days of season and years.Like many of these figure requests, we are basically looking to check that there isn't anything glaring going on underneath so we are comfortable later summarizing across this particular variable, or making sure we account for it via some sort of weighting of observations (perhaps by focusing on a variant of something known in fisheries as Catch Per Unit Effort).
+
+> MADELEINE: (that really isn't meant to come across as yelling, just trying to consistently label these so you can easily search for them) This is a very rough thought - feel free to suggest something else.
+We want to look at variation in 'duration' at each site across days of season and years.
+Maybe facet by season week, y axis is year, x axis is duration, horizontal boxplot of duration.
+This lumps all sites - maybe redo that graph again, once for each site?
 
@@
 (e.g., use facets).
-LET"S DISCUSS - Think about appropriate estimator so annual value accounts for variation in survey effort across years.) likely needs to be a full page figure (e.g., we'll need to set  fig.width and fig.height)
+x axis is Day of Season, y axis is % as decimal [0, 1.0], curves for each year; facet by species.
+LET"S DISCUSS - Think about appropriate estimator so annual value accounts for variation in survey effort across years.) likely needs to be a full page figure (e.g., we'll need to set  fig.width and fig.height).
+TAKE A STAB AT THIS BUT DON'T get lost trying to complete all the refinements.
+Really we should build up to this.
 
@@
 
+> MADELEINE: need to 'stage' the new script file when you do a 'commit' so that it gets incorporated into the commit.
+There isn't a copy of target_data.R in the repository yet...
+I set 'eval = FALSE' to skip the source call for now.
+```{r Target_Species, eval=FALSE, echo=FALSE, warning = TRUE}
+# selects only target species listed above
+#
+source("../scripts/target_data.R")
+```
+
 *Temporal* ??how define ‘season’??
@@
 Questions: what is cue to start of survey season?
-*Madeleine later* Diagnostic analysis => check dates & unit of first non-zero counts each year for each spp (relative to dates of initiating surveys)
+
+*Survey Timing and Changes in Phenology*
+Want to see if the survey design's start and end dates each season need to be modified to account for any observed changes in phenology (due to climate change, really).
+Ultimately need to review Audubon's Climate Change Watch List and see if any of their target species are observed in this effort.
+
+> MADELEINE: FIRST graph - x axis is year, y axis is Day of Season, curve for each of the target species defined in next sentence; possibly consider facet by site if anything of interest is revealed when doing so.
+Curve should be made of points of the first day the species was observed each year; maybe do one with just raw points connected by line segments and one with some sort of smooth - use default until we get a chance to look at these and talk about smoothers.
+Different color for each of the target species.
+Might be too many curves (species), in which case define some grouping on species or facet by species (more important than site).
+
+> SECOND graph - similar but showing last day of the season the species was observed each year.
+
+> THIRD graph - x axis is time of day, y axis is week of season, open circles for days with completed surveys, filled circles for surveys (times of day) when target species was observed in any year; facet on target spp.
+Or consider using the 'time of day' binning discussed in the Site Accessibility graphs (above).
+THIS IS SECONDARY PRIORITY - don't really like lumping all years together but no other ideas at the moment.
 

### 2019-02-27 analysis/paper/paper.Rmd
Figure requests to ensure Madeleine doesn't get bored.

@@
 output:
+  bookdown::word_document2:
+    fig_caption: yes
+    toc: yes
   html_document:
@@
     number_sections: yes
-    toc: yes
-  bookdown::word_document2:
-    fig_caption: yes
     toc: yes
@@
 
-> MADELEINE: need to 'stage' the new script file when you do a 'commit' so that it gets incorporated into the commit.
-There isn't a copy of target_data.R in the repository yet...
-I set 'eval = FALSE' to skip the source call for now.
-```{r Target_Species, eval=FALSE, echo=FALSE, warning = TRUE}
+```{r Target_Species, eval=TRUE, echo=FALSE, warning = TRUE}
 # selects only target species listed above

### 2019-02-27 analysis/paper/paper.Rmd
Fixed filename error in code chunk Target_Species.

--- 2019-02-27 analysis/paper/paper.Rmd
@@
 
-> MADELEINE: need to 'stage' the new script file when you do a 'commit' so that it gets incorporated into the commit.
-There isn't a copy of target_data.R in the repository yet...
-I set 'eval = FALSE' to skip the source call for now.
-```{r Target_Species, eval=FALSE, echo=FALSE, warning = TRUE}
+```{r Target_Species, eval=TRUE, echo=FALSE, warning = TRUE}
 # selects only target species listed above
 #
-source("../scripts/target_data.R")
+source("../scripts/target_species.R")
 ```

### 2019-03-15 analysis/scripts/survey_sites.R
Survey site setup; Trying to consistently define "failed" survey dates

@@ -1,0 +1,38 @@
+############
+#Visualize consistency of completion of planned surveys at each survey unit.
+########
+# Libraries used: 
+########
+# Madeleine Ward
+# Last Edit: 2019 Mar 15
+########
+
+unit_data <-
+  left_join(species_data <- left_join(tbl_Field_Data, tbl_Events,
+                                      by = "Event_ID") %>%
+              left_join(tlu_species_codes,
+                        by = c("Species" = "Code")) %>%
+              separate(Start_Date, c("Date", "Time_0"), sep = " ") %>%
+              separate(Start_Time, c("Date_01", "Time_Start"), sep = " ") %>%
+              #how to fit all these into one argument??
+              separate(End_Time, c("Date_02", "Time_End"), sep = " ") %>%
+              mutate(Date = mdy(Date)) %>%
+              filter(Year >= 2003) %>%
+              group_by(Year) %>%
+              arrange(Date) %>%
+              mutate(Days_since_first = Date - Date[row_number()==1]),
+            tbl_Locations,
+            by = "Location_ID") %>%
+  group_by(GIS_Location_ID, Year, Days_since_first, Species) %>%
+  summarize(total_Observed = sum(`Num_ Observed`)) %>%
+  left_join(tlu_species_codes, by = c("Species" = "Code"))
+
+plot_units_1 <- species_target_data %>%
+  ggplot(aes(YearDay,Year)) +
+  geom_point(col="black", size=1.5, aes(shape = as.factor(is_target))) +   # Draw points
+  geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
+             linetype=2, colour="grey") +
+  labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)")+
+  scale_shape_manual(values = c(16, 21)) +
+  theme_bw() +
+  facet_wrap(~Common_Name)

### 2019-03-15 analysis/scripts/data_wrangling.R
Finished prelim script for unit viability including the first of the two graphs described in paper.Rmd; for now, data frame is not filtered for only target species. Also, chose arbitrary method for determining failure at a site; needs updated.

@@ -14,7 +14,7 @@
 # 1st left_join - attaches relevant Event data to each event in tbl_Field_Data (using Event_ID)
 species_data <- left_join(tbl_Field_Data, tbl_Events,
                           by = "Event_ID") %>%
-  # 2nd lef_join - attaches relevant Species names (Common, Latin, etc.) (using Species Code)
+  # 2nd left_join - attaches relevant Species names (Common, Latin, etc.) (using Species Code)
   left_join(tlu_species_codes,
             by = c("Species" = "Code")) %>%
   #subset out just columns we care about, re-order by interest, rename where necessary

### 2019-03-15 analysis/scripts/survey_sites.R

@@ -1,38 +1,72 @@
 ############
 #Visualize consistency of completion of planned surveys at each survey unit.
 ########
-# Libraries used: 
+# Libraries used: dplyr, lubridate, ggplot2
 ########
 # Madeleine Ward
 # Last Edit: 2019 Mar 15
 ########
 
-unit_data <-
-  left_join(species_data <- left_join(tbl_Field_Data, tbl_Events,
-                                      by = "Event_ID") %>%
-              left_join(tlu_species_codes,
-                        by = c("Species" = "Code")) %>%
-              separate(Start_Date, c("Date", "Time_0"), sep = " ") %>%
-              separate(Start_Time, c("Date_01", "Time_Start"), sep = " ") %>%
-              #how to fit all these into one argument??
-              separate(End_Time, c("Date_02", "Time_End"), sep = " ") %>%
-              mutate(Date = mdy(Date)) %>%
-              filter(Year >= 2003) %>%
-              group_by(Year) %>%
-              arrange(Date) %>%
-              mutate(Days_since_first = Date - Date[row_number()==1]),
-            tbl_Locations,
+#create data frame with all survey dates at each unit, along with 
+#number of observations and duration at each
+unit_data_1 <-
+  #combine field data (observation), location data, and events (key)
+  left_join(tbl_Field_Data, tbl_Events,
+            by = "Event_ID") %>%
+  left_join(tbl_Locations,
             by = "Location_ID") %>%
-  group_by(GIS_Location_ID, Year, Days_since_first, Species) %>%
-  summarize(total_Observed = sum(`Num_ Observed`)) %>%
-  left_join(tlu_species_codes, by = c("Species" = "Code"))
+  #find sum of all observations for all species, incl. non-targets, at each date
+  group_by(GIS_Location_ID, Event_ID) %>%
+  summarize(Total_Obs = sum(`Num_ Observed`)) %>%
+  
+  #re-join with events data to get date
+  left_join(tbl_Events,
+            by = "Event_ID") %>%
+  #Julian dates (reused from data_wrangling.R)
+  mutate(Date = floor_date(mdy_hm(Start_Date), unit = "day"),
+         YearDay = yday(Date)) %>%
+  filter(Year > 2002) %>%
+  #select and order relevant variables
+  select(GIS_Location_ID, Date, YearDay, Day, Month, Year, Total_Obs, 
[…]

### 2019-03-15 analysis/paper/paper.Rmd
called survey site script within paper.Rmd. Runs without issue; I will need to add new captions.
Had to rearrange call to occur after calls for data_wrangling and target species data frame creation (so no object errors).

@@
 
-> MADELEINE: add two graphs (they will end up rather similar to ones for looking at timing of survey effort) to visualize how consistently they are able to complete planned surveys at each of the survey 'sites'.
-Probably okay to do both at dotplots or variants thereof.
-
-> FIRST graph: x axis is Day of Season (Julian Date; be nice to ultimately include vertical reference lines to denote beginnings of months), y axis is year; open dots (circles/points) for the days of planned surveys - e.g., every day there was any observation effort at any site; closed dots (filled in circles, pch=19?) for days where the survey was successfully completed, facet by site.
-So we are seeing whethere there are consistently problematic sites (like site 1, half of which consists of cruise ship docking areas).
-
 > SECOND graph: [THIS IS SECONDARY PRIORITY] x axis is Time of Day, perhaps broken in to 1 or 2 hour blocks , y axis is ....this is a tough one so totally open to your ideas.
@@
 source("../scripts/target_species.R")
+```
+
+> MADELEINE: add two graphs (they will end up rather similar to ones for looking at timing of survey effort) to visualize how consistently they are able to complete planned surveys at each of the survey 'sites'.
+Probably okay to do both at dotplots or variants thereof.
+
+> FIRST graph: x axis is Day of Season (Julian Date; be nice to ultimately include vertical reference lines to denote beginnings of months), y axis is year; open dots (circles/points) for the days of planned surveys - e.g., every day there was any observation effort at any site; closed dots (filled in circles, pch=19?) for days where the survey was successfully completed, facet by site.
+So we are seeing whethere there are consistently problematic sites (like site 1, half of which consists of cruise ship docking areas).
+
+```{r SurveySites, eval=TRUE, cache = FALSE, include = FALSE}
+# dotplot of survey occurrences by unit and whether there were observations
+source("../scripts/survey_sites.R", echo=FALSE)
+```
+
+```{r Fig-SurveySites, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
+
+print(plot_units_1)
 ```

### 2019-03-15 analysis/scripts/survey_sites.R

@@ -69,4 +69,3 @@ plot_units_1 <- unit_data_1 %>%
   scale_shape_manual(values = c(21, 16), breaks = c("Yes", "No")) +
   #facet by unit location
   facet_wrap(~GIS_Location_ID)
-plot_units_1

### 2019-03-18 analysis/scripts/survey_timing_EDA.R
Tried to use a join to get a set with all dates for all species, then an indicator of whether each was spotted on that day. Probably needs a separate set for each species?

@@ -5,11 +5,11 @@
 # Libraries used: ggplot2,
 ########
 # Joel Reynolds (repurposing code by Madeleine Ward)
-# Last Edit: 2019 Feb 8
+# Last Edit: 2019 Mar 17
 ########
 
 # Display the dates surveys were conducted across each year
-Plot1<- species_data %>%
+plot_surveydates <- species_data %>%
   ggplot(aes(YearDay,Year)) +
   geom_point(col="black", size=1.5) +   # Draw points
   geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
@@ -17,21 +17,36 @@ Plot1<- species_data %>%
   labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)") +
   theme_bw()
 
-species_target_data <- species_data %>%
-  mutate(is_target = ifelse(Common_Name %in% target_data$Common_Name, 1, 0))
+#UNFINISHED
+all_dates <- species_data %>%
+  select(Date) %>%
+  unique() %>%
+  mutate(all_date = 1)
+
+all_targets <- target_data %>%
+  mutate(sighting_date = ifelse(Date %in% all_dates$Date, 1, 0))
+
+target_dates <- full_join(all_dates, all_targets, by = "Date")
+
+full_join(all_dates, by = "Date")
 
 #### STILL TO FINISH
-  # figure out which dates observed any of the key species - maybe facet on each?
-  # then plot an open circle over that survey date (so we see which surveys were bust)
-  # build toward species specific phenology plots.
-Plot2<- species_target_data %>%
-  ggplot(aes(YearDay,Year)) +
-  geom_point(col="black", size=1.5, aes(shape = as.factor(is_target))) +   # Draw points
-  geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
-             linetype=2, colour="grey") +
+# figure out which dates observed any of the key species - maybe facet on each?
+# then plot an open circle over that survey date (so we see which surveys were bust)
+# build toward species specific phenology plots.
+
+
+Plot2<- ggplot(target_dates, aes(YearDay, Year)) +
+  geom_point(shape = 21)
+
+Plot3 <- Plot2 +
+  geom_point(aes(shape = as.factor(is_sighted))) +
+  facet_wrap(~Common_Name)
+
[…]

### 2019-03-18 analysis/paper/paper.Rmd
survey_dates.R looks at the first and last days of target species sightings during surveys to account for possible phenology changes. survey_duration.R attempts to partially examine survey effort by looking at boxplots of duration in each site by year.

@@
 
+```{r SurveyEffort, eval=TRUE, cache = FALSE, include = FALSE}
+#source("../scripts/survey_duration.R", echo=FALSE)
+```
+
+```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap=""}
+#print(plot_duration)
+```
+
 Species: 
@@
 # dotplot of survey times by year
-source("../scripts/survey_timing_EDA.R", echo=FALSE)
+#source("../scripts/survey_timing_EDA.R", echo=FALSE)
 
@@
 
-print(Plot1)
+#print(Plot1)
 ```
@@
 
+```{r SurveyDates, eval=TRUE, cache = FALSE, include = FALSE}
+# scatterplot of species' first and last sightings
+source("../scripts/survey_dates.R", echo=FALSE)
+```
+
+```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The first sighting of each species each year."}
+print(plot_timing_1)
+```
+
 > SECOND graph - similar but showing last day of the season the species was observed each year.
+
+```{r Fig-SurveyDates2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The last sighting of each species each year."}
+print(plot_timing_2)
+```
 

### 2019-03-18 analysis/scripts/survey_dates.R

@@ -1,0 +1,58 @@
+############
+# Graphs to see if the survey design's start and end dates each season 
+# need to be modified to account for any observed changes in phenology
+########
+# Libraries used: dplyr, lubridate, tidyr
+########
+# Madeleine Ward
+# Last Edit: 2019 Mar 17
+########
+
+#dataset with both start and end dates for each target species
+target_start_end <- target_data %>%
+  left_join(tbl_Locations, by = "Location_ID") %>%
+  # group by year
+  group_by(Common_Name, Year
+           #         , Location_ID
+  ) %>%
+  arrange(YearDay) %>%
+  # take first and last day for each group
+  slice(c(1, n())) %>%
+  # mark whether it is the first or last day each year
+  mutate(Start_End = ifelse(row_number()==1, "Start", "End")) %>%
+  ungroup() 
+
+# separate sets for both start and end dates
+target_start <- target_start_end %>%
+  filter(Start_End == "Start")
+
+target_end <- target_start_end %>%
+  filter(Start_End == "End")
+
+# this plot shows start dates connected with piecewise lines
+plot_timing_1 <- 
+  ggplot(target_start, aes(x= Year, y= YearDay)) +
+  #ggplot(target_start_end, aes(x= Year, y= YearDay)) +
+  geom_point(
+    # to put both start/end on the same plot:
+    #aes(color = as.factor(Start_End))
+  ) +
+  geom_line()  +
+  #geom_smooth() +
+  facet_wrap(~Common_Name
+             #, scales = "free_y"
+  ) +
+  labs(title = "Date of First Observation by Year", y = "Julian Date \n 
+       (vertical lines denote 1st of April, May, ... and Oct in non-leap years)") 
+
+# same graph but for end dates
+plot_timing_2 <- 
+  ggplot(target_end, aes(x= Year, y= YearDay)) +
+  geom_point() +
+  geom_line()  +
+  #geom_smooth() +
+  facet_wrap(~Common_Name
+             #, scales = "free_y"
+  ) +
+  labs(title = "Date of First Observation by Year", y = "Julian Date \n 
+       (vertical lines denote 1st of April, May, ... and Oct in non-leap years)") 

### 2019-03-18 analysis/scripts/survey_duration.R

@@ -1,0 +1,25 @@
+############
+# Looks at variation in survey duration at each site across days of season, year, and survey site.
+########
+# Libraries used: ggplot2, dplyr
+########
+# Madeleine Ward
+# Last Edit: 2019 Mar 17
+########
+
+# Use unit_data_2 from survey_sites.R
+duration_data <- unit_data_2 %>%
+  # filter for negative (error) durations, NA Site IDs
+  filter(Duration >= 0, !is.na(GIS_Location_ID))
+
+# group by year to get separate boxplot for each year
+plot_duration <- ggplot(duration_data, aes(Year, Duration, group = Year
+)) +
+  geom_boxplot() +
+  # will result in vertical plots; flip
+  coord_flip() +
+  # facet by site and free the x scale to accommodate site-specific outliers
+  facet_wrap(~GIS_Location_ID, scales = "free_x") +
+  labs(title = "Survey Duration by Site", y = "Duration (min)")
+
+# TO DO: add descriptions of the sites as subtitles

### 2019-03-18 analysis/scripts/survey_sites.R

@@ -27,7 +27,7 @@ unit_data_1 <-
          YearDay = yday(Date)) %>%
   filter(Year > 2002) %>%
   #select and order relevant variables
-  select(GIS_Location_ID, Date, YearDay, Day, Month, Year, Total_Obs, 
+  select(GIS_Location_ID, Year, Date, YearDay, Day, Month, Year, Total_Obs, 
          Duration, Start_Time, End_Time, Event_Notes, 
          Location_ID, Event_ID) %>%
   #chose arbitrary value for considering a success based on observations above 0

### 2019-03-25 analysis/paper/paper.Rmd
Joel's additions and some comments on survey_dates, survey_sites, and target_species scripts.

@@
 # Issues to Resolve with content (narrative, figures, analyses)
+- **Joel** needs to add commands to target_species.R to define spp & Common Name as **ordinal** variables following AOU sequencing (e.g., so all ducks show up together, all gulls show up together, etc.).
+- **Joel & Madeleine** will discuss survey_dates.R commands and how to modify them so that we get a sense of when surveys were & when target species was first (or last) seen each year.
+E.g., separate out any filtering process stemming from survey effort levels and timing.
+(thinking some thing like showing open circles for survey events PRIOR to first sighting...)
+- Need to think about way to winnow out target species that do not occurr w/ adequate frequency to report on.
+E.g., those only seen <3 or 5 years???
+Look at phenology graphs and determine threshold to apply.
+Consider some species N-ward and S-ward migration routes may differ.
 - What is up with 2008 surveys? why such distinct timing?
@@
 ```{r Target_Species, eval=TRUE, echo=FALSE, warning = TRUE}
-# selects only target species listed above
+# selects out only target species listed above
 #
@@
 
-> MADELEINE: add two graphs (they will end up rather similar to ones for looking at timing of survey effort) to visualize how consistently they are able to complete planned surveys at each of the survey 'sites'.
-Probably okay to do both at dotplots or variants thereof.
+*Survey Completion*
 
-> FIRST graph: x axis is Day of Season (Julian Date; be nice to ultimately include vertical reference lines to denote beginnings of months), y axis is year; open dots (circles/points) for the days of planned surveys - e.g., every day there was any observation effort at any site; closed dots (filled in circles, pch=19?) for days where the survey was successfully completed, facet by site.
-So we are seeing whethere there are consistently problematic sites (like site 1, half of which consists of cruise ship docking areas).
-
-```{r SurveySites, eval=TRUE, cache = FALSE, include = FALSE}
+```{r SiteSurveyCompletion, eval=TRUE, cache = FALSE, include = FALSE}
 # dotplot of survey occurrences by unit and whether there were observations
@@
 
-```{r Fig-SurveySites, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
+Add sentence or two about success in completing surveys.
+Doesn't appear to be any consistently problematic sites, not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).
+**MADELEINE: Dig in to Site 1 to make sure there aren't messages hidden in 'notes' since that is the location that consists of cruise ship docking areas for a good chunk of the space.**
+```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="Need to add caption.",warning=FALSE}
 
-print(plot_units_1)
+print(plot_unitSuccess)
 ```
@@
 
-```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The first sighting of each species each year."}
+```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The first sighting of each species each year.", warning=FALSE}
 print(plot_timing_1)

### 2019-03-25 analysis/scripts/survey_dates.R

@@ -1,6 +1,7 @@
 ############
 # Graphs to see if the survey design's start and end dates each season 
 # need to be modified to account for any observed changes in phenology
+# of the target species.
 ########
 # Libraries used: dplyr, lubridate, tidyr
 ########
@@ -30,6 +31,11 @@ target_end <- target_start_end %>%
   filter(Start_End == "End")
 
 # this plot shows start dates connected with piecewise lines
+# MADELEINE: let's discuss how this could be modified to include a sense
+# of WHEN, each year, the surveys actually started. E.g., so the viewer can
+# separate out any changes in species phenology from changes in survey timing and effort levels.
+# ALSO may need to think about either dropping some spp (e.g., Am Black Duck) or otherwise
+# constraining the range of the y axis so really focus on early (or late) period.
 plot_timing_1 <- 
   ggplot(target_start, aes(x= Year, y= YearDay)) +
   #ggplot(target_start_end, aes(x= Year, y= YearDay)) +
@@ -42,8 +48,11 @@ plot_timing_1 <-
   facet_wrap(~Common_Name
              #, scales = "free_y"
   ) +
+  # add horizontal reference lines to denote start of months...
   labs(title = "Date of First Observation by Year", y = "Julian Date \n 
-       (vertical lines denote 1st of April, May, ... and Oct in non-leap years)") 
+       (horizontal lines denote 1st of April, May, ... and Oct in non-leap years)") +
+  geom_hline(yintercept=c(yday(ymd(paste("2007-",c(4:6),"-1",sep="")))),
+             linetype=2, colour="grey")
 
 # same graph but for end dates
 plot_timing_2 <- 

### 2019-03-25 analysis/scripts/survey_sites.R

@@ -3,10 +3,20 @@
 ########
 # Libraries used: dplyr, lubridate, ggplot2
 ########
-# Madeleine Ward
-# Last Edit: 2019 Mar 15
+# Madeleine Ward, Joel Reynolds
+# Last Edit: 2019 Mar 24
 ########
 
+# Note to Madeleine, 2019 March 24
+# It is better to just select out, and reformat where necessary, the subsets you want from the 
+# data object created in data_wrangling.R. Then you wouldn't have to re-filter by year, mutate date, etc.
+###
+# E.g., a good programming goal is minimize the number of new objects and ensure always pulling from 
+# the single source object so that anytime there is new data we only have to recreate that main source object
+# and we know all the other scripts will be updated appropriately.
+# That will also help so that when the Database is fixed (which I think may actually have happened), we
+# only need to update that one script for data ingesting & the rest doesn't need changed.
+####################
 #create data frame with all survey dates at each unit, along with 
 #number of observations and duration at each
 unit_data_1 <-
@@ -55,17 +65,17 @@ unit_data_2 <-
          Duration, Start_Time, End_Time, Event_Notes, 
          Location_ID, Event_ID)
 
-#based on non-target species; to-do: use target data
-plot_units_1 <- unit_data_1 %>%
+#based on non-target species; 
+# TO DO: use target data
+plot_unitSuccess <- unit_data_1 %>%
   ggplot(aes(YearDay,Year)) +
   #filled vs. unfilled shape shows viable survey
   geom_point(col="black", size=1.5, 
              aes(shape = as.factor(is_success))) +
   geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
              linetype=2, colour="grey") +
-  labs(x="Julian Date \n (vertical lines 
-       denote 1st of April, May, ... and Oct in non-leap years)") +
+  labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)") +
   #manually define shape values
-  scale_shape_manual(values = c(21, 16), breaks = c("Yes", "No")) +
+  scale_shape_manual(values = c(21, 16), breaks = c("Yes", "No"), name="Survey Completed Survey?") +
   #facet by unit location
   facet_wrap(~GIS_Location_ID)

### 2019-03-25 analysis/scripts/target_species.R

@@ -7,9 +7,12 @@
 ########
 
 #create vector of target species types
-targetList <- c("loon", "grebe", "cormorant", "eron", "duck", "goose", "plover", "sandpiper", "gull", "alcid", "kingfisher")
+targetList <- c("loon", "grebe", "cormorant", "heron", "duck", "goose", "plover", "sandpiper", "gull", "alcid", "kingfisher")
 
 #use data frame species_data from data_wrangling.R
 target_data <- species_data %>%
   #filter for species in targetList
   filter(str_detect(str_to_lower(Common_Name, locale = "en"), paste(targetList, collapse="|")))
+
+# Need to define the species list as an ordered factor following the AOU sequence.
+# Joel will do this.

### 2019-03-27 analysis/paper/paper.Rmd
Moved wrangling and target species sets higher in paper.Rmd. Also changed some small data sets made in scripts so paper.Rmd knits without issue.

@@
 ```
+
+```{r Data_Wrangling, eval=TRUE, echo=FALSE, warning = TRUE}
+# reconstructs species_data tibble from tables read in by DataInput
+#
+source("../scripts/data_wrangling.R")
+
+```
+
+```{r Target_Species, eval=TRUE, echo=FALSE, warning = TRUE}
+# selects out only target species listed above
+#
+source("../scripts/target_species.R")
+```
+
 *Site Accessibility*
@@
 ```{r SurveyEffort, eval=TRUE, cache = FALSE, include = FALSE}
-#source("../scripts/survey_duration.R", echo=FALSE)
+source("../scripts/survey_duration.R", echo=FALSE)
 ```
@@
 ```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap=""}
-#print(plot_duration)
+print(plot_duration)
 ```
@@
 All data, original data forms, and field notes were archived at KLGO.
-
-```{r Data_Wrangling, eval=TRUE, echo=FALSE, warning = TRUE}
-# reconstructs species_data tibble from tables read in by DataInput
-#
-source("../scripts/data_wrangling.R")
-
-```
-
-```{r Target_Species, eval=TRUE, echo=FALSE, warning = TRUE}
-# selects out only target species listed above
-#
-source("../scripts/target_species.R")
-```
 

### 2019-03-27 analysis/scripts/survey_duration.R

@@ -7,6 +7,25 @@
 # Last Edit: 2019 Mar 17
 ########
 
+#will delete this when all datasets are put into wrangling
+unit_data_2 <-
+  #combine field data (observation), location data, and events (key)
+  left_join(target_data, tbl_Locations,
+            by = "Location_ID") %>%
+  #find sum of all observations for only target species at each date
+  group_by(GIS_Location_ID, Event_ID) %>%
+  summarize(Total_Obs = sum(Num_Obs)) %>%
+  #re-join with events data to get date
+  left_join(tbl_Events,
+            by = "Event_ID") %>%
+  #Julian dates (reused from data_wrangling.R)
+  mutate(Date = floor_date(mdy_hm(Start_Date), unit = "day"),
+         YearDay = yday(Date)) %>%
+  #select and order relevant variables
+  select(GIS_Location_ID, Year, Date, YearDay, Total_Obs, 
+         Duration, Start_Time, End_Time, Event_Notes, 
+         Location_ID, Event_ID)
+
 # Use unit_data_2 from survey_sites.R
 duration_data <- unit_data_2 %>%
   # filter for negative (error) durations, NA Site IDs

### 2019-03-27 analysis/scripts/survey_sites.R

@@ -61,7 +61,7 @@ unit_data_2 <-
   mutate(Date = floor_date(mdy_hm(Start_Date), unit = "day"),
          YearDay = yday(Date)) %>%
   #select and order relevant variables
-  select(GIS_Location_ID, Date, YearDay, Total_Obs, 
+  select(GIS_Location_ID, Year, Date, YearDay, Total_Obs, 
          Duration, Start_Time, End_Time, Event_Notes, 
          Location_ID, Event_ID)
 

### 2019-04-01 analysis/scripts/data_wrangling.R
In process of simplifying data frames based on the one created in data_wrangling. Have for the time being moved all data frame code to data_wrangling, which will be moved back to respective scripts after. Not done yet.

@@ -8,6 +8,11 @@
 ########
 
 
+### TO DO: have copied all data frames created in their respective scripts.
+# Want to re-frame species_data objetc as all-inclusive (join: tbl_Field_Data, tbl_Events, tlu_species_codes, tbl_Locations)
+# and make all subsequent data frames created from the species_data object.
+# Will eventually replace script data frames with modified versions.
+
 # re-build the necessary data frame from the various CSV files
 # (basically reconstructing the query as if the database was functioning).
 
@@ -38,3 +43,104 @@ species_data <- left_join(tbl_Field_Data, tbl_Events,
   mutate(total_obs = sum(Num_Obs)) %>%
   ungroup()
 
+###temp data sets
+###
+###target_species.R
+#create vector of target species types
+targetList <- c("loon", "grebe", "cormorant", "heron", "duck", "goose", "plover", "sandpiper", "gull", "alcid", "kingfisher")
+
+#use data frame species_data from data_wrangling.R
+target_data <- species_data %>%
+  #filter for species in targetList
+  filter(str_detect(str_to_lower(Common_Name, locale = "en"), paste(targetList, collapse="|")))
+
+###temp
+###survey_dates.R
+#dataset with both start and end dates for each target species
+target_start_end <- target_data %>%
+  left_join(tbl_Locations, by = "Location_ID") %>%
+  # group by year
+  group_by(Common_Name, Year
+           #         , Location_ID
+  ) %>%
+  arrange(YearDay) %>%
+  # take first and last day for each group
+  slice(c(1, n())) %>%
+  # mark whether it is the first or last day each year
+  mutate(Start_End = ifelse(row_number()==1, "Start", "End")) %>%
+  ungroup() 
+
+# separate sets for both start and end dates
+target_start <- target_start_end %>%
+  filter(Start_End == "Start")
+
+target_end <- target_start_end %>%
+  filter(Start_End == "End")
+
+### temp
+###survey_sites.R
+
+#create data frame with all survey dates at each unit, along with 
+#number of observations and duration at each
+unit_data_1 <-
+  #combine field data (observation), location data, and events (key)
+  left_join(tbl_Field_Data, tbl_Events,
+            by = "Event_ID") %>%
+  left_join(tbl_Locations,
+            by = "Location_ID") %>%
[…]

### 2019-04-02 analysis/scripts/data_wrangling.R
standardized every data frame to work from data_wrangling. hopefully there is less excess now. Also ensured it still runs on paper.Rmd

@@ -9,7 +9,7 @@
 
 
 ### TO DO: have copied all data frames created in their respective scripts.
-# Want to re-frame species_data objetc as all-inclusive (join: tbl_Field_Data, tbl_Events, tlu_species_codes, tbl_Locations)
+# Want to re-frame species_data object as all-inclusive (join: tbl_Field_Data, tbl_Events, tlu_species_codes, tbl_Locations)
 # and make all subsequent data frames created from the species_data object.
 # Will eventually replace script data frames with modified versions.
 
@@ -22,118 +22,29 @@ species_data <- left_join(tbl_Field_Data, tbl_Events,
   # 2nd left_join - attaches relevant Species names (Common, Latin, etc.) (using Species Code)
   left_join(tlu_species_codes,
             by = c("Species" = "Code")) %>%
-  #subset out just columns we care about, re-order by interest, rename where necessary
-  select(Start_Date, Year, Month, Day, Start_Time, End_Time, Duration,
-         Location_ID,
-         Species, ID, Category, Family, Common_Name, Latin_Name, Latin_Code,
-         Num_Obs = 'Num_ Observed',
-         Temp, Cloud_Cover, Precip_Code,
-         Wind_code, Wind_dir, Wave_ht, Tide, Tide_Level, Tide_State,
-         Event_Group_ID, Protocol_Name, Observer, Event_Notes,
-         Data_ID, Event_ID, Data_Location_ID) %>%
-  #subset out just nonzero observations after 2003 (using renamed 'Num_Obs')
-  filter(Num_Obs > 0, Year >= 2003) %>%
+  # 3rd left join - attaches location & site information (using Location_ID)
+  left_join(tbl_Locations,
+            by = "Location_ID") %>%
   #strip HMS from Start_Date, create new Julian date column ('YearDay') & Fractional Month_Day
   mutate(Date = floor_date(mdy_hm(Start_Date), unit = "day"),
          YearDay = yday(Date)) %>%
   #         Month_DayFrac = format(Date, "%m-%d"),
+  #subset out just columns we care about, re-order by interest, rename where necessary
+  select(Date, Start_Date, Year, Month, Day, YearDay, Start_Time, End_Time, Duration,
+         Species, ID, Category, Family, Common_Name, Latin_Name, Latin_Code,
+         Num_Obs = 'Num_ Observed',
+         GIS_Location_ID, X_Coord, Y_Coord, Coord_Units, Loc_Name, Loc_Notes,
[…]

### 2019-04-02 analysis/scripts/survey_dates.R

@@ -11,7 +11,6 @@
 
 #dataset with both start and end dates for each target species
 target_start_end <- target_data %>%
-  left_join(tbl_Locations, by = "Location_ID") %>%
   # group by year
   group_by(Common_Name, Year
            #         , Location_ID
@@ -23,21 +22,15 @@ target_start_end <- target_data %>%
   mutate(Start_End = ifelse(row_number()==1, "Start", "End")) %>%
   ungroup() 
 
-# separate sets for both start and end dates
-target_start <- target_start_end %>%
-  filter(Start_End == "Start")
-
-target_end <- target_start_end %>%
-  filter(Start_End == "End")
-
 # this plot shows start dates connected with piecewise lines
 # MADELEINE: let's discuss how this could be modified to include a sense
 # of WHEN, each year, the surveys actually started. E.g., so the viewer can
 # separate out any changes in species phenology from changes in survey timing and effort levels.
 # ALSO may need to think about either dropping some spp (e.g., Am Black Duck) or otherwise
 # constraining the range of the y axis so really focus on early (or late) period.
-plot_timing_1 <- 
-  ggplot(target_start, aes(x= Year, y= YearDay)) +
+plot_timing_1 <- target_start_end %>%
+  filter(Start_End == "Start") %>%
+  ggplot(aes(x= Year, y= YearDay)) +
   #ggplot(target_start_end, aes(x= Year, y= YearDay)) +
   geom_point(
     # to put both start/end on the same plot:
@@ -55,8 +48,9 @@ plot_timing_1 <-
              linetype=2, colour="grey")
 
 # same graph but for end dates
-plot_timing_2 <- 
-  ggplot(target_end, aes(x= Year, y= YearDay)) +
+plot_timing_2 <- target_start_end %>%
+  filter(Start_End == "End") %>%
+  ggplot(aes(x= Year, y= YearDay)) +
   geom_point() +
   geom_line()  +
   #geom_smooth() +

### 2019-04-02 analysis/scripts/survey_duration.R

@@ -7,33 +7,24 @@
 # Last Edit: 2019 Mar 17
 ########
 
-#will delete this when all datasets are put into wrangling
-unit_data_2 <-
-  #combine field data (observation), location data, and events (key)
-  left_join(target_data, tbl_Locations,
-            by = "Location_ID") %>%
-  #find sum of all observations for only target species at each date
-  group_by(GIS_Location_ID, Event_ID) %>%
-  summarize(Total_Obs = sum(Num_Obs)) %>%
-  #re-join with events data to get date
-  left_join(tbl_Events,
-            by = "Event_ID") %>%
-  #Julian dates (reused from data_wrangling.R)
-  mutate(Date = floor_date(mdy_hm(Start_Date), unit = "day"),
-         YearDay = yday(Date)) %>%
-  #select and order relevant variables
-  select(GIS_Location_ID, Year, Date, YearDay, Total_Obs, 
-         Duration, Start_Time, End_Time, Event_Notes, 
-         Location_ID, Event_ID)
 
 # Use unit_data_2 from survey_sites.R
-duration_data <- unit_data_2 %>%
+#duration_data <- unit_data_2 %>%
   # filter for negative (error) durations, NA Site IDs
-  filter(Duration >= 0, !is.na(GIS_Location_ID))
+ # filter(Duration >= 0, !is.na(GIS_Location_ID))
+
+survey_duration <- target_data %>%
+  #find sum of all observations for all target species, at each date
+  group_by(GIS_Location_ID, Event_ID) %>%
+  mutate (total_obs_site = sum(Num_Obs)) %>%
+  ungroup()
 
 # group by year to get separate boxplot for each year
-plot_duration <- ggplot(duration_data, aes(Year, Duration, group = Year
-)) +
+plot_duration <- 
+  ggplot(survey_duration %>% 
+           # filter for negative (error) durations, NA Site IDs
+           filter(Duration >= 0, !is.na(GIS_Location_ID)), 
+         aes(Year, Duration, group = Year)) +
   geom_boxplot() +
   # will result in vertical plots; flip
   coord_flip() +

### 2019-04-02 analysis/scripts/survey_sites.R

@@ -17,57 +17,45 @@
 # That will also help so that when the Database is fixed (which I think may actually have happened), we
 # only need to update that one script for data ingesting & the rest doesn't need changed.
 ####################
+
 #create data frame with all survey dates at each unit, along with 
 #number of observations and duration at each
-unit_data_1 <-
-  #combine field data (observation), location data, and events (key)
-  left_join(tbl_Field_Data, tbl_Events,
-            by = "Event_ID") %>%
-  left_join(tbl_Locations,
-            by = "Location_ID") %>%
+#unit_data_1 <-
+ # #combine field data (observation), location data, and events (key)
+#  left_join(tbl_Field_Data, tbl_Events,
+ #           by = "Event_ID") %>%
+#  left_join(tbl_Locations,
+ #           by = "Location_ID") %>%
   #find sum of all observations for all species, incl. non-targets, at each date
-  group_by(GIS_Location_ID, Event_ID) %>%
-  summarize(Total_Obs = sum(`Num_ Observed`)) %>%
+  #group_by(GIS_Location_ID, Event_ID) %>%
+  #summarize(Total_Obs = sum(`Num_ Observed`)) %>%
   
   #re-join with events data to get date
-  left_join(tbl_Events,
-            by = "Event_ID") %>%
+  #left_join(tbl_Events,
+   #         by = "Event_ID") %>%
   #Julian dates (reused from data_wrangling.R)
-  mutate(Date = floor_date(mdy_hm(Start_Date), unit = "day"),
-         YearDay = yday(Date)) %>%
-  filter(Year > 2002) %>%
+  #mutate(Date = floor_date(mdy_hm(Start_Date), unit = "day"),
+   #      YearDay = yday(Date)) %>%
+  #filter(Year > 2002) %>%
   #select and order relevant variables
-  select(GIS_Location_ID, Year, Date, YearDay, Day, Month, Year, Total_Obs, 
-         Duration, Start_Time, End_Time, Event_Notes, 
-         Location_ID, Event_ID) %>%
+  #select(GIS_Location_ID, Year, Date, YearDay, Day, Month, Year, Total_Obs, 
+   #      Duration, Start_Time, End_Time, Event_Notes, 
+  #       Location_ID, Event_ID) %>%
   #chose arbitrary value for considering a success based on observations above 0
-  mutate(is_success = ifelse(Total_Obs == 0, "No", "Yes")) %>%
-  filter(!is.na(GIS_Location_ID))
[…]

### 2019-04-12 analysis/scripts/survey_duration.R
Change reference grid on Survey_Duration, minor updates on text, add official AOU data object to raw data file for sequencing target species.

@@ -3,7 +3,7 @@
 ########
 # Libraries used: ggplot2, dplyr
 ########
-# Madeleine Ward
+# Madeleine Ward (2019 Mar 17); Joel Reynolds (2019 Apr 11)
 # Last Edit: 2019 Mar 17
 ########
 
@@ -20,16 +20,26 @@ survey_duration <- target_data %>%
   ungroup()
 
 # group by year to get separate boxplot for each year
-plot_duration <- 
-  ggplot(survey_duration %>% 
+plot_duration <-
+  ggplot(survey_duration %>%
            # filter for negative (error) durations, NA Site IDs
-           filter(Duration >= 0, !is.na(GIS_Location_ID)), 
+           filter(Duration >= 0, !is.na(GIS_Location_ID)),
          aes(Year, Duration, group = Year)) +
   geom_boxplot() +
   # will result in vertical plots; flip
   coord_flip() +
   # facet by site and free the x scale to accommodate site-specific outliers
-  facet_wrap(~GIS_Location_ID, scales = "free_x") +
-  labs(title = "Survey Duration by Site", y = "Duration (min)")
+  #facet_wrap(~GIS_Location_ID, scales = "free_x") +
+
+  # JR: actually, prefer keeping x scale fixed to support easier comparison
+  # across panels (sites).
+  facet_wrap(~GIS_Location_ID, scales = "fixed") +
+  # Resolve extreme observations by fixing scale limits at 180 min (3 hrs)
+  # with vertical reference grid at 30 min intervals (rather than default 50)
+  scale_y_continuous(limits=c(0,185),breaks = seq(0,180,30),labels = c("0","","60","","120","","180"))+
+  theme(panel.grid.minor = element_blank()) + #, panel.grid.major.y=element_blank()) +
+  labs(title = "Survey Duration by Site", y = "Duration (min)",
+       caption="(1 survey > 180 minutes at Site 6 not shown)")
+
 
 # TO DO: add descriptions of the sites as subtitles

### 2019-04-12 analysis/scripts/target_species.R

@@ -1,4 +1,4 @@
-# Select data from target species from field dataset
+# Subset out data on target species from field dataset
 ########
 # Libraries used: dplyr
 ########
@@ -16,3 +16,10 @@ target_data <- species_data %>%
 
 # Need to define the species list as an ordered factor following the AOU sequence.
 # Joel will do this.
+
+# 4/11/19 - JR - Question:
+# is subsetting on Common_Name getting everything we want?
+#
+# Target Species AOU order: (using AOS 2018 data)
+# FINISH EDITING - or put in Data_wrangling?
+

### 2019-04-12 analysis/paper/paper.Rmd
In process of simplifying data frames based on the one created in data_wrangling. Have for the time being moved all data frame code to data_wrangling, which will be moved back to respective scripts after. Not done yet.
Change reference grid on Survey_Duration, minor updates on text, begin developing sequencing of target species.

@@
 ```
-# Things To Discuss with Madeleine
-- add all libraries used in scripts to 'setup' code chunk (just above) rather than just loading in script so that it is easy to see what needs to be loaded before you can successfully knit.
+# NOTES To Madeleine
+- see edits in survey_duration.R to aid interpretation by the viewer (e.g., changes in x axis tick locations, etc.)
 - documentation in scripts
@@
 # Issues to Resolve with content (narrative, figures, analyses)
-- **Joel** needs to add commands to target_species.R to define spp & Common Name as **ordinal** variables following AOU sequencing (e.g., so all ducks show up together, all gulls show up together, etc.).
 - **Joel & Madeleine** will discuss survey_dates.R commands and how to modify them so that we get a sense of when surveys were & when target species was first (or last) seen each year.
@@
 Assessing a monitoring program focuses on four broad topics following the major Phases of a monitoring program (ibid):  
-1. _why_ - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?  
+1. _Why_ - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?  
 2.
-_how_ - [Data Collection Design] Does the monitoring data collection design effectively addresses the information objectives?
+_How_ - [Data Collection Design] Does the monitoring data collection design effectively addresses the information objectives?
 Given interim learning regarding system characteristics and improvements in measurement technologies, are the information goals (including accuracy and precision goals), actually feasible?
@@
 3.
-_how_ - [Data Analysis Design] Do the data analysis methods effectively address the information objectives?
+_How_ - [Data Analysis Design] Do the data analysis methods effectively address the information objectives?
 Do the data analysis methods properly account for impactful features of the data collection design?
@@
 4.
-_how_ - [Workflows & Reporting] Are workflows and related software systems in place to allow timely, and efficient, data management, analysis updating, reporting, and distribution of findings to target audiences of information users?
+_How_ - [Workflows & Reporting] Are workflows and related software systems in place to allow timely, and efficient, data management, analysis updating, reporting, and distribution of findings to target audiences of information users?
 
@@
 
+> **Joel** needs to add commands to target_species.R to define spp & Common Name as **ordinal** variables following AOU sequencing (e.g., so all ducks show up together, all gulls show up together, etc.).
+
 ```{r Target_Species, eval=TRUE, echo=FALSE, warning = TRUE}
@@
 
-> MADELEINE: With all these initial round of figures, don't get too caught up in refining fine details as we'll likely iterate a couple times and very well may end up deciding a totally different figure is more effective.
-For example, this initial round of requests is a lot of dotplots, more or less, but we may end up realizing it is better to look at a summary of '% success per week' or something else that is a more continuous outcome variable.
-Hopefully some of that makes sense.
-
 > SECOND graph: [THIS IS SECONDARY PRIORITY] x axis is Time of Day, perhaps broken in to 1 or 2 hour blocks , y axis is ....this is a tough one so totally open to your ideas.
@@
 So we are seeing whethere there are consistently problematic times of day at each site.
+```{r SiteAccess, eval=TRUE, cache = FALSE, include = FALSE}
+# STILL NEED TO CREATE
+# source("../scripts/survey_access.R", echo=FALSE)
+```
+
+```{r Fig-SiteAccess, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="",warning= FALSE}
+# print(plot_access)
+```
 
 *Survey Effort*
-Would like to get a sense of the consistency of (alternatively, variation in) survey effort (e.g., duration at each site) across Days of season and years.Like many of these figure requests, we are basically looking to check that there isn't anything glaring going on underneath so we are comfortable later summarizing across this particular variable, or making sure we account for it via some sort of weighting of observations (perhaps by focusing on a variant of something known in fisheries as Catch Per Unit Effort).
-
-> MADELEINE: (that really isn't meant to come across as yelling, just trying to consistently label these so you can easily search for them) This is a very rough thought - feel free to suggest something else.
-We want to look at variation in 'duration' at each site across days of season and years.
-Maybe facet by season week, y axis is year, x axis is duration, horizontal boxplot of duration.
-This lumps all sites - maybe redo that graph again, once for each site?
+The time spent at each site during a survey has varied widely across the years, especially at Sites 1, 7, and 8 (Figure \@ref(fig:Fig-SurveyEffort)).
+This variation in survey effort may be due to changes in numbers of birds occurring (and hence needing to be ID'd and recorded), variation among observers in identification skills, changes in visibility and area available for consistent viewing at each site, as well as potential changes in protocol across the years.
+This variation in survey effort should be considered in any analysis of changes across years in occurrence or frequency of a species, or richness.
+Depending on the analysis, this might best be done by adjusting the response variable from 'counts' to 'counts per unit effort', weighting observations by survey effort, or considering how survey duration might be accounted for as a covariate.
 
@@
 
-```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap=""}
+```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="",warning= FALSE}
 print(plot_duration)
@@
 *Survey Completion*
+> Ultimately, re-order these sections to flow better. Completion then Duration, etc., etc.
 
@@
 
-Add sentence or two about success in completing surveys.
-Doesn't appear to be any consistently problematic sites, not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).
+Add sentence or two about success in completing surveys. Doesn't appear to be any consistently problematic sites, not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).
+
 **MADELEINE: Dig in to Site 1 to make sure there aren't messages hidden in 'notes' since that is the location that consists of cruise ship docking areas for a good chunk of the space.**
+
 ```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="Need to add caption.",warning=FALSE}
@@
 #edit to use the name of the new plot
-print(Plot1)
+#print(Plot1)
 ```

### 2019-04-12 analysis/scripts/survey_sites.R

@@ -8,17 +8,17 @@
 ########
 
 # Note to Madeleine, 2019 March 24
-# It is better to just select out, and reformat where necessary, the subsets you want from the 
+# It is better to just select out, and reformat where necessary, the subsets you want from the
 # data object created in data_wrangling.R. Then you wouldn't have to re-filter by year, mutate date, etc.
 ###
-# E.g., a good programming goal is minimize the number of new objects and ensure always pulling from 
+# E.g., a good programming goal is minimize the number of new objects and ensure always pulling from
 # the single source object so that anytime there is new data we only have to recreate that main source object
 # and we know all the other scripts will be updated appropriately.
 # That will also help so that when the Database is fixed (which I think may actually have happened), we
 # only need to update that one script for data ingesting & the rest doesn't need changed.
 ####################
 
-#create data frame with all survey dates at each unit, along with 
+#create data frame with all survey dates at each unit, along with
 #number of observations and duration at each
 #unit_data_1 <-
  # #combine field data (observation), location data, and events (key)
@@ -29,7 +29,7 @@
   #find sum of all observations for all species, incl. non-targets, at each date
   #group_by(GIS_Location_ID, Event_ID) %>%
   #summarize(Total_Obs = sum(`Num_ Observed`)) %>%
-  
+
   #re-join with events data to get date
   #left_join(tbl_Events,
    #         by = "Event_ID") %>%
@@ -38,27 +38,28 @@
    #      YearDay = yday(Date)) %>%
   #filter(Year > 2002) %>%
   #select and order relevant variables
-  #select(GIS_Location_ID, Year, Date, YearDay, Day, Month, Year, Total_Obs, 
-   #      Duration, Start_Time, End_Time, Event_Notes, 
+  #select(GIS_Location_ID, Year, Date, YearDay, Day, Month, Year, Total_Obs,
+   #      Duration, Start_Time, End_Time, Event_Notes,
   #       Location_ID, Event_ID) %>%
   #chose arbitrary value for considering a success based on observations above 0
   #mutate(is_success = ifelse(Total_Obs == 0, "No", "Yes")) %>%
   #filter(!is.na(GIS_Location_ID))
 
-#create data frame with all survey dates at each unit, along with 
+#create data frame with all survey dates at each unit, along with
 #number of observations and duration at each
+# NOTE: based on all species;
+#TO DO: filter to just target species data
[…]

### 2019-04-19 analysis/scripts/data_wrangling.R
unfinished - added a temp script to figure out what's going on with site 1; added occurrence script we talked about (not in paper.Rmd yet, have to make sure it runs); renamed a var in data_wrangling

@@ -43,7 +43,7 @@ species_data <- left_join(tbl_Field_Data, tbl_Events,
   #define grouping by Species
   group_by(Species, Year, YearDay) %>%
   #add total observations column (by species by observation date)
-  mutate(total_obs_all = sum(Num_Obs)) %>%
+  mutate(obs_date = sum(Num_Obs)) %>%
   ungroup()
 
 

### 2019-04-19 analysis/scripts/occurrence_freq.R

@@ -1,0 +1,44 @@
+############
+# Create a graphic of species frequency of occurrence.
+########
+# Libraries used: dplyr, ggplot2,
+########
+# Madeleine Ward
+# Last Edit: 2019 Apr 16
+########
+
+#MADELEINE: add a graphic of spp frequency of occurrence 
+#(maybe some panel-rich figure, one panel per spp, 
+#showing annual average percentage of daily observations 
+#in which spp was sighted - as line-connected points?- 
+#ordered from most frequently observed spp to least frequently. 
+#(e.g., use facets). x axis is Day of Season, 
+#y axis is % as decimal [0, 1.0], curves for each year; 
+#facet by species. LET'S DISCUSS - 
+#Think about appropriate estimator so annual value accounts 
+#for variation in survey effort across years.) 
+#likely needs to be a full page figure (e.g., 
+#we’ll need to set fig.width and fig.height).
+
+occurrence_data <- target_data %>%
+  group_by(Year) #%>%
+  #mutate()
+
+#second graph shows observations each year relative to other species
+#percent_obs will be the percent of observations each date that 
+#the spp composed
+occurrence_species <- target_data %>%
+  group_by(Year, YearDay) %>%
+  #find all observations of all species on certain date
+  mutate(all_obs = sum(obs_date)) %>%
+  ungroup() %>%
+  #find percent of observations accreditable to each species
+  group_by(Species, Year, YearDay) %>%
+  mutate(percent_obs = obs_date/all_obs) #%>%
+  #filter(Species == "HERG")
+
+ggplot(occurrence_species, aes(Year, percent_obs)) +
+  geom_boxplot(aes(group = Year)) +
+  #for now, limit set at 50% - how to account for variability?
+  scale_y_continuous(limits = c(0, .50)) +
+  facet_wrap(~Species)

### 2019-04-19 analysis/paper/paper.Rmd
occurrence_freq still in progress but I have a better idea; small changes to data_wrangling to accommodate surveys w/o sightings; changed survey_dates to your recommendations & adapted into paper.Rmd

@@
 ```{r SurveyDates, eval=TRUE, cache = FALSE, include = FALSE}
-# scatterplot of species' first and last sightings
+# scatterplot of species' first and sightings compared to survey season start
 source("../scripts/survey_dates.R", echo=FALSE)
@@
 
-```{r Fig-SurveyDates2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The last sighting of each species each year."}
-print(plot_timing_2)
-```
+#to do
 

### 2019-04-19 analysis/scripts/data_wrangling.R

@@ -16,8 +16,8 @@
 # re-build the necessary data frame from the various CSV files
 # (basically reconstructing the query as if the database was functioning).
 
-# 1st left_join - attaches relevant Event data to each event in tbl_Field_Data (using Event_ID)
-species_data <- left_join(tbl_Field_Data, tbl_Events,
+# 1st join - attaches relevant Event data to each event in tbl_Field_Data (using Event_ID)
+species_data <- full_join(tbl_Field_Data, tbl_Events,
                           by = "Event_ID") %>%
   # 2nd left_join - attaches relevant Species names (Common, Latin, etc.) (using Species Code)
   left_join(tlu_species_codes,

### 2019-04-19 analysis/scripts/occurrence_freq.R

@@ -7,18 +7,6 @@
 # Last Edit: 2019 Apr 16
 ########
 
-#MADELEINE: add a graphic of spp frequency of occurrence 
-#(maybe some panel-rich figure, one panel per spp, 
-#showing annual average percentage of daily observations 
-#in which spp was sighted - as line-connected points?- 
-#ordered from most frequently observed spp to least frequently. 
-#(e.g., use facets). x axis is Day of Season, 
-#y axis is % as decimal [0, 1.0], curves for each year; 
-#facet by species. LET'S DISCUSS - 
-#Think about appropriate estimator so annual value accounts 
-#for variation in survey effort across years.) 
-#likely needs to be a full page figure (e.g., 
-#we’ll need to set fig.width and fig.height).
 
 occurrence_data <- target_data %>%
   group_by(Year) #%>%

### 2019-04-19 analysis/scripts/survey_dates.R

@@ -3,22 +3,21 @@
 # need to be modified to account for any observed changes in phenology
 # of the target species.
 ########
-# Libraries used: dplyr, lubridate, tidyr
+# Libraries used: dplyr, ggplot2
 ########
 # Madeleine Ward
 # Last Edit: 2019 Mar 17
 ########
 
+first_surveys <- species_data %>%
+  group_by(Year) %>%
+  arrange(YearDay) %>%
+  slice(1) 
 
 #dataset with both start and end dates for each target species
 target_start_end <- target_data %>%
-  #one way to address when the surveys started is to use a 
-  #survey variable that grabs the # of days since the first survey date
-  group_by(Year) %>%
-  arrange(YearDay) %>%
-  mutate(Days_since_first = YearDay - YearDay[row_number()==1]) %>%
-  ungroup() %>%
-# now this will find first/last days
+  mutate(is_first = ifelse(Event_ID %in% first_surveys$Event_ID, 1, 0)) %>%
+  # now this will find first/last days
   # group by year
   group_by(Common_Name, Year
            #         , Location_ID
@@ -30,43 +29,44 @@ target_start_end <- target_data %>%
   mutate(Start_End = ifelse(row_number()==1, "Start", "End")) %>%
   ungroup() 
 
-# this plot shows start dates connected with piecewise lines
-# MADELEINE: let's discuss how this could be modified to include a sense
-# of WHEN, each year, the surveys actually started. E.g., so the viewer can
-# separate out any changes in species phenology from changes in survey timing and effort levels.
-# ALSO may need to think about either dropping some spp (e.g., Am Black Duck) or otherwise
-# constraining the range of the y axis so really focus on early (or late) period.
-plot_timing_1 <- target_start_end %>%
-  filter(Start_End == "Start") %>%
-  ggplot(aes(x= Year)) +
-  #ggplot(target_start_end, aes(x= Year, y= YearDay)) +
-  geom_point(aes(y=YearDay)
-    # to put both start/end on the same plot:
-    #aes(color = as.factor(Start_End))
-  ) +
-  geom_line(aes(y=YearDay))  +
-  #using Days_since_first
-  geom_line(aes(y = Days_since_first), color = "blue") +
-  #geom_smooth() +
-  facet_wrap(~Common_Name
-             #, scales = "free_y"
-  ) +
-  # add horizontal reference lines to denote start of months...
[…]

### 2019-04-20 analysis/scripts/survey_dates.R
changes we made during call

@@ -12,11 +12,13 @@
 first_surveys <- species_data %>%
   group_by(Year) %>%
   arrange(YearDay) %>%
-  slice(1) 
+  slice(1) %>%
+  ungroup() %>%
+  select(Year, YearDay)
 
 #dataset with both start and end dates for each target species
 target_start_end <- target_data %>%
-  mutate(is_first = ifelse(Event_ID %in% first_surveys$Event_ID, 1, 0)) %>%
+  mutate(is_first = ifelse(Date %in% first_surveys$Date, 1, 0)) %>%
   # now this will find first/last days
   # group by year
   group_by(Common_Name, Year
@@ -34,28 +36,29 @@ target_start_end <- target_data %>%
 
 # used species HERG as placeholder until I can get facet to work
 plot_timing_1 <- ggplot() +
-  geom_line(data =target_start_end %>%
-              filter(is_first == 1
-                     ), 
+  geom_line(data =first_surveys 
+            #%>%
+  #            filter(is_first == 1
+  #                   )
+  , 
             aes(x=Year, y=YearDay, color="blue")) +
   geom_line(data =target_start_end %>%
               filter(Start_End == "Start"
-                     , Species == "HERG"
+                     #, Species == "HERG"
                      ), 
             aes(x=Year, y=YearDay, color="red")) +
 labs(title = "Date of First Observation by Year", y = "Julian Date \n 
        (vertical lines denote 1st of April, May, ... 
      and Oct in non-leap years)")  +
-scale_color_discrete(name = "", labels = c("First Day of Survey Season",
-                                           "First Day of Species Sighting")) + 
+scale_color_discrete(name = "", labels = c("First Day of Species Sighting",
+                                           "First Day of Survey Season")) + 
   geom_hline(yintercept=c(yday(ymd(paste("2007-",c(4:6),"-1",sep="")))),
              linetype=2, colour="grey") + 
-  theme(legend.position="bottom") 
-#+
-#  facet_wrap(~target_start_end$Common_Name
+  theme(legend.position="bottom") +
+  facet_wrap(~Common_Name
              #, scales = "free_y"
-#  )
-
+  )
+plot_timing_1
 
 #explore more succinct way with dplyr
 #df1 <- data.frame(dates = x,Variable = rnorm(mean = 0.75,nmonths))

### 2019-04-20 analysis/scripts/survey_dates.R
Finally got the facet to work! For now, first sightings are dots while first surveys are still lines.

@@ -36,31 +36,36 @@ target_start_end <- target_data %>%
 
 # used species HERG as placeholder until I can get facet to work
 plot_timing_1 <- ggplot() +
-  geom_line(data =first_surveys 
+  geom_point(data =target_start_end %>%
+               filter(Start_End == "Start"
+                      #, Species == "HERG"
+               ), 
+             aes(x=Year, y=YearDay), color="red") +
+  geom_line(data =first_surveys
+            #%>%
+            #            filter(is_first == 1
+            #                   )
+            , 
+            aes(x=Year, y=YearDay), color="black") +
+  #geom_line(data =first_surveys 
             #%>%
   #            filter(is_first == 1
   #                   )
-  , 
-            aes(x=Year, y=YearDay, color="blue")) +
-  geom_line(data =target_start_end %>%
-              filter(Start_End == "Start"
-                     #, Species == "HERG"
-                     ), 
-            aes(x=Year, y=YearDay, color="red")) +
+  #, 
+   #         aes(x=Year, y=YearDay, color="blue")) +
 labs(title = "Date of First Observation by Year", y = "Julian Date \n 
        (vertical lines denote 1st of April, May, ... 
      and Oct in non-leap years)")  +
-scale_color_discrete(name = "", labels = c("First Day of Species Sighting",
-                                           "First Day of Survey Season")) + 
+#scale_color_discrete(values = c("First Day of Survey Season" = "black")) + 
   geom_hline(yintercept=c(yday(ymd(paste("2007-",c(4:6),"-1",sep="")))),
              linetype=2, colour="grey") + 
   theme(legend.position="bottom") +
-  facet_wrap(~Common_Name
+  facet_wrap(~Common_Name)
              #, scales = "free_y"
-  )
 plot_timing_1
 
 #explore more succinct way with dplyr
+
 #df1 <- data.frame(dates = x,Variable = rnorm(mean = 0.75,nmonths))
 #df2 <- data.frame(dates = x,Variable = rnorm(mean = -0.75,nmonths))
 

### 2019-04-20 analysis/scripts/survey_dates.R
Forgot to remove printing command from script

@@ -62,7 +62,8 @@ labs(title = "Date of First Observation by Year", y = "Julian Date \n
   theme(legend.position="bottom") +
   facet_wrap(~Common_Name)
              #, scales = "free_y"
-plot_timing_1
+
+#plot_timing_1
 
 #explore more succinct way with dplyr
 

### 2019-04-20 analysis/scripts/occurrence_freq.R
some changes to explore possibilities for species_occurrence graphs

@@ -7,10 +7,20 @@
 # Last Edit: 2019 Apr 16
 ########
 
+all_surveys <- species_data %>%
+  select(Date, Year, Month, Day, YearDay, Duration) %>%
+  distinct(Date
+           #, .keep_all = TRUE
+           )
 
 occurrence_data <- target_data %>%
-  group_by(Year) #%>%
-  #mutate()
+  filter(Species == "MEGU")
+left_join(all_surveys, occurrence_data, by = "Date")
+#%>%
+#mutate()
+
+
+
 
 #second graph shows observations each year relative to other species
 #percent_obs will be the percent of observations each date that 
@@ -29,4 +39,15 @@ ggplot(occurrence_species, aes(Year, percent_obs)) +
   geom_boxplot(aes(group = Year)) +
   #for now, limit set at 50% - how to account for variability?
   scale_y_continuous(limits = c(0, .50)) +
-  facet_wrap(~Species)
+  coord_flip() +
+  facet_wrap(~Common_Name)
+
+
+occurrence_species2 <- target_data %>%
+  group_by(Year) %>%
+  mutate(all_obs = sum(obs_date)) %>%
+  ungroup() %>%
+  #find percent of observations accreditable to each species
+  group_by(Species, Year) %>%
+  mutate(percent_obs = obs_date/all_obs)
+  

### 2019-04-26 analysis/paper/paper.Rmd
ID'd next milestones for Madeleine to focus on (listed in intro bullets).

@@
 - documentation in scripts
+- **NEXT Priority Milestones**: note - all references to line numbers are prior to any other edits!
+- finish graphs of first & last observation per year for each target species (around line 386);**ultimately** add a fitted linear trend line to each of those (we need to account for frequency of surveying, so some subtle effort things here, but can resolve that later);
+- graphs of species occurrence (around line 407) for each of target species; 
+- covariate distribution graphs (lots of univariate & bivariate things; around line 534);
+- plot of total # spp sighted each season across years (around line 600)
 
@@
 Consider some species N-ward and S-ward migration routes may differ.
+- Joel will finish script uploading AOU data csv object & using to order Target Species
 - What is up with 2008 surveys? why such distinct timing?
@@
 #### Detection issues associated with survey conditions
+>MADELIENE: Next couple plots are priority.
+
+.distribution of weather conditions
+.distribution of survey time by weather conditions
+.quick diagnostics to ID covariates seldom recorded, not consistently recorded, consistently valued (e.g., no basis for contrast), etc.
 .species specific counts w/in season by weather conditions and wave conditions
@@
 .richness by major weather conditions (Northerlies, Southerlies, etc.) and for each spp
-.distribution of weather conditions
-.distribution of survey time by weather conditions
 .diagnostic to ID key survey times to ensure high quality observers?
 .consistency in ‘breeding periods’ across years (as captured in data)
-.quick diagnostics to ID covariates seldom recorded, not consistently recorded, consistently valued (e.g., no basis for contrast), etc.
 

### 2019-04-27 analysis/paper/paper.Rmd
Added 2 new plots to paper, both from survey_timing_EDA: one showing all survey dates, and then a layered scatterplot with all species-specific sightings across all dates. Deleted duplicate code & cleaned up scripts

@@
 output:
-  html_document:
-    df_print: paged
+  pdf_document:
     fig_caption: yes
-    number_sections: yes
     toc: yes
@@
     reference_docx: NRR_Author_Template_V4.2.docx
-  pdf_document:
+  html_document:
+    df_print: paged
     fig_caption: yes
+    number_sections: yes
     toc: yes
@@
 # dotplot of survey times by year
-#source("../scripts/survey_timing_EDA.R", echo=FALSE)
+source("../scripts/survey_timing_EDA.R", echo=FALSE)
 
@@
 
-```{r Fig-SurveyTiming, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
+```{r Fig-SurveyTiming1, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
 
-#print(Plot1)
+print(plot_alldates)
+```
+
+JOEL - this graph below looks VERY good zoomed in, but kinda crowded in HTML; still working on a better way to facet with specified rows
+```{r Fig-SurveyTiming2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap=""}
+
+print(plot_phenology)
 ```

### 2019-04-27 analysis/scripts/survey_sites.R

@@ -7,44 +7,6 @@
 # Last Edit: 2019 Mar 24
 ########
 
-# Note to Madeleine, 2019 March 24
-# It is better to just select out, and reformat where necessary, the subsets you want from the
-# data object created in data_wrangling.R. Then you wouldn't have to re-filter by year, mutate date, etc.
-###
-# E.g., a good programming goal is minimize the number of new objects and ensure always pulling from
-# the single source object so that anytime there is new data we only have to recreate that main source object
-# and we know all the other scripts will be updated appropriately.
-# That will also help so that when the Database is fixed (which I think may actually have happened), we
-# only need to update that one script for data ingesting & the rest doesn't need changed.
-####################
-
-#create data frame with all survey dates at each unit, along with
-#number of observations and duration at each
-#unit_data_1 <-
- # #combine field data (observation), location data, and events (key)
-#  left_join(tbl_Field_Data, tbl_Events,
- #           by = "Event_ID") %>%
-#  left_join(tbl_Locations,
- #           by = "Location_ID") %>%
-  #find sum of all observations for all species, incl. non-targets, at each date
-  #group_by(GIS_Location_ID, Event_ID) %>%
-  #summarize(Total_Obs = sum(`Num_ Observed`)) %>%
-
-  #re-join with events data to get date
-  #left_join(tbl_Events,
-   #         by = "Event_ID") %>%
-  #Julian dates (reused from data_wrangling.R)
-  #mutate(Date = floor_date(mdy_hm(Start_Date), unit = "day"),
-   #      YearDay = yday(Date)) %>%
-  #filter(Year > 2002) %>%
-  #select and order relevant variables
-  #select(GIS_Location_ID, Year, Date, YearDay, Day, Month, Year, Total_Obs,
-   #      Duration, Start_Time, End_Time, Event_Notes,
-  #       Location_ID, Event_ID) %>%
-  #chose arbitrary value for considering a success based on observations above 0
-  #mutate(is_success = ifelse(Total_Obs == 0, "No", "Yes")) %>%
-  #filter(!is.na(GIS_Location_ID))
-
 #create data frame with all survey dates at each unit, along with
 #number of observations and duration at each
 # NOTE: based on all species;

### 2019-04-27 analysis/scripts/survey_timing_EDA.R

@@ -5,48 +5,40 @@
 # Libraries used: ggplot2,
 ########
 # Joel Reynolds (repurposing code by Madeleine Ward)
-# Last Edit: 2019 Mar 17
+# Last Edit: 2019 Apr 27
 ########
 
-# Display the dates surveys were conducted across each year
-plot_surveydates <- species_data %>%
+## Display the dates surveys were conducted across each year
+
+plot_alldates <- species_data %>%
   ggplot(aes(YearDay,Year)) +
   geom_point(col="black", size=1.5) +   # Draw points
   geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
              linetype=2, colour="grey") +
-  labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)") +
-  theme_bw()
+  labs(x="Julian Date \n (vertical lines denote 1st of April, May,
+       ... and Oct in non-leap years)")
 
-#UNFINISHED
-all_dates <- species_data %>%
-  select(Date) %>%
-  unique() %>%
-  mutate(all_date = 1)
+## Display sightings for each species collapsed across sites for given survey date
 
-all_targets <- target_data %>%
-  mutate(sighting_date = ifelse(Date %in% all_dates$Date, 1, 0))
-
-target_dates <- full_join(all_dates, all_targets, by = "Date")
-
-full_join(all_dates, by = "Date")
-
-#### STILL TO FINISH
-# figure out which dates observed any of the key species - maybe facet on each?
-# then plot an open circle over that survey date (so we see which surveys were bust)
-# build toward species specific phenology plots.
-
-
-Plot2<- ggplot(target_dates, aes(YearDay, Year)) +
-  geom_point(shape = 21)
-
-Plot3 <- Plot2 +
-  geom_point(aes(shape = as.factor(is_sighted))) +
-  facet_wrap(~Common_Name)
-
-geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
-           linetype=2, colour="grey") +
-  labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)")+
-  scale_shape_manual(values = c(16, 21)) +
+plot_phenology <- ggplot() +
+  #add layer for all possible survey dates
+  geom_point(data = species_data %>%
+               distinct(Date, .keep_all = TRUE) %>%
+               select(Date, Year, YearDay),
[…]

### 2019-04-29 temp_covariates.R
forgot to commit these changes: creation of distribution script

@@ -1,0 +1,16 @@
+############
+# Examine various covariates and distributions.
+########
+# Libraries used: 
+########
+# Madeleine Ward
+# Last Edit: 2019 Apr 27
+########
+
+# Distribution of weather conditions
+#plot_winddist <- ggplot(tbl_Events, aes(as.factor(Year))) +
+ # geom_bar(fill = as.factor(Wind_code))
+
+
+#plot_winddist
+

### 2019-04-29 analysis/scripts/temp_covariates.R
some distributions

@@ -1,0 +1,29 @@
+############
+# Examine various covariates and distributions.
+########
+# Libraries used: ggplot2
+########
+# Madeleine Ward
+# Last Edit: 2019 Apr 27
+########
+
+### Distribution of weather conditions
+
+# Distribution of wind speed by year
+plot_wind <- ggplot(tbl_Events %>%
+                      left_join(tbl_Locations, by = "Location_ID") %>%
+                      filter(Year > 2003, !is.na(Wind_code), !is.na(GIS_Location_ID)),
+                    aes(as.factor(Year))) +
+  geom_bar(aes(fill = as.factor(Wind_code))) +
+  scale_fill_discrete(labels = tlu_wind_codes$Wind_Speed,
+                      name = "Wind Speed") +
+  labs(title = "Distribution of Wind Conditions Across Years",
+       x = "Year", y = "Count") +
+  theme(axis.text.x = element_text(angle=65, vjust = .6))
+plot_wind
+# Wind speed by year & site
+plot_windsite <- plot_wind +
+  facet_wrap(~GIS_Location_ID)
+plot_windsite
+
+

### 2019-04-29 temp_covariates.R

@@ -1,16 +1,0 @@
-############
-# Examine various covariates and distributions.
-########
-# Libraries used: 
-########
-# Madeleine Ward
-# Last Edit: 2019 Apr 27
-########
-
-# Distribution of weather conditions
-#plot_winddist <- ggplot(tbl_Events, aes(as.factor(Year))) +
- # geom_bar(fill = as.factor(Wind_code))
-
-
-#plot_winddist
-

### 2019-04-29 analysis/scripts/temp_covariates.R
more rough distribution graphs of weahter; started covariation plots

@@ -9,7 +9,7 @@
 
 ### Distribution of weather conditions
 
-# Distribution of wind speed by year
+## Distribution of wind speed by year
 plot_wind <- ggplot(tbl_Events %>%
                       left_join(tbl_Locations, by = "Location_ID") %>%
                       filter(Year > 2003, !is.na(Wind_code), !is.na(GIS_Location_ID)),
@@ -18,12 +18,43 @@ plot_wind <- ggplot(tbl_Events %>%
   scale_fill_discrete(labels = tlu_wind_codes$Wind_Speed,
                       name = "Wind Speed") +
   labs(title = "Distribution of Wind Conditions Across Years",
-       x = "Year", y = "Count") +
-  theme(axis.text.x = element_text(angle=65, vjust = .6))
+       x = "Year", y = "Count")
 plot_wind
-# Wind speed by year & site
+## Wind speed by year & site
 plot_windsite <- plot_wind +
   facet_wrap(~GIS_Location_ID)
 plot_windsite
 
+## Distribution of wave height by year
+plot_waves <- ggplot(tbl_Events %>%
+                      left_join(tbl_Locations, by = "Location_ID") %>%
+                      filter(Year > 2003, !is.na(Wave_ht), !is.na(GIS_Location_ID)),
+                    aes(as.factor(Year))) +
+  # still need to change order of fill var
+  geom_bar(aes(fill = as.factor(Wave_ht))) +
+  scale_fill_discrete(labels = tlu_wave_codes$Description,
+                      name = "Waves") +
+  labs(title = "Distribution of Wave Conditions Across Years",
+       x = "Year", y = "Count") +
+  coord_flip() + 
+  theme(legend.position="bottom")
+plot_waves
+## Wind speed by year & site
+plot_windsite <- plot_wind +
+  facet_wrap(~GIS_Location_ID, ncol =2)
+plot_windsite
 
+### Covariates
+
+## Distribution of survey time by weather conditions
+# Needs survey_timing_EDA.R plots created first
+plot_timewind <- species_data %>%
+  distinct(Date, .keep_all = TRUE) %>%
+  filter(!is.na(Wind_code)) %>%
+  ggplot(aes(YearDay,Year)) +
+  geom_point(aes(fill = as.factor(Wind_code))) +   # Draw points
+  geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
+             linetype=2, colour="grey") +
[…]

### 2019-04-30 analysis/paper/paper.Rmd
Attempts to address faceting problem again with command; paper html may or may not work.

@@
 
-JOEL - this graph below looks VERY good zoomed in, but kinda crowded in HTML; still working on a better way to facet with specified rows
+JOEL - I specified ncol = 3 in the facet command but my html doesn't update with it (I know the command itself works since I can see the zoom in the script)?
+Are you able to see it?
+and/or does it work in solving the space problem?
 ```{r Fig-SurveyTiming2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap=""}

### 2019-04-30 analysis/scripts/survey_timing_EDA.R

@@ -39,7 +39,7 @@ plot_phenology <- ggplot() +
   #temp; make background white for clearer contrast
   theme_bw() +
   #facet by target species; common name used
-  facet_wrap(~Common_Name)
+  facet_wrap(~Common_Name, ncol = 3)
 
 
 ####### Ideas to pursue later

### 2019-04-30 analysis/scripts/survey_dates.R
Added last sightings graph to paper. Continue distributions.

@@ -9,12 +9,17 @@
 # Last Edit: 2019 Mar 17
 ########
 
-first_surveys <- species_data %>%
+firstlast_surveys <- species_data %>%
   group_by(Year) %>%
-  arrange(YearDay) %>%
+  arrange(YearDay) 
+
+first_surveys <- firstlast_surveys %>%
   slice(1) %>%
-  ungroup() %>%
   select(Year, YearDay)
+last_surveys <- firstlast_surveys %>%
+  slice(n()) %>%
+  select(Year, YearDay)
+
 
 #dataset with both start and end dates for each target species
 target_start_end <- target_data %>%
@@ -63,19 +68,24 @@ labs(title = "Date of First Observation by Year", y = "Julian Date \n
   facet_wrap(~Common_Name)
              #, scales = "free_y"
 
-#plot_timing_1
-
-#explore more succinct way with dplyr
-
-#df1 <- data.frame(dates = x,Variable = rnorm(mean = 0.75,nmonths))
-#df2 <- data.frame(dates = x,Variable = rnorm(mean = -0.75,nmonths))
-
-#df3 <- df1 %>%  mutate(Type = 'Amocycillin') %>%
-#  bind_rows(df2 %>%
-#              mutate(Type = 'Penicillin'))
-
-
-#ggplot(df3,aes(y = Variable,x = dates,color = Type)) + 
-#  geom_line() +
-#  ggtitle("Merged datasets")
+plot_timing_2 <- ggplot() +
+  geom_point(data =target_start_end %>%
+               filter(Start_End == "End"
+                      #, Species == "HERG"
+               ), 
+             aes(x=Year, y=YearDay), color="blue") +
+  geom_line(data =last_surveys
+            #%>%
+            #            filter(is_first == 1
+            #                   )
+            , 
+            aes(x=Year, y=YearDay), color="black") +
+  labs(title = "Date of Last Observation by Year", y = "Julian Date \n 
+       (vertical lines denote 1st of April, May, ... 
+       and Oct in non-leap years)")  +
+  #scale_color_discrete(values = c("First Day of Survey Season" = "black")) + 
+  geom_hline(yintercept=c(yday(ymd(paste("2007-",c(4:6),"-1",sep="")))),
+             linetype=2, colour="grey") + 
+  theme(legend.position="bottom") +
+  facet_wrap(~Common_Name, ncol = 3)
 

### 2019-04-30 analysis/scripts/temp_covariates.R

@@ -47,14 +47,51 @@ plot_windsite
 ### Covariates
 
 ## Distribution of survey time by weather conditions
-# Needs survey_timing_EDA.R plots created first
-plot_timewind <- species_data %>%
-  distinct(Date, .keep_all = TRUE) %>%
-  filter(!is.na(Wind_code)) %>%
-  ggplot(aes(YearDay,Year)) +
-  geom_point(aes(fill = as.factor(Wind_code))) +   # Draw points
+plot_timeweather <- species_data %>%
+  filter(Year >= 2003) %>%
+  #distinct(Date, .keep_all = TRUE) %>%
+  #filter(!is.na(Wind_code)) %>%
+  ggplot(aes(YearDay,Year)) +   
   geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
              linetype=2, colour="grey") +
   labs(x="Julian Date \n (vertical lines denote 1st of April, May,
-       ... and Oct in non-leap years)")
+       ... and Oct in non-leap years)") +
+  facet_wrap(~GIS_Location_ID)
+
+# To do: change color scheme so no one goes blind looking at this
+
+# By wind code
+plot_timewind <- plot_timeweather +
+  geom_point(aes(color = as.factor(Wind_code)),size = 2) +
+  scale_color_discrete(labels = tlu_wind_codes$Wind_Speed,
+                       name = "Wind Speed") +
+  labs(title = "Survey Timing and Wind Conditions")
 plot_timewind
+
+# By wave height
+plot_timewave <- plot_timeweather +
+  geom_point(aes(color = as.factor(Wave_ht)),size = 2) +
+  scale_color_discrete(labels = tlu_wave_codes$Description,
+                       name = "Waves") +
+  theme(legend.position = "bottom") +
+  labs(title = "Survey Timing and Wave Height")
+plot_timewave
+
+## Total number species by weather conditions
+# By wind code
+plot_targetsweather <- target_data %>%
+  group_by(Date) %>%
+  mutate(total_obs = sum(obs_date)) 
+plot_targetswind <- plot_targetsweather %>%
+  ggplot(aes(as.factor(Wind_code), total_obs)) +
+  geom_boxplot() +
+  coord_flip() +
+  coord_cartesian(ylim = c(0, 10000)) 
+#+ facet_wrap(~Common_Name)
+plot_targetswind
+
[…]

### 2019-05-01 analysis/scripts/survey_dates.R
Finish up survey_dates

@@ -3,32 +3,22 @@
 # need to be modified to account for any observed changes in phenology
 # of the target species.
 ########
-# Libraries used: dplyr, ggplot2
+# Libraries used: dplyr, ggplot2, ggforce
 ########
 # Madeleine Ward
 # Last Edit: 2019 Mar 17
 ########
 
+# collect all survey dates and arrange by timing
 firstlast_surveys <- species_data %>%
+  select(Year, YearDay) %>%
   group_by(Year) %>%
   arrange(YearDay) 
 
-first_surveys <- firstlast_surveys %>%
-  slice(1) %>%
-  select(Year, YearDay)
-last_surveys <- firstlast_surveys %>%
-  slice(n()) %>%
-  select(Year, YearDay)
-
-
 #dataset with both start and end dates for each target species
 target_start_end <- target_data %>%
-  mutate(is_first = ifelse(Date %in% first_surveys$Date, 1, 0)) %>%
-  # now this will find first/last days
-  # group by year
-  group_by(Common_Name, Year
-           #         , Location_ID
-  ) %>%
+  # find first/last days, by year
+  group_by(Common_Name, Year) %>%
   arrange(YearDay) %>%
   # take first and last day for each group
   slice(c(1, n())) %>%
@@ -36,56 +26,45 @@ target_start_end <- target_data %>%
   mutate(Start_End = ifelse(row_number()==1, "Start", "End")) %>%
   ungroup() 
 
-# this plot shows start dates of each survey, as well as
+## this plot shows start dates of each survey, as well as
 # first sighting of each target species
-
-# used species HERG as placeholder until I can get facet to work
 plot_timing_1 <- ggplot() +
+  # select all species-specific start dates
   geom_point(data =target_start_end %>%
                filter(Start_End == "Start"
                       #, Species == "HERG"
                ), 
              aes(x=Year, y=YearDay), color="red") +
-  geom_line(data =first_surveys
-            #%>%
-            #            filter(is_first == 1
-            #                   )
-            , 
+  # select all survey start dates
+  geom_line(data =firstlast_surveys %>%
+              slice(1), 
             aes(x=Year, y=YearDay), color="black") +
-  #geom_line(data =first_surveys 
-            #%>%
-  #            filter(is_first == 1
-  #                   )
-  #, 
-   #         aes(x=Year, y=YearDay, color="blue")) +
+  # titles and Julian date labels
[…]

### 2019-05-01 analysis/paper/paper.Rmd
Added last sightings graph to paper. Continue distributions.
Change cache

@@
 
-```{r Fig-AllSppFreq, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="blah blah about consistency in target species occurrence. Maybe something about new spp showing up?"}
+```{r Fig-AllSppFreq, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="blah blah about consistency in target species occurrence. Maybe something about new spp showing up?"}
 
@@
 
-```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="Need to add caption.",warning=FALSE}
+```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="Need to add caption.",warning=FALSE}
 
@@
 
-```{r Fig-SurveyTiming1, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
+```{r Fig-SurveyTiming1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
 
@@
 and/or does it work in solving the space problem?
-```{r Fig-SurveyTiming2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap=""}
+```{r Fig-SurveyTiming2, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap=""}
 
@@
 
-```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="The first sighting of each species each year.", warning=FALSE}
+```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The first sighting of each species each year.", warning=FALSE}
 print(plot_timing_1)
@@
 
-#to do
+```{r Fig-SurveyDates2, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The last sighting of each species each year.", warning=FALSE}
+print(plot_timing_2)
+```
 

### 2019-05-04 analysis/paper/paper.Rmd
I have fixed the facet compression problem! It is not(?) super reproducible but changes chunk options to manually expand the figure. Also added quick covariates to paper.rmd - definitly need cleaned up

@@
 
-```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The first sighting of each species each year.", warning=FALSE}
+```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The first sighting of each species each year.",fig.height=20, fig.width=12, warning=FALSE}
 print(plot_timing_1)
@@
 
-```{r SpeciesTiming, eval=TRUE, cache = FALSE, include = FALSE}
-# MADELEINE: write a script to extract, for just the target species listed at line 250, 
-# just the dates each year that each of the target species was observed (somewhere in the survey areas - don't care which, at the moment)
-# then add a layer to the ggplot 'Plot1' (horrid name, by the way) to visually distinguish those dates for that species.
-# will have to play around to see if it is feasible to do this by faceting on the species or just make a separate 
-# timing graph for each species - your call. It will be unreadable if try and put all of the target species on 
-# a single plot
+```{r Fig-SpeciesTiming2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="For each of the target species, the dates on which they were observed in any of the sample areas."}
 
-# tweak this call to source that new script, then un-comment, etc.
-# source("../scripts/species_timing_EDA.R", echo=FALSE)
-
-```
-
-```{r Fig-SpeciesTiming, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="For each of the target species, the dates on which they were observed in any of the sample areas."}
-
-#edit to use the name of the new plot
-#print(Plot1)
+print(plot_phenology)
 ```
@@
 .consistency in ‘breeding periods’ across years (as captured in data)
+
+```{r Covariates, eval=TRUE, cache = FALSE, include = FALSE}
+source("../scripts/temp_covariates.R")
+```
 

### 2019-05-04 analysis/scripts/temp_covariates.R

@@ -1,12 +1,12 @@
 ############
 # Examine various covariates and distributions.
 ########
-# Libraries used: ggplot2
+# Libraries used: ggplot2, stringr
 ########
 # Madeleine Ward
 # Last Edit: 2019 Apr 27
 ########
-
+library(stringr)
 ### Distribution of weather conditions
 
 ## Distribution of wind speed by year
@@ -19,11 +19,13 @@ plot_wind <- ggplot(tbl_Events %>%
                       name = "Wind Speed") +
   labs(title = "Distribution of Wind Conditions Across Years",
        x = "Year", y = "Count")
-plot_wind
+print(plot_wind)
 ## Wind speed by year & site
 plot_windsite <- plot_wind +
   facet_wrap(~GIS_Location_ID)
-plot_windsite
+print(plot_windsite)
+
+### Distribution of wave conditions
 
 ## Distribution of wave height by year
 plot_waves <- ggplot(tbl_Events %>%
@@ -38,11 +40,36 @@ plot_waves <- ggplot(tbl_Events %>%
        x = "Year", y = "Count") +
   coord_flip() + 
   theme(legend.position="bottom")
-plot_waves
-## Wind speed by year & site
-plot_windsite <- plot_wind +
+print(plot_waves)
+## Wave height by year & site
+plot_wavesite <- plot_waves +
   facet_wrap(~GIS_Location_ID, ncol =2)
-plot_windsite
+print(plot_wavesite)
+
+## Distribution of tide states by year
+#many problems with this graph...
+plot_tides <- ggplot(test <- tbl_Events %>%
+                       left_join(tbl_Locations, by = "Location_ID") %>%
+                                   mutate(All_tides = substr(Tide_Level, start = 1, stop = 1), locale = "en") %>%
+                       filter(Year >= 2003
+                              #,
+                              #!is.na(All_tides)
+                              , 
+                              !is.na(GIS_Location_ID)
+                              ),
+                     aes(as.factor(Year))) +
+  geom_bar(aes(fill = as.factor(Tide_Level))) +
+  scale_fill_discrete(labels = tlu_tide_level$tide_level,
+                      name = "Tide Level") +
+  labs(title = "Distribution of Tide Levels Across Years",
+       x = "Year", y = "Count") +
+  coord_flip() + 
[…]

### 2019-05-04 analysis/scripts/occurrence_freq.R
finished occurrence plots; will later add to paper.rmd

@@ -7,22 +7,7 @@
 # Last Edit: 2019 Apr 16
 ########
 
-all_surveys <- species_data %>%
-  select(Date, Year, Month, Day, YearDay, Duration) %>%
-  distinct(Date
-           #, .keep_all = TRUE
-           )
-
-occurrence_data <- target_data %>%
-  filter(Species == "MEGU")
-left_join(all_surveys, occurrence_data, by = "Date")
-#%>%
-#mutate()
-
-
-
-
-#second graph shows observations each year relative to other species
+# graph shows observations each year relative to other TARGET species
 #percent_obs will be the percent of observations each date that 
 #the spp composed
 occurrence_species <- target_data %>%
@@ -31,18 +16,10 @@ occurrence_species <- target_data %>%
   mutate(all_obs = sum(obs_date)) %>%
   ungroup() %>%
   #find percent of observations accreditable to each species
-  group_by(Species, Year, YearDay) %>%
-  mutate(percent_obs = obs_date/all_obs) #%>%
-  #filter(Species == "HERG")
-
-ggplot(occurrence_species, aes(Year, percent_obs)) +
-  geom_boxplot(aes(group = Year)) +
-  #for now, limit set at 50% - how to account for variability?
-  scale_y_continuous(limits = c(0, .50)) +
-  coord_flip() +
-  facet_wrap(~Common_Name)
-
-
+  group_by(Year, YearDay, Species) %>%
+  mutate(percent_obs = obs_date/all_obs) %>%
+  ungroup() 
+#each year
 occurrence_species2 <- target_data %>%
   group_by(Year) %>%
   mutate(all_obs = sum(obs_date)) %>%
@@ -50,4 +27,12 @@ occurrence_species2 <- target_data %>%
   #find percent of observations accreditable to each species
   group_by(Species, Year) %>%
   mutate(percent_obs = obs_date/all_obs)
-  
+
+occurrence_species$Common_Name <- reorder(occurrence_species$Common_Name, -occurrence_species$percent_obs)
+
+plot_occurrence <- ggplot(occurrence_species, aes(YearDay, percent_obs)) +
+  geom_point(alpha = .1) +
+  geom_line(alpha = .02) +
+  theme_bw() +
+  facet_wrap(~Common_Name, nrow = 5)
+plot_occurrence

### 2019-05-07 analysis/paper/paper.Rmd
changed paper.Rmd chunk options / deleted old scripts.

@@
 
-```{r Fig-SiteAccess, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="",warning= FALSE}
+```{r Fig-SiteAccess, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis',fig.height=20, fig.width=12, fig.cap="",warning= FALSE}
 # print(plot_access)
@@
 
-```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="",warning= FALSE}
+```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.height=15, fig.width=12,fig.cap="",warning= FALSE}
 print(plot_duration)
@@
 
-```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="Need to add caption.",warning=FALSE}
+```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=10, fig.width=18, fig.cap="Need to add caption.",warning=FALSE}
 
@@
 
-```{r Fig-SurveyTiming1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
+```{r Fig-SurveyTiming1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=10, fig.width=12, fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
 
 print(plot_alldates)
-```
-
-JOEL - I specified ncol = 3 in the facet command but my html doesn't update with it (I know the command itself works since I can see the zoom in the script)?
-Are you able to see it?
-and/or does it work in solving the space problem?
-```{r Fig-SurveyTiming2, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap=""}
-
-print(plot_phenology)
 ```
@@
 
-```{r Fig-SurveyDates2, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The last sighting of each species each year.", warning=FALSE}
+```{r Fig-SurveyDates2, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.height=20, fig.width=12, fig.cap="The last sighting of each species each year.", warning=FALSE}
 print(plot_timing_2)
@@
 
-```{r Fig-SpeciesTiming2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.cap="For each of the target species, the dates on which they were observed in any of the sample areas."}
+```{r Fig-SpeciesTiming2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.height=25, fig.width=12, fig.cap="For each of the target species, the dates on which they were observed in any of the sample areas."}
 

### 2019-05-07 analysis/scripts/Example.R

@@ -1,1 +1,0 @@
-# blah generic script file

### 2019-05-07 analysis/scripts/temp_site1.R

@@ -1,16 +1,0 @@
-############
-# Check Site 1's viability as survey site
-########
-# Libraries used:
-########
-# Madeleine Ward
-# Last Edit: 2019 Apr 17
-########
-
-event_notes <- species_data %>%
-  group_by(Year, YearDay, GIS_Location_ID) %>%
-  mutate(all_observed = sum(total_obs_all)) %>%
-  ungroup() %>%
-  filter(GIS_Location_ID == "Site 1") %>%
-  select(Event_Notes, all_observed, Year, YearDay, GIS_Location_ID) %>%
-  group_by(Event_Notes)

### 2019-05-07 analysis/scripts/data_wrangling.R
site accessibility attempt

@@ -45,13 +45,3 @@ species_data <- full_join(tbl_Field_Data, tbl_Events,
   #add total observations column (by species by observation date)
   mutate(obs_date = sum(Num_Obs)) %>%
   ungroup()
-
-
-###
-###survey_timing_EDA.R
-#unfinished
-
-###
-###UNUSED
-##time_duration.R
-##units_species.R

### 2019-05-07 analysis/scripts/survey_access.R

@@ -1,0 +1,81 @@
+############
+# Measure site accessibility
+########
+# Libraries used: 
+########
+# Madeleine Ward
+# Last Edit: 2019 May 7
+########
+
+#SECOND graph: [THIS IS SECONDARY PRIORITY] x axis is Time of Day, 
+#perhaps broken in to 1 or 2 hour blocks , y axis is ....
+#this is a tough one so totally open to your ideas. How about y axis is 
+#% [0, 1.0] and for each year there is a different curve showing the % of
+#planned surveys (as defined above) that were successfully completed that 
+#season within that time block; facet by site. So we are seeing whether there 
+#are consistently problematic times of day at each site. 
+
+temp_access <- target_data %>%
+  mutate(time_block = substr(Start_Time, 
+                             nchar(Start_Time)-4, nchar(Start_Time)-3)) %>%
+  group_by(GIS_Location_ID, Year, time_block) %>%
+  mutate(site_time_obs = sum(Num_Obs)) %>%
+  ungroup()
+
+# 
+plot_access <- ggplot(temp_access %>%
+                        filter(!is.na(GIS_Location_ID), !is.na(time_block))) +
+  geom_boxplot(aes(time_block, site_time_obs)) +
+  coord_flip(ylim = c(0, 1200)) +
+  labs() +
+  facet_wrap(~GIS_Location_ID, ncol = 3 
+             #, scales = "free_x"
+             )
+plot_access
+
+#plot_yearaccess <- ggplot(temp_access %>%
+#                            filter(!is.na(GIS_Location_ID)) %>%
+#                            group_by(Year, GIS_Location_ID) %>%
+#                            mutate(site_year_obs = sum(Num_Obs))) +
+#  geom_boxplot(aes(Year, site_year_obs)) +
+  #coord_flip(ylim = c(0, 1200)) +
+  # coord_cartesian(ylim = c(0, 1000)) +
+#  facet_wrap(~GIS_Location_ID, ncol = 3 
+             #, scales = "free_x"
+#  )
+#plot_yearaccess
+
+temp_access2 <- target_data %>%
+  mutate(time_block = substr(Start_Time, 
+                             nchar(Start_Time)-4, nchar(Start_Time)-3)) %>%
+  group_by(GIS_Location_ID, Year, time_block) %>%
[…]

### 2019-05-07 analysis/scripts/survey_dates.R

@@ -68,3 +68,22 @@ plot_timing_2 <- ggplot() +
   geom_hline(yintercept=c(yday(ymd(paste("2007-",c(4:6),"-1",sep="")))),
              linetype=2, colour="grey") + 
   facet_wrap_paginate(~Common_Name, ncol = 3)
+
+## layering both first/last sightings
+#plot_timing_3 <- ggplot() +
+ # geom_point(data =target_start_end, 
+#             aes(x=Year, y=YearDay, color = Start_End)) +
+#  geom_line(data =firstlast_surveys %>%
+#              slice(1), 
+#            aes(x=Year, y=YearDay), color="black") +
+#  geom_line(data =firstlast_surveys %>%
+#              slice(n()), 
+#            aes(x=Year, y=YearDay), color="black") +
+  # titles and Julian date labels
+#  labs(title = "Date of First & Last Observations by Year", y = "Julian Date \n 
+#       (vertical lines denote 1st of April, May, ... 
+#       and Oct in non-leap years)")  +
+  #scale_color_discrete(values = c("First Day of Survey Season" = "black")) + 
+#  geom_hline(yintercept=c(yday(ymd(paste("2007-",c(4:6),"-1",sep="")))),
+#             linetype=2, colour="grey") + 
+#  facet_wrap_paginate(~Common_Name, ncol = 3)

### 2019-05-08 analysis/paper/paper.Rmd
site accessibility attempt
Lots of changes:  clarity within scripts, new scripts in paper (survey access, species occurrence)

@@
   warning = TRUE, # set to FALSE when done
-  message = TRUE, # set to FALSE when doen
+  message = TRUE, # set to FALSE when done
   echo = FALSE,
@@
 # NOTES To Madeleine
-- see edits in survey_duration.R to aid interpretation by the viewer (e.g., changes in x axis tick locations, etc.)
-- documentation in scripts
+- see edits in survey_duration.R to aid interpretation by the viewer (e.g., changes in x axis tick locations, etc.) $\checkmark$
+- documentation in scripts $\checkmark$
 - **NEXT Priority Milestones**: note - all references to line numbers are prior to any other edits!
-- finish graphs of first & last observation per year for each target species (around line 386);**ultimately** add a fitted linear trend line to each of those (we need to account for frequency of surveying, so some subtle effort things here, but can resolve that later);
-- graphs of species occurrence (around line 407) for each of target species; 
+- finish graphs of first & last observation per year for each target species (around line 386);**ultimately** add a fitted linear trend line to each of those (we need to account for frequency of surveying, so some subtle effort things here, but can resolve that later); $\checkmark$
+- graphs of species occurrence (around line 407) for each of target species; $\checkmark$
 - covariate distribution graphs (lots of univariate & bivariate things; around line 534);
@@
 #NOTE: working directory when knitting markup is /analysis/paper. So relative pathway needs to be ../scripts etc.
+
 source("../scripts/data_retrievalCSV.R")
@@
 How about y axis is % [0, 1.0] and for each year there is a different curve showing the % of planned surveys (as defined above) that were successfully completed that season within that time block; facet by site.
-So we are seeing whethere there are consistently problematic times of day at each site.
-```{r SiteAccess, eval=TRUE, cache = FALSE, include = FALSE}
-# STILL NEED TO CREATE
-# source("../scripts/survey_access.R", echo=FALSE)
+So we are seeing whether there are consistently problematic times of day at each site.
+> JOEL: some updates on this!
+for now, times are in hour blocks.
+There can't be lines/curves for year since they are unreadable (up/down).
+I also tried to use color as a factor but then the graphs are crowded (so many years).
+So I have 2 alternatives, I don't really like either but some are more readable than others.
+See below.
+
+```{r SiteAccess, eval=TRUE, cache = FALSE}
+
+source("../scripts/survey_access.R", echo=FALSE)
 ```
 
-```{r Fig-SiteAccess, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis',fig.height=20, fig.width=12, fig.cap="",warning= FALSE}
-# print(plot_access)
+```{r Fig-SiteAccess1, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis',fig.height=20, fig.width=12, fig.cap="",warning= FALSE}
+
+print(plot_success)
+```
+
+```{r Fig-SiteAccess2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis',fig.height=15, fig.width=12, fig.cap="",warning= FALSE}
+
+print(plot_success3)
 ```
@@
 
-```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.height=15, fig.width=12,fig.cap="",warning= FALSE}
+```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.height=12, fig.width=12, fig.cap="",warning= FALSE}
+
 print(plot_duration)
@@
 x axis is Day of Season, y axis is % as decimal [0, 1.0], curves for each year; facet by species.
-LET"S DISCUSS - Think about appropriate estimator so annual value accounts for variation in survey effort across years.) likely needs to be a full page figure (e.g., we'll need to set  fig.width and fig.height).
-TAKE A STAB AT THIS BUT DON'T get lost trying to complete all the refinements.
-Really we should build up to this.
+> JOEL: I did some graphs for species occurrence that fit the description but not sure if they were what you were looking for?
+They do show species occurrence relative to others; as annual/daily percentage of daily observations in which spp was sighted (percentage calculated by annual/daily obs per species over all total obs - is this what you meant?
+let me know).
+Lines didn't work though (lots of ups and downs)
 
 ```{r EDA_AllSpp_Frequency, eval=TRUE, cache = FALSE, include = FALSE, fig.height=8, fig.width=6}
-# lets discuss what I'm thinking of. It is of secondary importance, so don't want this 
-# to consume tons of your time.
 
-#source("../scripts/survey_timing_EDA.R", echo=FALSE)
+source("../scripts/occurrence_freq.R", echo=FALSE)
 
@@
 
-```{r Fig-AllSppFreq, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="blah blah about consistency in target species occurrence. Maybe something about new spp showing up?"}
+```{r Fig-AllSppFreq1, echo=FALSE, eval=TRUE, warning=FALSE, cache=FALSE, fig.show= 'asis', fig.height=20, fig.width=12, fig.cap="blah blah about consistency in target species occurrence.
+Maybe something about new spp showing up?"}
 
-#print(Plot1)
+print(plot_occurrenceDay)
+```
+
+```{r Fig-AllSppFreq2, echo=FALSE, eval=TRUE, warning=FALSE, cache=FALSE, fig.show= 'asis', fig.height=25, fig.width=10, fig.cap="blah blah about consistency in target species occurrence.
+Maybe something about new spp showing up?"}
+
+print(plot_occurrenceYear)
 ```
@@
 # dotplot of survey occurrences by unit and whether there were observations
+
 source("../scripts/survey_sites.R", echo=FALSE)
@@
 
-```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=10, fig.width=18, fig.cap="Need to add caption.",warning=FALSE}
+```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=15, fig.width=18, fig.cap="Need to add caption.",warning=FALSE}
 
@@
 # dotplot of survey times by year
+
 source("../scripts/survey_timing_EDA.R", echo=FALSE)
-
 ```
 
-```{r Fig-SurveyTiming1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=10, fig.width=12, fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
+```{r Fig-SurveyTiming1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=8, fig.width=12, fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
 
@@
 ```{r SurveyDates, eval=TRUE, cache = FALSE, include = FALSE}
-# scatterplot of species' first and sightings compared to survey season start
+
+# scatterplot of species' first and last sightings compared to survey season start
 source("../scripts/survey_dates.R", echo=FALSE)
@@
 ```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The first sighting of each species each year.",fig.height=20, fig.width=12, warning=FALSE}
+
 print(plot_timing_1)
@@
 ```{r Fig-SurveyDates2, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.height=20, fig.width=12, fig.cap="The last sighting of each species each year.", warning=FALSE}
+
 print(plot_timing_2)
@@
 
+MADELEINE
+
 Percent of expected waterbird and breeding landbird species (174) confirmed (158; 91%)

### 2019-05-08 analysis/scripts/occurrence_freq.R

@@ -1,10 +1,10 @@
 ############
-# Create a graphic of species frequency of occurrence.
+# Create graphics of species frequency of occurrence by year and date.
 ########
-# Libraries used: dplyr, ggplot2,
+# Libraries used: dplyr, ggplot2
 ########
 # Madeleine Ward
-# Last Edit: 2019 Apr 16
+# Last Edit: 2019 May 8
 ########
 
 # graph shows observations each year relative to other TARGET species
@@ -12,27 +12,58 @@
 #the spp composed
 occurrence_species <- target_data %>%
   group_by(Year, YearDay) %>%
+  
   #find all observations of all species on certain date
   mutate(all_obs = sum(obs_date)) %>%
   ungroup() %>%
+  
   #find percent of observations accreditable to each species
   group_by(Year, YearDay, Species) %>%
   mutate(percent_obs = obs_date/all_obs) %>%
   ungroup() 
-#each year
+
+#attempt to reorder facets by total percent of observations
+#occurrence_species$Common_Name <- reorder(occurrence_species$Common_Name, 
+#                                          occurrence_species$percent_obs)
+## doesn't work?
+
+#plot of observation frequency
+plot_occurrenceDay <- ggplot(occurrence_species, aes(YearDay, percent_obs)) +
+  #set alpha so clustered points visible
+  geom_point(alpha = .1) +
+  #geom_line(aes(color = as.factor(Year))) +
+  labs(title = "Species Occurrence by Date",
+       x= "Julian Date \n (vertical lines denote 1st of April, May, ... 
+       and Oct in non-leap years)",
+       y = "Percent of All Observations Species Composed") + 
+  #geom_line(alpha = .08) +
+  #set background to white for clarity
+  theme_bw() +
+  facet_wrap(~Common_Name, ncol = 4)
+
+##...each year
 occurrence_species2 <- target_data %>%
+  filter(!is.na(obs_date)) %>%
   group_by(Year) %>%
   mutate(all_obs = sum(obs_date)) %>%
   ungroup() %>%
+  
   #find percent of observations accreditable to each species
-  group_by(Species, Year) %>%
-  mutate(percent_obs = obs_date/all_obs)
+  group_by(Year, Species) %>%
+  mutate(percent_obs = sum(obs_date)/all_obs) %>%
+  slice(1)
 
[…]

### 2019-05-08 analysis/scripts/survey_access.R

@@ -1,19 +1,11 @@
 ############
 # Measure site accessibility
 ########
-# Libraries used: 
+# Libraries used: dplyr, ggplot2
 ########
 # Madeleine Ward
 # Last Edit: 2019 May 7
 ########
-
-#SECOND graph: [THIS IS SECONDARY PRIORITY] x axis is Time of Day, 
-#perhaps broken in to 1 or 2 hour blocks , y axis is ....
-#this is a tough one so totally open to your ideas. How about y axis is 
-#% [0, 1.0] and for each year there is a different curve showing the % of
-#planned surveys (as defined above) that were successfully completed that 
-#season within that time block; facet by site. So we are seeing whether there 
-#are consistently problematic times of day at each site. 
 
 temp_access <- target_data %>%
   mutate(time_block = substr(Start_Time, 
@@ -31,7 +23,6 @@ plot_access <- ggplot(temp_access %>%
   facet_wrap(~GIS_Location_ID, ncol = 3 
              #, scales = "free_x"
              )
-plot_access
 
 #plot_yearaccess <- ggplot(temp_access %>%
 #                            filter(!is.na(GIS_Location_ID)) %>%
@@ -45,9 +36,20 @@ plot_access
 #  )
 #plot_yearaccess
 
+temp_access3 <- target_data %>%
+  mutate(time_block = as.numeric(substr(Start_Time,
+                                        nchar(Start_Time)-4, 
+                                        nchar(Start_Time)-3))) 
+plot1 <- ggplot(temp_access3, aes(time_block)) +
+  geom_histogram()
+
 temp_access2 <- target_data %>%
   mutate(time_block = substr(Start_Time, 
                              nchar(Start_Time)-4, nchar(Start_Time)-3)) %>%
+  #mutate(block = as.numeric(substr(Start_Time, 
+  #                                 nchar(Start_Time)-4, 
+  #                                 nchar(Start_Time)-3)),
+  #       time_block = ifelse())
   group_by(GIS_Location_ID, Year, time_block) %>%
   arrange(GIS_Location_ID, Year, time_block) %>%
   select(GIS_Location_ID, Year, YearDay, time_block, Species, Num_Obs) %>%
@@ -59,23 +61,44 @@ temp_access2 <- target_data %>%
   ungroup() %>%
[…]

### 2019-05-08 analysis/scripts/survey_dates.R

@@ -3,7 +3,7 @@
 # need to be modified to account for any observed changes in phenology
 # of the target species.
 ########
-# Libraries used: dplyr, ggplot2, ggforce
+# Libraries used: dplyr, ggplot2
 ########
 # Madeleine Ward
 # Last Edit: 2019 Mar 17
@@ -17,9 +17,11 @@ firstlast_surveys <- species_data %>%
 
 #dataset with both start and end dates for each target species
 target_start_end <- target_data %>%
+  
   # find first/last days, by year
   group_by(Common_Name, Year) %>%
   arrange(YearDay) %>%
+  
   # take first and last day for each group
   slice(c(1, n())) %>%
   # mark whether it is the first or last day each year
@@ -46,9 +48,7 @@ labs(title = "Date of First Observation by Year", y = "Julian Date \n
   geom_hline(yintercept=c(yday(ymd(paste("2007-",c(4:6),"-1",sep="")))),
              linetype=2, colour="grey") + 
   # facet by species; try to separate into visible columns
-  facet_wrap_paginate(~Common_Name, ncol = 3
-             #, scales = "free_y")
-  )
+  facet_wrap(~Common_Name, ncol = 3)
 
 ## last sighting of each species
 plot_timing_2 <- ggplot() +
@@ -64,26 +64,6 @@ plot_timing_2 <- ggplot() +
   labs(title = "Date of Last Observation by Year", y = "Julian Date \n 
        (vertical lines denote 1st of April, May, ... 
        and Oct in non-leap years)")  +
-  #scale_color_discrete(values = c("First Day of Survey Season" = "black")) + 
   geom_hline(yintercept=c(yday(ymd(paste("2007-",c(4:6),"-1",sep="")))),
              linetype=2, colour="grey") + 
-  facet_wrap_paginate(~Common_Name, ncol = 3)
-
-## layering both first/last sightings
-#plot_timing_3 <- ggplot() +
- # geom_point(data =target_start_end, 
-#             aes(x=Year, y=YearDay, color = Start_End)) +
-#  geom_line(data =firstlast_surveys %>%
-#              slice(1), 
-#            aes(x=Year, y=YearDay), color="black") +
-#  geom_line(data =firstlast_surveys %>%
-#              slice(n()), 
-#            aes(x=Year, y=YearDay), color="black") +
-  # titles and Julian date labels
[…]

### 2019-05-08 analysis/scripts/survey_sites.R

@@ -4,29 +4,32 @@
 # Libraries used: dplyr, lubridate, ggplot2
 ########
 # Madeleine Ward, Joel Reynolds
-# Last Edit: 2019 Mar 24
+# Last Edit: 2019 May 8
 ########
 
-#create data frame with all survey dates at each unit, along with
-#number of observations and duration at each
-# NOTE: based on all species;
-#TO DO: filter to just target species data
-survey_units <- survey_duration %>%
-  #chose arbitrary value for considering a success based on seeing any birds (observations above 0)
-  mutate(is_success = ifelse(total_obs_site == 0, "No", "Yes")) %>%
-  filter(!is.na(GIS_Location_ID)) # ignore observations not associated with a survey site
-
-# based on all species;
-#TO DO: filter to just target species data
-plot_unitSuccess <- survey_units %>%
-  ggplot(aes(YearDay,Year)) +
-  #filled vs. unfilled shape shows viable survey
-  geom_point(col="black", size=1,
-             aes(shape = as.factor(is_success))) +
+plot_unitSuccess <- ggplot() +
+  
+  # add layer with open circles for all dates in each site
+  geom_point(data = species_data %>%
+               distinct(Date, .keep_all = TRUE) %>%
+               select(Date, Year, YearDay, GIS_Location_ID) %>%
+               filter(!is.na(GIS_Location_ID)),
+             #plot with open circle shape & greater transparency
+             aes(x = YearDay, y = Year), shape = 1, size = 3, alpha = .8) + 
+  
+  # add layer with successful surveys
+  geom_point(data = survey_duration %>%
+               filter(total_obs_site > 0, !is.na(GIS_Location_ID)),
+             aes(YearDay, Year),
+             col="black", size=3) +
   geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
              linetype=2, colour="grey") +
-  labs(x="Julian Date \n (vertical lines denote 1st of April, May, ... and Oct in non-leap years)") +
-  #manually define shape values
-  scale_shape_manual(values = c(21, 16), breaks = c("Yes", "No"), name="Completed Survey?") +
-  #facet by unit location
+  labs(x = "Julian Date \n (vertical lines denote 1st of 
+       #April, May, ... and Oct in non-leap years)",
+       title = "Survey Success Across Sites (Measured by Non-Zero Obs)") +
+  #make background white
+  theme_bw() +
+  
+  # facet by unit
   facet_wrap(~GIS_Location_ID)
+

### 2019-05-08 analysis/scripts/survey_timing_EDA.R

@@ -2,31 +2,37 @@
 # Basic data visualization of survey timing (from species_data object),
 # ultimately leading to further filtering of that object for later analyses.
 ########
-# Libraries used: ggplot2,
+# Libraries used: ggplot2, dplyr
 ########
 # Joel Reynolds (repurposing code by Madeleine Ward)
-# Last Edit: 2019 Apr 27
+# Last Edit: 2019 May 8
 ########
 
 ## Display the dates surveys were conducted across each year
 
 plot_alldates <- species_data %>%
+  distinct(Date, .keep_all = TRUE) %>%
   ggplot(aes(YearDay,Year)) +
-  geom_point(col="black", size=1.5) +   # Draw points
+  geom_point(size=3, alpha = .9) +   # Draw points
+  # Add lines to denote months
   geom_vline(xintercept=c(yday(ymd(paste("2007-",c(4:10),"-1",sep="")))),
              linetype=2, colour="grey") +
   labs(x="Julian Date \n (vertical lines denote 1st of April, May,
-       ... and Oct in non-leap years)")
+       ... and Oct in non-leap years)") +
+  theme_bw()
+
 
 ## Display sightings for each species collapsed across sites for given survey date
 
 plot_phenology <- ggplot() +
+  
   #add layer for all possible survey dates
   geom_point(data = species_data %>%
                distinct(Date, .keep_all = TRUE) %>%
                select(Date, Year, YearDay),
              #plot with open circle shape & greater transparency
              aes(x = YearDay, y = Year), shape = 1, alpha = .5) +
+  
   #add layer for species-specific sighting dates, closed circles
   geom_point(data = target_data,
              aes(x = YearDay, y = Year)) +
@@ -36,6 +42,7 @@ plot_phenology <- ggplot() +
   labs(x="Julian Date \n (vertical lines denote 1st of April, May,
        ... and Oct in non-leap years)",
        title= "Survey Success for Target Species") +
+  
   #temp; make background white for clearer contrast
   theme_bw() +
   #facet by target species; common name used

### 2019-05-08 analysis/scripts/units_species.R

@@ -1,59 +1,0 @@
-#Value of all units - only some spp in unit X?; “unit by species occurrence” relationships?
-############
-#
-########
-# Madeleine Ward
-# Last Edit: 2018 Oct 22
-########
-
-tbl_Locations$GIS_Location_ID <-
-  str_sub(tbl_Locations$GIS_Location_ID,start=-1)
-
-
-species_units <-
-  left_join(species_data <- left_join(tbl_Field_Data, tbl_Events,
-                                      by = "Event_ID") %>%
-              left_join(tlu_species_codes,
-                        by = c("Species" = "Code")) %>%
-              separate(Start_Date, c("Date", "Time_0"), sep = " ") %>%
-              separate(Start_Time, c("Date_01", "Time_Start"), sep = " ") %>%
-              #how to fit all these into one argument??
-              separate(End_Time, c("Date_02", "Time_End"), sep = " ") %>%
-              mutate(Date = mdy(Date)) %>%
-              filter(Year >= 2003) %>%
-              group_by(Year) %>%
-              arrange(Date) %>%
-              mutate(Days_since_first = Date - Date[row_number()==1]),
-            tbl_Locations,
-            by = "Location_ID") %>%
-  group_by(GIS_Location_ID, Year, Days_since_first, Species) %>%
-  summarize(total_Observed = sum(`Num_ Observed`)) %>%
-  left_join(tlu_species_codes, by = c("Species" = "Code"))# %>%
-#optional: spread command
-#spread(GIS_Location_ID, total_Observed)
-
-#to do: make plot
-plot4 <- ggplot(species_units, aes(as.numeric(Days_since_first),
-                                   total_Observed,
-                                   color = as.factor(Year))) +
-  geom_jitter(height = .3, aes(fill = GIS_Location_ID)) +
-  #in ggforce package to supposedly distribute facets over pages
-  facet_wrap_paginate(~Species, ncol = 1, scales = "free") +
-  ylim(c(0, 2000))
-print(plot4)
-
-#how to
-#1. order bars with a y aesthetic,
-#2. have species not take up so much space
-#3. get rid of 0/NA values to take up less space?? ("free_x" not working)
-plot5 <- ggplot(species_units, aes(Species, total_Observed, fill = as.factor(Year))) +
-  geom_bar(stat = "identity") +
-  coord_flip() +
-  #in ggforce package to supposedly distribute facets over pages
[…]

### 2019-05-24 analysis/paper/paper.Rmd
changes to occurrence, survey access plots

@@
 So we are seeing whether there are consistently problematic times of day at each site.
-> JOEL: some updates on this!
-for now, times are in hour blocks.
-There can't be lines/curves for year since they are unreadable (up/down).
-I also tried to use color as a factor but then the graphs are crowded (so many years).
-So I have 2 alternatives, I don't really like either but some are more readable than others.
-See below.
+> JOEL: for now, times are in hour blocks
 
@@
 
-print(plot_success)
-```
-
-```{r Fig-SiteAccess2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis',fig.height=15, fig.width=12, fig.cap="",warning= FALSE}
-
-print(plot_success3)
+#print(plot_access)
 ```
@@
 x axis is Day of Season, y axis is % as decimal [0, 1.0], curves for each year; facet by species.
-> JOEL: I did some graphs for species occurrence that fit the description but not sure if they were what you were looking for?
-They do show species occurrence relative to others; as annual/daily percentage of daily observations in which spp was sighted (percentage calculated by annual/daily obs per species over all total obs - is this what you meant?
-let me know).
-Lines didn't work though (lots of ups and downs)
 
@@
 
-print(plot_occurrenceDay)
+print(plot_occurrenceYear)
 ```
@@
 
-print(plot_occurrenceYear)
+#print(plot_occurrenceDate)
 ```
@@
 ```{r Covariates, eval=TRUE, cache = FALSE, include = FALSE}
-source("../scripts/temp_covariates.R")
+#source("../scripts/temp_covariates.R")
 ```

### 2019-05-24 analysis/scripts/occurrence_freq.R

@@ -7,63 +7,66 @@
 # Last Edit: 2019 May 8
 ########
 
-# graph shows observations each year relative to other TARGET species
-#percent_obs will be the percent of observations each date that 
-#the spp composed
-occurrence_species <- target_data %>%
-  group_by(Year, YearDay) %>%
-  
-  #find all observations of all species on certain date
-  mutate(all_obs = sum(obs_date)) %>%
-  ungroup() %>%
-  
-  #find percent of observations accreditable to each species
-  group_by(Year, YearDay, Species) %>%
-  mutate(percent_obs = obs_date/all_obs) %>%
-  ungroup() 
+#percent of surveys, each year, in which species was sighted
 
-#attempt to reorder facets by total percent of observations
-#occurrence_species$Common_Name <- reorder(occurrence_species$Common_Name, 
-#                                          occurrence_species$percent_obs)
-## doesn't work?
+#collect total surveys each year
+all_surveys_yr <- species_data %>%
+  distinct(Event_ID, .keep_all = TRUE) %>%
+  group_by(Year) %>%
+  summarize(total_surveys = n())
 
-#plot of observation frequency
-plot_occurrenceDay <- ggplot(occurrence_species, aes(YearDay, percent_obs)) +
-  #set alpha so clustered points visible
-  geom_point(alpha = .1) +
-  #geom_line(aes(color = as.factor(Year))) +
-  labs(title = "Species Occurrence by Date",
-       x= "Julian Date \n (vertical lines denote 1st of April, May, ... 
-       and Oct in non-leap years)",
-       y = "Percent of All Observations Species Composed") + 
-  #geom_line(alpha = .08) +
-  #set background to white for clarity
+#collect total surveys each year for each species, the percent of sighted surveys
+occurrence_species_yr <- target_data %>%
+  group_by(Common_Name, Year) %>%
+  #all surveys in which Spp were sighted each year
+  summarize(species_surveys = n()) %>%
+  #join by Year to compare with total
+  left_join(all_surveys_yr, by = "Year") %>%
+  #calculate percent of yearly surveys in which Spp sighted
+  mutate(percent_surveys = species_surveys/total_surveys)
+
+#plot of species occurrence by Year
+plot_occurrenceYear <- ggplot(occurrence_species_yr,
+                          aes(Year, percent_surveys)) +
+  geom_point() +
[…]

### 2019-05-24 analysis/scripts/survey_access.R

@@ -4,101 +4,44 @@
 # Libraries used: dplyr, ggplot2
 ########
 # Madeleine Ward
-# Last Edit: 2019 May 7
+# Last Edit: 2019 May 24
 ########
 
-temp_access <- target_data %>%
+# create frame of all time blocks
+all_dates <- target_data %>%
+  #each hour is a block, from 0-24
   mutate(time_block = substr(Start_Time, 
                              nchar(Start_Time)-4, nchar(Start_Time)-3)) %>%
-  group_by(GIS_Location_ID, Year, time_block) %>%
-  mutate(site_time_obs = sum(Num_Obs)) %>%
-  ungroup()
+  
+  #keep only one copy of date and select year, date, and block
+  distinct(Date, .keep_all = TRUE) %>%
+  select(Date, Year, YearDay, time_block) %>%
+  group_by(Year, time_block) %>%
+  #count all surveys performed in the time block
+  summarize(block_instance = n())
 
-# 
-plot_access <- ggplot(temp_access %>%
-                        filter(!is.na(GIS_Location_ID), !is.na(time_block))) +
-  geom_boxplot(aes(time_block, site_time_obs)) +
-  coord_flip(ylim = c(0, 1200)) +
-  labs() +
-  facet_wrap(~GIS_Location_ID, ncol = 3 
-             #, scales = "free_x"
-             )
-
-#plot_yearaccess <- ggplot(temp_access %>%
-#                            filter(!is.na(GIS_Location_ID)) %>%
-#                            group_by(Year, GIS_Location_ID) %>%
-#                            mutate(site_year_obs = sum(Num_Obs))) +
-#  geom_boxplot(aes(Year, site_year_obs)) +
-  #coord_flip(ylim = c(0, 1200)) +
-  # coord_cartesian(ylim = c(0, 1000)) +
-#  facet_wrap(~GIS_Location_ID, ncol = 3 
-             #, scales = "free_x"
-#  )
-#plot_yearaccess
-
-temp_access3 <- target_data %>%
-  mutate(time_block = as.numeric(substr(Start_Time,
-                                        nchar(Start_Time)-4, 
-                                        nchar(Start_Time)-3))) 
-plot1 <- ggplot(temp_access3, aes(time_block)) +
-  geom_histogram()
-
-temp_access2 <- target_data %>%
+# UNFINISHED: MESSY AND ERRORED
+# this frame is meant to collect surveys in time block per year for each site;
[…]

### 2019-05-24 analysis/scripts/survey_sites.R

@@ -4,20 +4,19 @@
 # Libraries used: dplyr, lubridate, ggplot2
 ########
 # Madeleine Ward, Joel Reynolds
-# Last Edit: 2019 May 8
+# Last Edit: 2019 May 23
 ########
 
 plot_unitSuccess <- ggplot() +
   
-  # add layer with open circles for all dates in each site
+  # add layer with open circles for all dates
   geom_point(data = species_data %>%
                distinct(Date, .keep_all = TRUE) %>%
-               select(Date, Year, YearDay, GIS_Location_ID) %>%
-               filter(!is.na(GIS_Location_ID)),
+               select(Date, Year, YearDay),
              #plot with open circle shape & greater transparency
              aes(x = YearDay, y = Year), shape = 1, size = 3, alpha = .8) + 
   
-  # add layer with successful surveys
+  # add layer with successful surveys in each site
   geom_point(data = survey_duration %>%
                filter(total_obs_site > 0, !is.na(GIS_Location_ID)),
              aes(YearDay, Year),

### 2019-05-24 analysis/scripts/temp_covariates.R

@@ -1,127 +1,0 @@
-############
-# Examine various covariates and distributions.
-########
-# Libraries used: ggplot2, stringr
-########
-# Madeleine Ward
-# Last Edit: 2019 Apr 27
-########
-library(stringr)
-### Distribution of weather conditions
-
-## Distribution of wind speed by year
-plot_wind <- ggplot(tbl_Events %>%
-                      left_join(tbl_Locations, by = "Location_ID") %>%
-                      filter(Year > 2003, !is.na(Wind_code), !is.na(GIS_Location_ID)),
-                    aes(as.factor(Year))) +
-  geom_bar(aes(fill = as.factor(Wind_code))) +
-  scale_fill_discrete(labels = tlu_wind_codes$Wind_Speed,
-                      name = "Wind Speed") +
-  labs(title = "Distribution of Wind Conditions Across Years",
-       x = "Year", y = "Count")
-print(plot_wind)
-## Wind speed by year & site
-plot_windsite <- plot_wind +
-  facet_wrap(~GIS_Location_ID)
-print(plot_windsite)
-
-### Distribution of wave conditions
-
-## Distribution of wave height by year
-plot_waves <- ggplot(tbl_Events %>%
-                      left_join(tbl_Locations, by = "Location_ID") %>%
-                      filter(Year > 2003, !is.na(Wave_ht), !is.na(GIS_Location_ID)),
-                    aes(as.factor(Year))) +
-  # still need to change order of fill var
-  geom_bar(aes(fill = as.factor(Wave_ht))) +
-  scale_fill_discrete(labels = tlu_wave_codes$Description,
-                      name = "Waves") +
-  labs(title = "Distribution of Wave Conditions Across Years",
-       x = "Year", y = "Count") +
-  coord_flip() + 
-  theme(legend.position="bottom")
-print(plot_waves)
-## Wave height by year & site
-plot_wavesite <- plot_waves +
-  facet_wrap(~GIS_Location_ID, ncol =2)
-print(plot_wavesite)
-
-## Distribution of tide states by year
-#many problems with this graph...
-plot_tides <- ggplot(test <- tbl_Events %>%
-                       left_join(tbl_Locations, by = "Location_ID") %>%
-                                   mutate(All_tides = substr(Tide_Level, start = 1, stop = 1), locale = "en") %>%
-                       filter(Year >= 2003
[…]

### 2019-08-12 analysis/paper/paper.Rmd
edits to catch up post Madeleine, prep for KLGO call.

@@
 ---
-title: "Assessment of the KLGO Coastal Waterbird Survey Monitoring Program"
+title: "Status Update regarding \n Assessment of the KLGO coastal waterbird survey monitoring program"
 author:
@@
 output:
+  bookdown::word_document2:
+    fig_caption: yes
+    toc: no
+  pdf_document:
+    fig_caption: yes
+    toc: no
+  word_document:
+    toc: no
+    fig_caption: true
+    reference_docx: NRR_Author_Template_V4.2.docx
   html_document:
@@
     number_sections: yes
-    toc: yes
-  bookdown::word_document2:
-    fig_caption: yes
-    toc: yes
-  word_document:
-    toc: yes
-    fig_caption: true
-    reference_docx: NRR_Author_Template_V4.2.docx
-  pdf_document:
-    fig_caption: yes
-    toc: yes
+    toc: no
 editor_options: 
@@
 bibliography: KLGOCWS_References.bib
-abstract: The Coastal Waterbird Survey has been conducted, with varying levels of
+abstract: DRAFT ABSTRACT The Coastal Waterbird Survey has been conducted, with varying levels of
   effort and in-season frequency, since 2003 [@Surdyk2018]. The area of interest has
@@
   and assess how things are going. We review the current program's objectives and survey design,
-  identify statistical issues, and recommend potential solutions. Major topics include... blah blah. The study concludes with a series of recommended tasks that the park should
+  identify statistical issues, and recommend potential solutions. Major topics include... etc. The study concludes with a series of recommended tasks that the park should
   undertake to improve the coastal waterbird survey's efficiency and effectiveness.
 ---
-
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
 
@@
 ```
-# NOTES To Madeleine
-- see edits in survey_duration.R to aid interpretation by the viewer (e.g., changes in x axis tick locations, etc.) $\checkmark$
-- documentation in scripts $\checkmark$
-- **NEXT Priority Milestones**: note - all references to line numbers are prior to any other edits!
-- finish graphs of first & last observation per year for each target species (around line 386);**ultimately** add a fitted linear trend line to each of those (we need to account for frequency of surveying, so some subtle effort things here, but can resolve that later); $\checkmark$
-- graphs of species occurrence (around line 407) for each of target species; $\checkmark$
-- covariate distribution graphs (lots of univariate & bivariate things; around line 534);
-- plot of total # spp sighted each season across years (around line 600)
 
-# Issues to Resolve with Rmarkdown, Knitr, Github, etc.
-- conversion from paper.md to paper.pdf (pandoc issue?) 
-- adding files to .gitignore
+# Overview of Progress on this TAR
+* Brief overview of activities
+  + 2018 Summer - site visit KLGO; discussions w/ Jamie, Elaine Furbish; gather protocol and past report materials, database, etc.
+Develop assessment strategy and report outline; initial review of past reports and related materials.
+Recruit 10 hr / wk volunteer undergraduate student intern (Madeleine Ward).
+  + 2018 Fall-Winter - bulk of effort ended up focused on working with Madeleine and NPS WASO IT staff to set up computing systems for functional collaborative workflow (e.g., establich NPS github repository and train intern in use of 'version control systems' to automate synchronizing of our files for collaboration, resolving various IT issues, setting up project folder and file structures for long-term use and archiving); initial assessment of program goals and objectives; resolve data access issues;
+  + 2019 Spring - initiate basic summary graphics for clarifying timing and frequency of survey events, variation in duration of survey events (e.g., variation in survey effort at a site); timing and frequency of spp-specific sightings; intial development of phenology graphics.
+INTERNSHIP ENDED
+  + 2019 Summer (minimal time) - prep for discussion w/ KLGO Resources staff.
 
-# Issues to Resolve with content (narrative, figures, analyses)
-- **Joel & Madeleine** will discuss survey_dates.R commands and how to modify them so that we get a sense of when surveys were & when target species was first (or last) seen each year.
-E.g., separate out any filtering process stemming from survey effort levels and timing.
-(thinking some thing like showing open circles for survey events PRIOR to first sighting...)
-- Need to think about way to winnow out target species that do not occurr w/ adequate frequency to report on.
-E.g., those only seen <3 or 5 years???
-Look at phenology graphs and determine threshold to apply.
-Consider some species N-ward and S-ward migration routes may differ.
-- Joel will finish script uploading AOU data csv object & using to order Target Species
-- What is up with 2008 surveys? why such distinct timing?
-- Think about Covariate EDA, if any...change in observers?
-- Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
-- clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.
+The following excerpt from the planned report lists 4 major topics or phases that should be addressed in a full assessment of a monitoring program.
+The most important is the first topic - the *Why* motivating the program; e.g., the program's Problem Framing and Information Objectives.
+I've provided draft feedback on this and a number of key questions that require resolution by KLGO staff prior to diving into any detailed analyses *so as to ensure futher detailed review of data collection & analyses target KLGO's desired Information Objectives*.
+I've also included some of the basic survey effort data visualizations that we've already developed but haven't spent time on the associated narrative.
 
+***
 # Introduction
@@
 This is sufficient history to warrant stepping back and assessing the monitoring program.
-Such assessments are key to an adaptive monitoring program [@Lindenmayer2009], are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016], and align with guidance from NPS' Inventory and Monitoring Division [@Gallo2018a, @Mitchell2018].
+Such assessments are key to an adaptive monitoring program [@Lindenmayer2009], are an important and distinct phase to promote program, and institutional, learning and evolution [@Reynolds2016], and align with guidance from NPS' Inventory and Monitoring Division [@Gallo2018a], [@Mitchell2018].
 
-Assessing a monitoring program focuses on four broad topics following the major Phases of a monitoring program (ibid):  
+Assessing a monitoring program focuses on four broad topics, each associated with a major development phase of the monitoring program (ibid):  
 1. _Why_ - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?  
@@
 
-This assessment addressed all four topics, but focused predominantly on topics (i) Problem Framing & Information Objectives, (ii) Data Collection Design, and (iii) Data Analysis Design.
+This assessment aims to address all four topics, but will focus predominantly on topics (i) Problem Framing & Information Objectives, (ii) Data Collection Design, and (iii) Data Analysis Design.
 Since 'Problem Framing & Information Objectives' determine the relevancy of many of the potential assessment actions related to the other topics,  the assessment was drafted as a first stage of an iterative effort, providing high level feedback, including listings of potential detailed syntheses and workflow enhancement activities (such as improved database interface, automation of annual analyses and updating of syntheses, etc.).
 With this foundation, further dialogue with KLGO staff regarding relevant 'Problem Framing & Information Objectives' and other priorities will determine which follow-on activities to pursue.
-
 <!-- Key Findings / Next Steps -->
@@
 
-# Methods
+# Assessment Methods
 The CWS monitoring program was compared against a recommended checklist of monitoring program components developed from published literature (Appendix 1).
@@
 
-The management problem or question motivating the need for information about coastal waterbirds provides the foundation for the survey’s data collection and analysis designs [@Fancy2009].
-It clarifies the underlying goal of the monitoring program and answers “Why is this an important problem or question?”.
+Ideally, the survey’s data collection and analysis designs are founded on clearly articulated *Information Needs* stemming from a clearly articulated management problem or question or goal  [@Fancy2009].
+Such a problem or question or goal answers “Why is this information important?”.
 
@@
 ### Assessment findings:
+#### Motivating Management Problem, Question or Goal
 The 2003 coastal waterbird inventory and survey design was developed to address "a top priority bird inventory (need)" of the Southeast Alaska Inventory and Monitoring Network (Sharman et al.
@@
 
-The inventory's stated goal was to document the occurrence of 90% of the species of waterbirds likely to occur in KLGO (ibid).
-The principal objectives were “to develop scientifically valid sampling designs for inventorying waterbirds and breeding landbirds in KLGO, and to collect baseline information on bird species occurrence, distribution, abundance, and habitat associations in the park…[to] ...provide park managers with timely information that facilitates land management planning, helps track wildlife population and ecosystem change over time, and keeps resource managers apprised of changes that may require action before a crisis occurs” (ibid).
+The inventory's stated goal was to document the occurrence of 90% of the species of waterbirds likely to occur in KLGO "...
+[to] ...provide park managers with timely information that facilitates land management planning, helps track wildlife population and ecosystem change over time, and keeps resource managers apprised of changes that may require action before a crisis occurs" (ibid).
+
+=> Thus the program appears to be intended to serve management information needs associated with land use planning, provide 'surveillance monitoring' of population changes and trends through time, and provide red-flags for potential developing crises.
+
+**Recommendations**
+
+=> **Park Resource staff should clarified and further refined the motivating needs in order to allow assessment of the adequacy of the current survey design and effort levels.**
+
+What actual land management questions might such information inform?
+Have any occurred in the last 20 years?
+If that remains a valid goal, can the types of questions and decisions be refined and clarified?
+What sort of crises?
+How quickly do they need to be 'flagged'?
+What sort of ability to detect trend?
+Only local trend or regional or...?
+Trend in phenology or occurrence frequency or abundance or some other population feature?
+Clarifying these goals matters, for example, since monitoring phenology requires much less intensive survey effort than effectively monitoring abundance.
+
+#### Motivating Management Information Objectives
+The principal objectives were identified as (numbering added) “(i) to develop scientifically valid sampling designs for inventorying waterbirds and breeding landbirds in KLGO, and to collect baseline information on bird species (ii) occurrence, (iii) distribution, (iv) abundance, and (v) habitat associations in the park ” (ibid).
+
 While no Threatened or Endangered bird species were known to occur in KLGO, as of 2004, nine species on the Alaska Watchlist had been documented in the park (ibid), thus suggesting potential priority information needs associated with monitoring for changes in those species occurrence and abundance.
+
 Potential secondary objectives mentioned include "investigations into the breeding ecology, phenology, reproductive success and population age structure of KLGO breeding birds" (ibid).
@@
 
-Stakeholders mentioned or implied in park reports include park superintendent and related managers, those charged with monitoring regional and internationally important bird habitats, the Skagway Bird Club, the Alaska Landbird Monitoring Survey and the Boreal Partners in Flight.
+=> depending on how you classify things, this is up to 12 distinct information objectives.
+Distinct information objectives may well require very different surveys methods, if not at least different levels of survey effort, and analyses.
 
-The coastal waterbird surveys have been continued since 2003 [@Surdyk2018].
+**Recommendations**
 
-QUOTES
-Hahr and Trapp (2004):
-We intend to use the results of these inventories to create a revised bird checklist and an atlas database for the Taiya and Skagway River watersheds in collaboration with the Skagway Bird Club.
+=> **Park Resource staff should revisit and evaluate the relevance and importance or each of these distinct information objectives**, ideally ranking them at least into 'essential' versus 'nice if feasible' categories.
+This will provide the targets for actually assessing adequacy of the current survey design and analysis methods.
 
-This protocol addresses one monitoring question with three objectives: (1) what is the annual nest density and reproductive (fledgling) success in known Hawaiian petrel colonies, (2) what are the long-term trends in colony distribution and density monitored in approximate 5-year intervals, and (3) are these affected by predator control?
-The first goal of monitoring is to obtain unbiased estimates of Hawaiian petrel nest density and reproductive (fledging) success from known colonies in HAVO and HALE in order to detect changes in colony growth or decline.
-The second goal of monitoring is to periodically (approximately every five years) obtain unbiased estimates of nest density from potential Hawaiian petrel habitat.
-This information can be used to assess changes in density and distribution of subcolonies across the landscape.
-The third goal is to estimate nest density and fledging success in areas undergoing different management regimes to assess effectiveness of management.
-Benchmark levels of these estimates could serve as warnings of the need for modified management or further investigation of these colonies.
+#### Key Stakeholders
+Stakeholders mentioned or, more usually, implied in park reports include park superintendent and related managers, those charged with monitoring regional and internationally important bird habitats, the Skagway Bird Club (author of 'significant bird observations in SE AK' and contributor to AK summaries for 'North American Birds'), the Alaska Landbird Monitoring Survey and the Boreal Partners in Flight.
+For example, Hahr and Trapp (2004) mention the intention to develop inventories and revise the bird checklist and atlas database for the Taiya and Skagway River watersheds in collaboration with the Skagway Bird Club.
 
-Though not explicitly characterized in the protocol or reports reviewed for this assessment (Table 1), the following target audiences and information users were identified through conversations with KLGO staff, partners, and other waterbird researchers and stakeholders in Alaska.
-(current)
-- park
-- Skagway Bird Club; compiles for addition to 'significant bird observations in SE AK', which gets compiled in to AK sumamries for 'North American Birds' (mainly spp lists)
-(potential)
-- BPIF
-- PSG
-- Audubon Alaska (for IDing Important Bird Areas - though require accurate density estimates)
-- Audubon North America - cliamte species distribution models (Nicole Michel, Brooke Bateman, etc.); esp if associated w/ repeat sites & effort data (date, time of day, duration, distance/area covered) - see Nicole email 6/7/2018.
-Watchlist of projected changes - [@Wu2018]
+**Recommendations**
 
-### Recommendations:
+=> **KLGO Natural Resources staff should explicitly clarify intended stakeholders and 'information users', clarify their exact information needs in order to assure they will be met (or cannot be met), and develop list of contacts for distributing results and updates (perhaps includes as Appendix in SOP if not already).**
 
-Potential Follow Up Activities:
-
-Define the _Target Audiences_ for the information and their Objectives in using the information
-
-_Decision Makers_
-Who are target audiences for this information?
-Who uses it or may use it?
-Local contact: C. Elaine Furbish, Skagway Bird Club, Google Site
-Predominantly historical park; limited resource staff and capacity
-Visible natural resource of interest to local community and visitors; interp and educational opportunity; topic of information requests; provides ‘community service’ given absence of any other agency with natural resource staff;
-Annual report (time lag?); National Resource Condition Assessment; other? State of the Park?
-Potential for contribution to regional, flyway, and WASO programs (e.g., CCRP & Audubon climate watch, etc.)
-Current sightings passed to local bird club & used by avian-interested visitors
-Used in quarterly regional ‘significant bird observations’, which in turn is used to develop the Alaska-wide summary for the journal North American Birds (published by the American Birding Association http://www.aba.org/nab ) and includes ‘...sightings of birds that are out of range or out of season or occur regularly in small numbers, noteworthy breeding records, unusually large or small numbers of a particular species, unusual migration dates, etc….’ (S.
-C.
-Heinl, 2017, Summary of SE AK Bird Observations, June-July 2017, accessed 2018 June 6).
-Queries to Heather Renner & Robin Corcoran re: PSG-type users, etc.
-Opportunity for community based monitoring to track projected spp impacts of climate change, change in phenology, occurrence, abundance, etc.?
-
-_Decisions_
-What types of decisions is it / will it be used for?
-(specify for each target audience/ end user)
-No specific decisions
-Current:
-KLGO - spp lists, phenology of occurrence (guidance to visitors), trends in richness (NRCA), abundance?
-Skagway bird club - spp list, current occurrences (activity re: rare & unusual)
-SE AK / Regional - spp lists; unusual sightings, etc. (spp occurrences & broad seasonal timing).
-AKR -
-WASO/NPS - ??CCRP??
-Question to Gregor & Mel (AK Audubon).
-Could be used for garnering community engagement in Climate Watch & other national efforts associated with Audubon’s Climate Report.
-Not used in continental analyses.
-“One key message that I want to emphasize is that the focus of this work was purely on climate suitability.
-What this means, as we indicate in the briefs’ “Important” box up front and “Caveats” section further back, is that significant changes in climate suitability, as measured here, will not always result in a species response, and all projections should be interpreted as potential trends.
-Multiple other factors mediate responses to climate change, including habitat availability, ecological processes and biotic interactions, dispersal capacity, species' evolutionary adaptive capacity, and phenotypic plasticity (e.g., behavioral adjustments).
-Ultimately, models can tell us where to focus our concern and which species are most likely to be affected, but monitoring is the only way to validate these projections and should inform any on-the-ground conservation action.” Gregor Schuurman, email to DENA
-
-Potential:
-KLGO: 
-(easy) - changes in occurrence (spp, season scale); changes in phenology (migration timing, by spp; breeding timing, by spp); broad (order of magnitude) changes in abundance (index); broad (?) changes in composition (by spp & timing of survey)?; species occurrence x unit (~habitat?) - refine expected lists per survey and develop targeted SOP training materials;
-(harder) - finer changes in abundance; finer changes in composition and age; ???
-
-What Objectives are the decision makers hoping to achieve by using this information?
-
-Over what spatial scales is the information being used? 
-e.g., at the park, in the community, regional scale, flyway scale, nationally, internationally
-For each spatial scale of interest, what are the associated time scales of interest for detecting changes?
-E.g., annual, w/in season, spring vs fall, etc.
-
-From another report but quotes from Olsen et al. (1999):
-Olsen (et al.
-1999) noted that “Most of the thought that goes into a monitoring program should occur at this preliminary planning stage.
-The objectives guide, if not completely determine the scope of inference of the study and the data collected, both of which are crucial for attaining the stated objectives.” Olsen goes on to say that a “clear and concise statement of monitoring objectives is essential to realize the necessary compromises, select appropriate locations for inclusion in the study, take relevant and meaningful measurements at these locations, and perform analyses that will provide a basis for the conclusions necessary for meeting the stated objectives.”
-
-Fancy, Gross, and Carter (2009):
-The development of monitoring objectives, which provide additional focus about the purpose or desired outcome of the monitoring effort, was an iterative process that sometimes required several years to refine.
-Early in the design process, monitoring objectives were stated in more general terms, such as “Determine trends in the incidence of disease and infestation in selected plant communities and populations”, whereas the final monitoring plan and protocols provided monitoring objectives that met the test of being realistic, specific, and measurable (e.g., “Estimate trends in the proportion, severity, and survivorship of limber pine trees infected with white pine blister rust at Craters of the Moon National Monument”; Garrett et al.
-2007).
-
-Structure from Northeast Coastal and Barrier Network
-https://irma.nps.gov/DataStore/Reference/Profile/2195238
-Justification => Vital Sign => Monitoring Goal => Monitoring Questions => Monitoring Objectives => Measures (eg pg 55 & 56)
-
-Clarify Objectives of the Decision Makers 
-Among the decision maker objectives identified in 1.b.ii, clarify 
-the Fundamental Objective(s) from 
-the Means Objectives, then
-diagram the Objectives Hierarchy.
-Identify the Attributes that need to be monitored to assess achievement of each of the objectives. (‘Concepts by Theory’, Ford 2001)
-Identify the Measurements selected to characterize each attribute. (‘Concepts by Data’, Ford 2001) [WHAT]
-Identify any Threshold values associated w/ each Measurement
-
-Q: compiling and assessing regional trends in waterbirds in SE AK?
-Asked of Heather Renner (Ak Maritime), Robin Corcoran (Kodiak NWR)
-Answers:
-Not really; regular discussion at PSG but no one has taken it on. 
-Kathy Kuletz works on statewide trends in at sea seabird observations;  Piatt GLBA work (& Martin Renner); 
-Robin C: “There's always talk at the PSG meeting about the North Pacific seabird colony database, or some variation of that database for marine birds, but it never gets anywhere.
-Basically there is no money for a database manager.
-I've been submitting my data to Rob Kaler in Migratory Bird Management, but I think that's just for storage until the money for a database manager materializes.
-More recently Rob Suryan w/NOAA has been interested - for the North Pacific Pelagic Seabird Database.
-You might want to contact him, he's leading the Gulf Watch Alaska Long-Term Ecosystem Monitoring Program, and maybe interested in this work.”
-
-NPS/Audubon:
-Wu et al.
-2018.
-Projected avifaunal responses to climate change across the U.S.
-NPS.
-PLOS ONE.
-Appendices include projected park-specific turnover, colonization, and extirpations projections - potential to test with this survey.
-Would require discussions w/ CCRP & ARO - ideally as a coordinated effort with, say, AK Audubon and ACF(?) or LCCs as tracking metric.
-Appendices include “Potential management goals and activities for parks, organized by trend group.”
-
-Melanie AK Audubon: IBA requires demonstrating higher than ‘usual’ densities (relative to surrounding area).
-
-Nicole Michel & Brooke Batemen Audubon National Science Team: of interest and possible use in national scale quantitative analyses or climate species distribution models.
-“Scaling up and integrating datasets across regions, including data from different goals, is something that we’re starting to do in other regions.
-“
-
-Occurrence, Richness
-Phenology (migration, breeding)
-Abundance - Even without the structure needed to estimate detection probability, if the same protocol is used annually this data could be used to produce long-term trend indices for the study area.
-(competing staff priorities - peak (?) usually aligns with hooligan run. Possibly reduce summer frequency?)
-
-Diagram the Conceptual Model of the system
-E.g., create an influence diagram, being sure to denote the
-Fundamental & Means Objective(s)
-Factors influencing their achievement (aka ‘system drivers’)
-Associated management or policy decisions and/or potential actions.
-Include a written summary of the literature upon which the conceptual model is based.
-
-Summarize the management or policy decisions the monitoring is to inform (if any) based on 3.c.
+=> Additional potential stakeholders include Audubon North America as part of their national Climate Watch program, which develops species distribution models under projected future climates.
+This group is especially interested in data from areas with repeat observations at sites and that include recording of effort data (date, time of day, duration, distance/area covered, etc.).
+An example of their products is the recent watchlist of projected changes [@Wu2018].
+(Nicole Michel, Brooke Batemen; see email from Nicole to Joel 6/7/2018)
 
 ## Phase 2 - Monitoring & Analysis Design
+**NOTE TO READER**: The rest of this is a mix of outline and notes, with some initial data visualizations. This is far from a draft narrative. **Proceed with Caution**
+
 HOW monitor? Clarify the various design choices.
@@
 
-*Site Accessibility*
-How consistently are they able to conduct surveys at each site? E.g., how often is Site 1 dropped due to cruise ships?
+*Temporal* 
+```{r EDA_SurveyTiming, eval=FALSE, cache = FALSE, include = FALSE}
+# dotplot of survey times by year
 
-> SECOND graph: [THIS IS SECONDARY PRIORITY] x axis is Time of Day, perhaps broken in to 1 or 2 hour blocks , y axis is ....this is a tough one so totally open to your ideas.
-How about y axis is % [0, 1.0] and for each year there is a different curve showing the % of planned surveys (as defined above) that were successfully completed that season within that time block; facet by site.
-So we are seeing whether there are consistently problematic times of day at each site.
-> JOEL: for now, times are in hour blocks
-
-```{r SiteAccess, eval=TRUE, cache = FALSE}
-
-source("../scripts/survey_access.R", echo=FALSE)
+source("../scripts/survey_timing_EDA.R", echo=FALSE)
 ```
 
-```{r Fig-SiteAccess1, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis',fig.height=20, fig.width=12, fig.cap="",warning= FALSE}
+```{r Fig-SurveyTiming1, echo=FALSE, eval=FALSE, cache=FALSE, fig.show= 'asis',fig.height=8, fig.width=12, fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
 
-#print(plot_access)
+print(plot_alldates)
 ```
+The annual dates of the first and last survey each year has varied by over a month (Figure \@ref(fig:Fig-SurveyTiming)).
+
+Questions: what is cue to start of survey season?
+
+*Survey Completion*
+
+```{r SiteSurveyCompletion, eval=TRUE, cache = FALSE, include = FALSE}
+# dotplot of survey occurrences by unit and whether there were observations
+
+source("../scripts/survey_sites.R", echo=FALSE)
+```
+
+```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=15, fig.width=18, fig.cap="Need to add caption.",warning=FALSE}
+
+print(plot_unitSuccess)
+```
+Add sentence or two about success in completing surveys. Doesn't appear to be any consistently problematic sites, not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).
 
@@
 All data, original data forms, and field notes were archived at KLGO.
-
-*Survey Completion*
-> Ultimately, re-order these sections to flow better. Completion then Duration, etc., etc.
-
-```{r SiteSurveyCompletion, eval=TRUE, cache = FALSE, include = FALSE}
-# dotplot of survey occurrences by unit and whether there were observations
-
-source("../scripts/survey_sites.R", echo=FALSE)
-```
-
-Add sentence or two about success in completing surveys. Doesn't appear to be any consistently problematic sites, not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).
-
-**MADELEINE: Dig in to Site 1 to make sure there aren't messages hidden in 'notes' since that is the location that consists of cruise ship docking areas for a good chunk of the space.**
-
-```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=15, fig.width=18, fig.cap="Need to add caption.",warning=FALSE}
-
-print(plot_unitSuccess)
-```
-
-*Temporal* ??how define ‘season’??
-
-The annual dates of the first and last survey each year has varied by over a month (Figure \@ref(fig:Fig-SurveyTiming)).
-
-```{r EDA_SurveyTiming, eval=TRUE, cache = FALSE, include = FALSE}
-# dotplot of survey times by year
-
-source("../scripts/survey_timing_EDA.R", echo=FALSE)
-```
-
-```{r Fig-SurveyTiming1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=8, fig.width=12, fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
-
-print(plot_alldates)
-```
-
-Questions: what is cue to start of survey season?
 
@@
 
-##### pagebreak
+***
+#### Misc draft text currently removed
+_Decision Makers_
+Who are target audiences for this information?
+Who uses it or may use it?
+Local contact: C. Elaine Furbish, Skagway Bird Club, Google Site
+Predominantly historical park; limited resource staff and capacity
+Visible natural resource of interest to local community and visitors; interp and educational opportunity; topic of information requests; provides ‘community service’ given absence of any other agency with natural resource staff;
+Annual report (time lag?); National Resource Condition Assessment; other? State of the Park?
+Potential for contribution to regional, flyway, and WASO programs (e.g., CCRP & Audubon climate watch, etc.)
+Current sightings passed to local bird club & used by avian-interested visitors
+Used in quarterly regional ‘significant bird observations’, which in turn is used to develop the Alaska-wide summary for the journal North American Birds (published by the American Birding Association http://www.aba.org/nab ) and includes ‘...sightings of birds that are out of range or out of season or occur regularly in small numbers, noteworthy breeding records, unusually large or small numbers of a particular species, unusual migration dates, etc….’ (S.
+C.
+Heinl, 2017, Summary of SE AK Bird Observations, June-July 2017, accessed 2018 June 6).
+Queries to Heather Renner & Robin Corcoran re: PSG-type users, etc.
+Opportunity for community based monitoring to track projected spp impacts of climate change, change in phenology, occurrence, abundance, etc.?
+
+##### misc notes re: potential (Stakeholder, Info Needs) combos
+_Decisions_
+What types of decisions is it / will it be used for?
+(specify for each target audience/ end user)
+No specific decisions
+Current:
+KLGO - spp lists, phenology of occurrence (guidance to visitors), trends in richness (NRCA), abundance?
+Skagway bird club - spp list, current occurrences (activity re: rare & unusual)
+SE AK / Regional - spp lists; unusual sightings, etc. (spp occurrences & broad seasonal timing).
+AKR -
+WASO/NPS - ??CCRP??
+Question to Gregor & Mel (AK Audubon).
+Could be used for garnering community engagement in Climate Watch & other national efforts associated with Audubon’s Climate Report.
+Not used in continental analyses.
+“One key message that I want to emphasize is that the focus of this work was purely on climate suitability.
+What this means, as we indicate in the briefs’ “Important” box up front and “Caveats” section further back, is that significant changes in climate suitability, as measured here, will not always result in a species response, and all projections should be interpreted as potential trends.
+Multiple other factors mediate responses to climate change, including habitat availability, ecological processes and biotic interactions, dispersal capacity, species' evolutionary adaptive capacity, and phenotypic plasticity (e.g., behavioral adjustments).
+Ultimately, models can tell us where to focus our concern and which species are most likely to be affected, but monitoring is the only way to validate these projections and should inform any on-the-ground conservation action.” Gregor Schuurman, email to DENA
+
+Potential:
+KLGO: 
+(easy) - changes in occurrence (spp, season scale); changes in phenology (migration timing, by spp; breeding timing, by spp); broad (order of magnitude) changes in abundance (index); broad (?) changes in composition (by spp & timing of survey)?; species occurrence x unit (~habitat?) - refine expected lists per survey and develop targeted SOP training materials;
+(harder) - finer changes in abundance; finer changes in composition and age; ???
+
+What Objectives are the decision makers hoping to achieve by using this information?
+
+Over what spatial scales is the information being used? 
+e.g., at the park, in the community, regional scale, flyway scale, nationally, internationally
+For each spatial scale of interest, what are the associated time scales of interest for detecting changes?
+E.g., annual, w/in season, spring vs fall, etc.
+
+Q: compiling and assessing regional trends in waterbirds in SE AK?
+Asked of Heather Renner (Ak Maritime), Robin Corcoran (Kodiak NWR)
+Answers:
+Not really; regular discussion at PSG but no one has taken it on. 
+Kathy Kuletz works on statewide trends in at sea seabird observations;  Piatt GLBA work (& Martin Renner); 
+Robin C: “There's always talk at the PSG meeting about the North Pacific seabird colony database, or some variation of that database for marine birds, but it never gets anywhere.
+Basically there is no money for a database manager.
+I've been submitting my data to Rob Kaler in Migratory Bird Management, but I think that's just for storage until the money for a database manager materializes.
+More recently Rob Suryan w/NOAA has been interested - for the North Pacific Pelagic Seabird Database.
+You might want to contact him, he's leading the Gulf Watch Alaska Long-Term Ecosystem Monitoring Program, and maybe interested in this work.”
+
+NPS/Audubon:
+Wu et al.
+2018.
+Projected avifaunal responses to climate change across the U.S.
+NPS.
+PLOS ONE.
+Appendices include projected park-specific turnover, colonization, and extirpations projections - potential to test with this survey.
+Would require discussions w/ CCRP & ARO - ideally as a coordinated effort with, say, AK Audubon and ACF(?) or LCCs as tracking metric.
+Appendices include “Potential management goals and activities for parks, organized by trend group.”
+
+Melanie AK Audubon: IBA requires demonstrating higher than ‘usual’ densities (relative to surrounding area).
+
+Nicole Michel & Brooke Batemen Audubon National Science Team: of interest and possible use in national scale quantitative analyses or climate species distribution models.
+“Scaling up and integrating datasets across regions, including data from different goals, is something that we’re starting to do in other regions.
+“
+
+Occurrence, Richness
+Phenology (migration, breeding)
+Abundance - Even without the structure needed to estimate detection probability, if the same protocol is used annually this data could be used to produce long-term trend indices for the study area.
+(competing staff priorities - peak (?) usually aligns with hooligan run. Possibly reduce summer frequency?)
+
+Diagram the Conceptual Model of the system
+E.g., create an influence diagram, being sure to denote the
+Fundamental & Means Objective(s)
+Factors influencing their achievement (aka ‘system drivers’)
+Associated management or policy decisions and/or potential actions.
+Include a written summary of the literature upon which the conceptual model is based.
+
+#### Misc QUOTES for potential use later
+This protocol addresses one monitoring question with three objectives: (1) what is the annual nest density and reproductive (fledgling) success in known Hawaiian petrel colonies, (2) what are the long-term trends in colony distribution and density monitored in approximate 5-year intervals, and (3) are these affected by predator control?
+The first goal of monitoring is to obtain unbiased estimates of Hawaiian petrel nest density and reproductive (fledging) success from known colonies in HAVO and HALE in order to detect changes in colony growth or decline.
+The second goal of monitoring is to periodically (approximately every five years) obtain unbiased estimates of nest density from potential Hawaiian petrel habitat.
+This information can be used to assess changes in density and distribution of subcolonies across the landscape.
+The third goal is to estimate nest density and fledging success in areas undergoing different management regimes to assess effectiveness of management.
+Benchmark levels of these estimates could serve as warnings of the need for modified management or further investigation of these colonies.
+
+From another report but quotes from Olsen et al. (1999):
+Olsen (et al.
+1999) noted that “Most of the thought that goes into a monitoring program should occur at this preliminary planning stage.
+The objectives guide, if not completely determine the scope of inference of the study and the data collected, both of which are crucial for attaining the stated objectives.” Olsen goes on to say that a “clear and concise statement of monitoring objectives is essential to realize the necessary compromises, select appropriate locations for inclusion in the study, take relevant and meaningful measurements at these locations, and perform analyses that will provide a basis for the conclusions necessary for meeting the stated objectives.”
+
+Fancy, Gross, and Carter (2009):
+The development of monitoring objectives, which provide additional focus about the purpose or desired outcome of the monitoring effort, was an iterative process that sometimes required several years to refine.
+Early in the design process, monitoring objectives were stated in more general terms, such as “Determine trends in the incidence of disease and infestation in selected plant communities and populations”, whereas the final monitoring plan and protocols provided monitoring objectives that met the test of being realistic, specific, and measurable (e.g., “Estimate trends in the proportion, severity, and survivorship of limber pine trees infected with white pine blister rust at Craters of the Moon National Monument”; Garrett et al.
+2007).
+
+Structure from Northeast Coastal and Barrier Network
+https://irma.nps.gov/DataStore/Reference/Profile/2195238
+Justification => Vital Sign => Monitoring Goal => Monitoring Questions => Monitoring Objectives => Measures (eg pg 55 & 56)
+
+Clarify Objectives of the Decision Makers 
+Among the decision maker objectives identified in 1.b.ii, clarify 
+the Fundamental Objective(s) from 
+the Means Objectives, then
+diagram the Objectives Hierarchy.
+Identify the Attributes that need to be monitored to assess achievement of each of the objectives. (‘Concepts by Theory’, Ford 2001)
+Identify the Measurements selected to characterize each attribute. (‘Concepts by Data’, Ford 2001) [WHAT]
+Identify any Threshold values associated w/ each Measurement
+
+# To Do
+* Issues to Resolve with content (narrative, figures, analyses)
+  + Framework - review & incorporate relevant from I&M national program guidance for frequency of analyses & reporting, etc., etc.
+  + Winnow out target species that do not occurr w/ adequate frequency to report on.
+E.g., those only seen <3 or 5 years???
+Look at phenology graphs and determine threshold to apply.
+Consider some species N-ward and S-ward migration routes may differ.
+  + finish script uploading AOU data csv object & using to order Target Species
+  + What is up with 2008 surveys? why such distinct timing?
+  + Think about Covariate EDA, if any...change in observers?
+  + Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
+  + clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.
+  + covariate distribution graphs (lots of univariate & bivariate things; around line 534);
+  +plot of total # spp sighted each season across years (around line 600)
+
+# Issues to Resolve with Rmarkdown, Knitr, Github, etc.
+- conversion from paper.md to paper.pdf (pandoc issue?) 
+- adding files to .gitignore
 

### 2019-08-12 analysis/scripts/survey_duration.R

@@ -4,14 +4,14 @@
 # Libraries used: ggplot2, dplyr
 ########
 # Madeleine Ward (2019 Mar 17); Joel Reynolds (2019 Apr 11)
-# Last Edit: 2019 Mar 17
+# Last Edit: 2019 Mar 17; 2019 Aug 12 moved creation of survey_duration to target_species.R
 ########
 
-survey_duration <- target_data %>%
-  #find sum of all observations for all target species, on each date
-  group_by(GIS_Location_ID, Date) %>%
-  mutate (total_obs_site = sum(Num_Obs)) %>%
-  ungroup()
+# survey_duration <- target_data %>%
+#   #find sum of all observations for all target species, on each date
+#   group_by(GIS_Location_ID, Date) %>%
+#   mutate (total_obs_site = sum(Num_Obs)) %>%
+#   ungroup()
 
 # group by year to get separate boxplot for each year
 plot_duration <-

### 2019-08-12 analysis/scripts/target_species.R

@@ -3,7 +3,7 @@
 # Libraries used: dplyr
 ########
 # Madeleine Ward
-# Last Edit: 2019 Feb 23
+# Last Edit: 2019 Feb 23, 2019 Aug 12 added survey_duration creation
 ########
 
 #create vector of target species types
@@ -14,8 +14,17 @@ target_data <- species_data %>%
   #filter for species in targetList
   filter(str_detect(str_to_lower(Common_Name, locale = "en"), paste(targetList, collapse="|")))
 
+survey_duration <- target_data %>%
+  #find sum of all observations for all target species, on each date
+  group_by(GIS_Location_ID, Date) %>%
+  mutate (total_obs_site = sum(Num_Obs)) %>%
+  ungroup()
+
+
 # Need to define the species list as an ordered factor following the AOU sequence.
 # Joel will do this.
+
+
 
 # 4/11/19 - JR - Question:
 # is subsetting on Common_Name getting everything we want?

### 2019-08-12 analysis/paper/paper.Rmd
edits to catch up post Madeleine, prep for KLGO call.

--- 2019-05-24 analysis/paper/paper.Rmd
@@
 ```
-# NOTES To Madeleine
-- see edits in survey_duration.R to aid interpretation by the viewer (e.g., changes in x axis tick locations, etc.) $\checkmark$
-- documentation in scripts $\checkmark$
-- **NEXT Priority Milestones**: note - all references to line numbers are prior to any other edits!
-- finish graphs of first & last observation per year for each target species (around line 386);**ultimately** add a fitted linear trend line to each of those (we need to account for frequency of surveying, so some subtle effort things here, but can resolve that later); $\checkmark$
-- graphs of species occurrence (around line 407) for each of target species; $\checkmark$
-- covariate distribution graphs (lots of univariate & bivariate things; around line 534);
-- plot of total # spp sighted each season across years (around line 600)
-
-# Issues to Resolve with Rmarkdown, Knitr, Github, etc.
-- conversion from paper.md to paper.pdf (pandoc issue?) 
-- adding files to .gitignore
-
-# Issues to Resolve with content (narrative, figures, analyses)
-- **Joel & Madeleine** will discuss survey_dates.R commands and how to modify them so that we get a sense of when surveys were & when target species was first (or last) seen each year.
-E.g., separate out any filtering process stemming from survey effort levels and timing.
-(thinking some thing like showing open circles for survey events PRIOR to first sighting...)
-- Need to think about way to winnow out target species that do not occurr w/ adequate frequency to report on.
-E.g., those only seen <3 or 5 years???
-Look at phenology graphs and determine threshold to apply.
-Consider some species N-ward and S-ward migration routes may differ.
-- Joel will finish script uploading AOU data csv object & using to order Target Species
-- What is up with 2008 surveys? why such distinct timing?
-- Think about Covariate EDA, if any...change in observers?
-- Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
-- clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.
 
@@
 
-##### pagebreak
+***
+# To Do
+## Issues to Resolve with content (narrative, figures, analyses)
+- Winnow out target species that do not occurr w/ adequate frequency to report on.
+E.g., those only seen <3 or 5 years???
+Look at phenology graphs and determine threshold to apply.
+Consider some species N-ward and S-ward migration routes may differ.
+- finish script uploading AOU data csv object & using to order Target Species
+- What is up with 2008 surveys? why such distinct timing?
+- Think about Covariate EDA, if any...change in observers?
+- Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
+- clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.
+- covariate distribution graphs (lots of univariate & bivariate things; around line 534);
+- plot of total # spp sighted each season across years (around line 600)
+
+# Issues to Resolve with Rmarkdown, Knitr, Github, etc.
+- conversion from paper.md to paper.pdf (pandoc issue?) 
+- adding files to .gitignore
 

### 2020-02-06 analysis/paper/paper.Rmd

--- 2019-08-12 analysis/paper/paper.Rmd
@@
 I've also included some of the basic survey effort data visualizations that we've already developed but haven't spent time on the associated narrative.
+
+# NOTES To Madeleine
+- covariate distribution graphs (lots of univariate & bivariate things; around line 534);
+- plot of total # spp sighted each season across years (around line 600)
+
+# Issues to Resolve with Rmarkdown, Knitr, Github, etc.
+- conversion from paper.md to paper.pdf (pandoc issue?) 
+- adding files to .gitignore
+
+# Issues to Resolve with content (narrative, figures, analyses)
+- **Joel & Madeleine** will discuss survey_dates.R commands and how to modify them so that we get a sense of when surveys were & when target species was first (or last) seen each year.
+E.g., separate out any filtering process stemming from survey effort levels and timing.
+(thinking some thing like showing open circles for survey events PRIOR to first sighting...)
+- Need to think about way to winnow out target species that do not occurr w/ adequate frequency to report on.
+E.g., those only seen <3 or 5 years???
+Look at phenology graphs and determine threshold to apply.
+Consider some species N-ward and S-ward migration routes may differ.
+- Joel will finish script uploading AOU data csv object & using to order Target Species
+- What is up with 2008 surveys? why such distinct timing?
+- Think about Covariate EDA, if any...change in observers?
+- Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
+- clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.
 
@@
 
+##### pagebreak
+
 ### Colophon

### 2020-03-19 analysis/paper/paper.Rmd
Edits to intro, weekly frequency plots

--- 2019-08-12 analysis/paper/paper.Rmd
@@
   fig.path = "../figures/",
+  fig.width = 6,
+#  out.width = "75%", crashes when try to add any of these as defaults & knitr to Word docx
+#  fig.align = "center",
+#  fig.asp = 0.618,
   error = FALSE
@@
   + 2019 Summer (minimal time) - prep for discussion w/ KLGO Resources staff.
+  + 2020 Feb/Mar (as time permitted) -
 
@@
 I've also included some of the basic survey effort data visualizations that we've already developed but haven't spent time on the associated narrative.
+
+* Issues to Resolve with content (narrative, figures, analyses)
+  + **Joel** needs to add commands to target_species.R to define spp & Common Name as **ordinal** variables following AOU sequencing (e.g., so all ducks show up together, all gulls show up together, etc.), so that spp panels are ordered 'as expected' in any figures.
+  + **Joel & Madeleine** will discuss survey_dates.R commands and how to modify them so that we get a sense of when surveys were & when target species was first (or last) seen each year.
+E.g., separate out any filtering process stemming from survey effort levels and timing.
+(thinking some thing like showing open circles for survey events PRIOR to first sighting...)
+  + Need to think about way to winnow out target species that do not occurr w/ adequate frequency to report on.
+E.g., those only seen <3 or 5 years???
+Look at phenology graphs and determine threshold to apply.
+Consider some species N-ward and S-ward migration routes may differ.
+  + Joel will finish script uploading AOU data csv object & using to order Target Species
+  + What is up with 2008 surveys? why such distinct timing?
+  + Think about Covariate EDA, if any...change in observers?
+  + Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
+  + clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.
+  + covariate distribution graphs (lots of univariate & bivariate things; around line 534);
+  + plot of total # spp sighted each season across years (around line 600)
 
@@
 Assessing a monitoring program focuses on four broad topics, each associated with a major development phase of the monitoring program (ibid):
+
 1. _Why_ - [Problem Framing & Information Objectives] Are the management objective(s) motivating the monitoring well articulated? Do they remain relevant and important?
+
 2.
@@
 If so, can the data collection be made more efficient, reducing the observation burden on staff resources?
+
 3.
@@
 Are more informative analyses available given interim improvements in methods?
+
 4.
@@
 
-# Background
+(add overview of key findings?)
 
@@
 
-## Phase 2 - Monitoring & Analysis Design
-**NOTE TO READER**: The rest of this is a mix of outline and notes, with some initial data visualizations. This is far from a draft narrative. **Proceed with Caution**
-
-HOW monitor? Clarify the various design choices.
-
-Clarify rationale for monitoring versus some other means of addressing the information objectives (e.g., literature review, controlled experiment, etc.), and type of monitoring being undertaken (status & trends or surveillance, threshold, effectiveness, or adaptive management).
-
-Questions: Why? What are information goals & objectives?
-What sort of observations would trigger action? What actions?
-
-### Survey Design
-
-*Target Frame* - space, time, scale, target populations
-
-*Spatial* Upper Lynn Canal visible from coastline from Skagway to Dyea, including Skagway & Taiya river mouths and Nahku Bay (between Skagway and Dyea Point)
-
-> INSERT FIGURE FROM NRSS report
-
 ```{r Data-Retrieval, eval=TRUE, echo=FALSE, warning = FALSE}
@@
 
-> **Joel** needs to add commands to target_species.R to define spp & Common Name as **ordinal** variables following AOU sequencing (e.g., so all ducks show up together, all gulls show up together, etc.).
+## Phase 2 - Monitoring & Analysis Design
+**NOTE TO READER**: The rest of this is work in progress and consists of a mix of outline and notes, with some initial data visualizations.
+This is far from a draft narrative.
+**Proceed with Caution.**
 
-```{r Target_Species, eval=TRUE, echo=FALSE, warning = TRUE}
-# selects out only target species listed above
-#
-source("../scripts/target_species.R")
+### Survey Design
+
+*Spatial Component*
+
+IDEAL: The Upper Lynn Canal, including Skagway & Taiya river mouths and Nahku Bay (between Skagway and Dyea Point).
+
+REALIZED: The portions of Upper Lynn Canal visible in the eight census units defined based on standardized observation points established at public access points between Skagway and Dyea, including Skagway & Taiya river mouths and Nahku Bay [@Hahr2004].
+The standardized observation points were established for each census unit to ensure the most complete coverage of the study area, and to promote consistency of area surveyed across personnel changes (ibid).
+
+> INSERT FIGURE FROM NRSS report, include survey points and denote visible portions of the area actually visible from each survey point.
+
+```{r SiteSurveyCompletion, eval=TRUE, cache = FALSE, include = FALSE}
+# dotplot of survey occurrences by unit and whether there were observations
+
+source("../scripts/survey_sites.R", echo=FALSE)
 ```
 
-*Temporal* 
+```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.asp=1.3, fig.cap="Each survey in a given year (row) is shown as a circle on the Julian date of the survey.
+Filled circles denote surveys where at least one of the target species of waterbirds was observed (e.g., one of the 82 species denoted as 'Waterbird' in the data field 'Category').",warning=FALSE, out.height="60%"}
+
+print(plot_unitSuccess)
+```
+
+* **Considerations**:
+  + At this broad level, there don't appear to be any sites at which surveys are consistently unable to be completed (using the flag of detecting any of the target species as a signal of successful survey), not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).
+A more restricted list of target species might change this conclusion.
+  + Over the course of decades the regions visible from the standardized observation points may change due to vegetation growth, new structures on private lands, etc.
+Similarly, seasonal patterns of vegetation growth may constrain the visible region as the season progresses at some standardized observation points.
+This is a minor point but should be considered in analyses focusing on those time scales.
+If it isn't part of the SOP already, provide a map clearly depicting the intended survey area as a guide for field staff to promote consistency of area observed across changes in personnel.
+  + (Design point to ponder further) Potential confounding of sites or route sequencing and tidal influence?
+The SOP says to survey units in the sequence 1, 2, 3, 4, 5, 6, 8, 7, or the reverse, depending on the timing of tide (due to impact on census unit 7).
+  + Other?
+
+*Temporal Component: Season initiation and end*
+
+IDEAL: (season initiation & duration) Apparently (this should be confirmed) from the initiation of northward migration by the earliest of the target species that pass through the study area to the conclusion of southward migration by the latest of the target species that pass through the study area.
+
+REALIZED: [@Hahr2004]: "Census units were [scheduled to be] surveyed once per week in late April and May and bi-weekly during late summer and fall (August – mid September).
+Surveys were conducted as weather and scheduling permitted...
+Surveys were only conducted under optimal weather conditions; however, occasionally conditions deteriorated during the course of a survey making species identification difficult (especially in the case of low clouds, fog and wind)."
+
+* **Considerations**:
+  + The annual dates of the first and last survey each year has varied by over a month (Figure \@ref(fig:Fig-SiteSurveyCompletion)), even ignoring late 2007 and early 2008.
+  + Ideally, there should be clear description of the cue used to prompt the start of a survey season and, importantly, the information goal that underlies that decision.
+This is increasingly important as the potential impact of climate change on migration phenology increases.
+  + The period for 'once per week' surveys is intended to capture migration (based on reports).
+If the goal is to assess potential changes in migration timing, should assess whether it is feasible to survey with an adequate frequency *prior* to onset of migration for the target species so as to be sure to capture both the beginning and the end of the northward / southward migration periods.
+This may require a greater frequency of surveying than is feasible to maintain.
+Even if capturing phenology changes is of secondary importance, those changes will impact when surveys should be conducted (and the intensity at which they are conducted) even if the information goal is to assess frequency of occurrence or the even more challenging goal of abundance.
+  + the information goal isn't to assess potential changes in phenology for target species period of 'Ideally, the reasoning for the different survey frequencies ('once per week' and 'bi-weekly' periods) should be directly tied to explicit information goals of the survey and the historic data used to assess the feasibility of actually meeting those goals.
+It may be found that less frequent surveys will meet the same information goals w/ acceptable levels of precision.
+  + consider graphical summary assessing consistency in attaining weekly targets defined above.
+
+```{r SurveyWeeklyFrequency, eval=TRUE, cache = FALSE, include = FALSE}
+# dotplot of survey frequency each week
+
+source("../scripts/survey_sitesWeekly.R", echo=FALSE)
+```
+
+```{r Fig-SurveyWeeklyFrequency, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.asp=1.2, fig.cap="Each week with any surveys in a given year (row) is shown as a circle on the week of the survey.
+The size and fill intensity of the circles denote the number of surveys that week (bigger circle and lighter fill denote more surveys).
+Vertical lines denote, approximately, the beginning of April, May, June (black line), July, August (black line), and September.",warning=FALSE, out.height="60%"}
+
+print(plot_WeeklyFrequency)
+```
+
+*Survey Timing and Changes in Phenology*
+Want to see if the survey design's start and end dates each season need to be modified to account for any observed changes in phenology (due to climate change, really).
+Ultimately need to review Audubon's Climate Change Watch List and see if any of their target species are observed in this effort.
+
+> MADELEINE: FIRST graph - x axis is year, y axis is Day of Season, curve for each of the target species defined in next sentence; possibly consider facet by site if anything of interest is revealed when doing so.
+Curve should be made of points of the first day the species was observed each year; maybe do one with just raw points connected by line segments and one with some sort of smooth - use default until we get a chance to look at these and talk about smoothers.
+Different color for each of the target species.
+Might be too many curves (species), in which case define some grouping on species or facet by species (more important than site).
+
+```{r SurveyDates, eval=TRUE, cache = FALSE, include = FALSE}
+
+# scatterplot of species' first and last sightings compared to survey season start
+source("../scripts/survey_dates.R", echo=FALSE)
+```
+
+```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The first sighting of each species each year.",fig.height=20, fig.width=12, warning=FALSE}
+
+print(plot_timing_1)
+```
+
+(migration) 1 per wk, Apr 27 - May 20; (breeding) 1 per 2 wks, Jun 6 - Jul 18; (migration) 1 per wk, Aug 1 - Sept 28.
+2016: 17 surveys total
+ Questions: How define beginning & end of breeding period?
+Info goals from migration periods? From breeding periods? 
+Design options to reduce frequency?
+
+Diagnostic analysis=> 
+.assessment of impact on priority info goals of different frequencies?
+.spp specific occurrency by survey date across years - capturing onset and end of migration?
+.other assessments of survey timing relative to phenology of spp
+.mine FB, Eagle Chat, others for date of first occurrences (see 2016 report); Elaine. Requires clarifying information goals of survey vrs ‘first occurrence’. E.g., not tail behavior.
+
+*Diurnal Temporal Component*
+(diurnal timing) Determined by tidal cycle at census unit 7 in relationship to daylight period.
+
+REALIZED: ([@Hahr2004], pg 13): "Surveys were conducted as weather and scheduling permitted and dates/time were not selected at random."
+
+*Observation Process*
+[@Hahr2004], pg 13: Observers counted and identified all birds visible within each unit and
+documented birds opportunistically observed while moving between units.
+Counts were made by scanning the census units with Bushnell Legend 10 x 42
+binoculars and a Bausch & Lomb 20-60 x 70 spotting scope until the observer
+was able to quantify all visible birds in the unit.
+
+Data were recorded on standardized data
+forms developed after Collins et al. (2001) (Appendix B). ArcView shapefiles of
+the observation points and census units were created for use in the park’s GIS.
+All data, original data forms, and field notes were archived at KLGO.
+
+**Comments**
+
+*  Detection issues
+*  Covariates impacting detection; spp-specific rates, implicit estimation via BHM?
+
 ```{r EDA_SurveyTiming, eval=FALSE, cache = FALSE, include = FALSE}
@@
 ```
-The annual dates of the first and last survey each year has varied by over a month (Figure \@ref(fig:Fig-SurveyTiming)).
-
-Questions: what is cue to start of survey season?
-
-*Survey Completion*
-
-```{r SiteSurveyCompletion, eval=TRUE, cache = FALSE, include = FALSE}
-# dotplot of survey occurrences by unit and whether there were observations
-
-source("../scripts/survey_sites.R", echo=FALSE)
-```
-
-```{r Fig-SiteSurveyCompletion, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis',fig.height=15, fig.width=18, fig.cap="Need to add caption.",warning=FALSE}
-
-print(plot_unitSuccess)
-```
-Add sentence or two about success in completing surveys. Doesn't appear to be any consistently problematic sites, not even site 1 (Figure \@ref(fig:Fig-SiteSurveyCompletion)).
 
 *Survey Effort*
+
 The time spent at each site during a survey has varied widely across the years, especially at Sites 1, 7, and 8 (Figure \@ref(fig:Fig-SurveyEffort)).
@@
 
-```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.height=12, fig.width=12, fig.cap="",warning= FALSE}
+```{r Fig-SurveyEffort, echo=FALSE, eval=FALSE, cache=FALSE, fig.show= 'asis', fig.asp=0.618, fig.cap="The amount of time spent at each site each survey varied widely across each season, each year, and across sites.",warning= FALSE}
 
@@
 
-[@Hahr2004], pg 13: Areas of coastline that were visible from public access points
-between Skagway and the Taiya River delta were divided into eight census units
-and surveyed once per week in late April and May and bi-weekly during late
-summer and fall (August – mid September) (Figure 5). Surveys were conducted
-as weather and scheduling permitted and dates/time were not selected at
-random. Standardized observation points were established for each census unit
-to ensure the most complete coverage of the study area, and to minimize
-observer bias.
-
-Observers counted and identified all birds visible within each unit and
-documented birds opportunistically observed while moving between units.
-Counts were made by scanning the census units with Bushnell Legend 10 x 42
-binoculars and a Bausch & Lomb 20-60 x 70 spotting scope until the observer
-was able to quantify all visible birds in the unit. Surveys were only conducted
-under optimal weather conditions; however, occasionally conditions deteriorated
-during the course of a survey making species identification difficult (especially in
-the case of low clouds, fog and wind). Data were recorded on standardized data
-forms developed after Collins et al. (2001) (Appendix B). ArcView shapefiles of
-the observation points and census units were created for use in the park’s GIS.
-All data, original data forms, and field notes were archived at KLGO.
-
-*Survey Timing and Changes in Phenology*
-Want to see if the survey design's start and end dates each season need to be modified to account for any observed changes in phenology (due to climate change, really).
-Ultimately need to review Audubon's Climate Change Watch List and see if any of their target species are observed in this effort.
-
-> MADELEINE: FIRST graph - x axis is year, y axis is Day of Season, curve for each of the target species defined in next sentence; possibly consider facet by site if anything of interest is revealed when doing so.
-Curve should be made of points of the first day the species was observed each year; maybe do one with just raw points connected by line segments and one with some sort of smooth - use default until we get a chance to look at these and talk about smoothers.
-Different color for each of the target species.
-Might be too many curves (species), in which case define some grouping on species or facet by species (more important than site).
-
-```{r SurveyDates, eval=TRUE, cache = FALSE, include = FALSE}
-
-# scatterplot of species' first and last sightings compared to survey season start
-source("../scripts/survey_dates.R", echo=FALSE)
-```
-
-```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The first sighting of each species each year.",fig.height=20, fig.width=12, warning=FALSE}
-
-print(plot_timing_1)
-```
-
 > SECOND graph - similar but showing last day of the season the species was observed each year.
@@
 
-*Sample Selection process* - any strata or clusters? Two-stage processes? Probabilistic selection? [get in to enumerative vs analytical distinction of Deming?]
+*Sample Selection process* - 
 Currently survey all of frame (‘census’ of units); no spatial sampling uncertainty. 
@@
 Diagnostic analysis => any ‘representative’ units? How would that be defined?
-
-*Survey timing* 
-Per season -
-
-Questions: how define when to begin surveys? Zeros? Triggering events?
-Potential confounding factors? (changes in phenology, ClCh); important objective or confounder? Design options for addressing?
-Diagnostic analysis=>  Signs of confounding?
-
-W/in season - (e.g., LEAU St. George); potential refinement?
-Temporal: 
-(migration) 1 per wk, Apr 27 - May 20; (breeding) 1 per 2 wks, Jun 6 - Jul 18; (migration) 1 per wk, Aug 1 - Sept 28.
-2016: 17 surveys total
- Questions: How define beginning & end of breeding period?
-Info goals from migration periods? From breeding periods? 
-Design options to reduce frequency?
-Diagnostic analysis=> 
-.assessment of impact on priority info goals of different frequencies?
-.spp specific occurrency by survey date across years - capturing onset and end of migration?
-.other assessments of survey timing relative to phenology of spp
-.mine FB, Eagle Chat, others for date of first occurrences (see 2016 report); Elaine. Requires clarifying information goals of survey vrs ‘first occurrence’. E.g., not tail behavior.
-
-Per survey
-route sequencing and tidal influence; 
-SOP: 1,2,3,4,5,6,8,7 or reverse, depending on tide timing (determines when do unit 7).
 

### 2020-03-19 analysis/scripts/survey_sitesWeekly.R

@@ -14,19 +14,23 @@ blah<- species_data %>%
   group_by(Year) %>% count(Week)
 
 plot_WeeklyFrequency <- ggplot() +
+  # add vertical lines for April,May, June, July, Aug, Sept
+  # actual dates used: Monday 2 April, 2007, May (30 April), June (black line at 28 May),
+  # July (2 July), and Aug (black line at 30 July), and September (3 Sept)
+  geom_vline(xintercept=c(14,18,22,27,31,36),colour="grey") +
+  geom_vline(xintercept=c(22,31), colour="black") +
   # add layer with circles for all weeks & scale & transparency a function of
   # number of surveys w/in that week
   geom_point(data = blah,
              #plot with open circle shape & set size = freq of surveys per week
              aes(x = Week, y = Year, size=n, alpha=1/n), col="black", shape = 16)+
-  # add vertical lines for April,May, June, July, Aug, Sept
-  geom_vline(xintercept=c(14,18,22,27,31,36),colour="grey") +
   labs(x = "Julian Date",# \n(vertical lines: Monday April 2, April 30, May 28, July 2, July 30, and Sept 3, 2007)",
        title="") +
   #make background white
   theme_bw() +
   theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank()) +
   guides(alpha=FALSE)
+
 
 # week 14 = April 2, 2007
 # Week 18 April 30

### 2020-03-20 analysis/paper/paper.Rmd
Adding high-level take home thoughts for guiding KLGO staff discussions while project is on hold.

@@
 INTERNSHIP ENDED
-  + 2019 Summer (minimal time) - prep for discussion w/ KLGO Resources staff.
+  + 2019 Summer (minimal time) - prep for August discussion w/ KLGO Resources staff (Jason Taylor, Anne Matsov, Mary Hake); discussed major concern over lack of explicit, specific motivating Objectives (see **Introduction** and **Assessment Results: Phase 1 - Problem Framing and Information Objectives**).
+  + 2019 Fall (no progress) - all efforts on hold during Regional Biometrician's detail as NRST Lead
+  + 2020 Winter (high level summary of initial concerns added) - in light of the forthcoming transition of the Regional Biometrician to another NPS program, there was a follow up discussion (late Feb) with KLGO Resource Staff (Anne Matsov, Mary Hake) touching on a variety of initial concerns (see **Summary of (Potential) Concerns** below) identified but not yet detailed or assessed quantitatively and thus not yet detailed in the incomplete draft report; conversation ultimately focused on implications of those concerns regarding decisions of (a) potentially reducing or (b) potentially skipping the FY20 field season.
+Regional Biometrician pointed out that reducing the data collection effort requires further thought and familiarity with existing program's limitations and KLGO's 'hierarchy of objectives' for undertaking this monitoring effort and that, from a statistical perspective, skipping one year in an 15+ year monitoring effort was not problematic in terms of summary analyses of trends and changes.
   + 2020 Feb/Mar (as time permitted) -
 
-The following excerpt from the planned report lists 4 major topics or phases that should be addressed in a full assessment of a monitoring program.
-The most important is the first topic - the *Why* motivating the program; e.g., the program's Problem Framing and Information Objectives.
-I've provided draft feedback on this and a number of key questions that require resolution by KLGO staff prior to diving into any detailed analyses *so as to ensure futher detailed review of data collection & analyses target KLGO's desired Information Objectives*.
+# Summary of (Potential) Concerns
+The report was planned to focus on each of 4 major assessment areas (see **Introduction** below).
+The current draft is near complete only with regard to the first, focused on problem Framing.
+The other points immediately below are other priority concerns raised in developing this project and of importance for the Park and the next Regional Biometrician to be aware of but that have not yet been captured in the draft text or figures.
+They should be considered by the Park in deciding on Next Steps.
+
+## Unclear Problem Framing and Motivating Objectives
+Of the 4 major topics or phases that should be addressed in a full assessment of a monitoring program, the most important is the first topic - the *Why* motivating the program; e.g., the program's Problem Framing and Information Objectives.
+I've provided draft feedback on this and a number of key questions that require resolution by KLGO staff prior to diving into any detailed analyses so as to ensure futher detailed review of data collection & analyses target KLGO's desired Information Objectives.
+
+## Imporant aspects of the Survey Design 
+**Timing**: the start and end dates of the survey vary across years by over a month (see draft Figure 2 below).
+The timing should be clarified and consistently implemented, as much as possible, to reduce this source of uncertainty and problems in analyzing changes in phenology and other outcomes of interest.
+The clarification should be targeted as the priority information goals the Park has identified in the Problem Framing.
+
+**Timing re: Target Species Phenology**: if migration phenology is a priority outcome of interest, the simple analyses related to the draft Figures 3 and 5 should be pursued to ID those target species whose northward and/or southward passages are regularly 'captured'.
+Note that if species is already observed on the first (last) survey of the season, then the survey is **not** capturing the timing of their northward (southward) migration.
+To capture that event requires at least one survey suggesting the species is not yet arrived in the area *prior* to the first observation (same thing for southward: at least one survey w/o observation *after* the last observation of the species).
+This hasn't been thoroughly investigated, but glancing at draft Figure 3 (or 5) suggests that the current survey design, as implemented, is capturing the northward (southward) migration of very few species; e.g., most of the species are seen on the first (last) survey of the season.
+If Phenology is an important outcome of interest, these simple assessments should be pursued to clarify which, if any, regularly seen species this survey may be providing consistent information on.
+
+**Frequency**: the stated intention is for survey frequency to vary in 'early & mid' versus 'late' season.
+Quick superficial review suggests the target intensity changes have not been achieved consistently and that, in many years, surveys have been conducted much more frequently than targeted (see draft Figure 2 below).
+While 'greater frequency' doesn't really create problems for the analysis, it may be something the NR program lead wants to explore further to be aware of a need to constrain effort in order to have more capacity to attend to other projects.
+The variation in frequency across years needs to be considered and potentially accounted for in analyses of outcomes of interest.
+The motivation for the variation in frequency needs to be clarified and then (relatively simple) assessments completed to identify *those species* for which the motivating information objectives are being achieved.
+E.g., is the breeding status of the target species being reqularly measured with adequate frequency?
+
+**Duration**: the time spent at each site varies widely across visits (see draft Figure 4).
+This needs to be accounted for in any analysis of 'frequency of occurrence' or 'apparent abundance' following an approach akin to 'catch per unit effort' in fisheries.
+E.g., duration becomes a covariate in all models.
+Consider implications for clarifying SOP.
+Consider implications for different priority outcomes of interest - e.g., does it need to be variable and up to the observer or are there other rule sets that lead to more consistent implementation?
+
+## Measurement Process Concerns
+**Detection**: the current survey design does not account for the need to account for (by estimating) the impact of changing rates of detection.
+Detection rates will differ among species, among observers, and among observation conditions.
+It might be possible to implicitly account for detection in a Bayesian analysis framework that  accounted for a parsimonious set of covariates related to survey condition, date or season of survey, and/or survey duration; **this would not be an insubstantial undertaking**, so whether it was worth doing would depend on the information objectives and goals.
+Ultimately, detection must be addressed in any future survey SOP, but the importance and approach will depend on the Park clarifying the priority outcomes of interest (e.g., information needs).
+Detection is less of an issue for occurrence (and, perhaps, breeding state) than for phenology than  for questions related to abundance (for which it is can be of central importance).
+
+**Phenology**: if phenology is a priority outcome of interest, consider analyses treating the response as interval censored data.
+
+**Covariates**: the draft assessment didn't get to the point of exploring the potential value of the many covariates collected.
+In general, their value will differ with different outcomes of interest.
+On first blush, many of the most intensively measured covariates may not be of great value - e.g., a 'snapshot in time' covariate measured with the goal of accounting for variation in detection may not be a very useful characterization of conditions that change over the course of a site observation period whose duration was 30 or more minutes.
+
+The choice of covariates should be determined based on the priority outcomes of interest and the planned analyses to address the Park's priority information needs.
+E.g., all of those components should be figured out, then consideration given to important and **feasible-to-collect** covariates, then revisit the planned analyses, etc...
+My gut take is that much energy is being devoted to collecting data that, ultimately, won't be useful.
+See also, related comments from Elaine Furbish (summer 2019, relayed by Mary Hake).
+
+* **Taxa for which it is feasible to learn something**:  An important next step is to review the frequency of detection of each taxa, both in terms of the various phases of the season, and across years, to figure out:
+  + which are detected with adequate frequency to say anything reliably about their occurrence
+  + are potential candidates for being monitoring for northward migratory phenology; ...for southward migratory phenology;
+  + are potential candidates for being monitored for seasonal local abundance and/or breeding
+  + and other possible categories based on whatever outcomes of interest the Park identifies.
+
+Note that synthesizing that information is a valuable outcome from the monitoring program and is a prime example of 'Step 10' of the 'Road Map for Monitoring': *Learning how to Learn* and an essential step before figuring out the survey effort levels required to meet the Parks information objectives (e.g., by conducting power analyses, etc.)
+
+## Reducing Data Collection vs skipping a field season
+An important immediate Park concern regards the tradeoffs between reducing current survey effort vs skipping a field season.
+This was discussed on the conference call in Mar 2020.
+In general, rather than arbitrarily modify the existing survey, it seems fine (from a statistical perspective) to skip a field season if limited staff capacity has to be redirected elsewhere.
+In the interim, the Park should clarify their motivating information needs and, ideally, this assessment advanced either by the incoming Regional Biometrician, via contract with an experienced applied statistician familiar with issues in natural resource monitoring, or perhaps by a student under the mentorship of such a statistician.
+Relatedly, the Park should be aware that many natural resource monitoring programs have shifted to regularly scheduled but intermittent monitoring - e.g., every other year or every third year or (there are many possible combinations) 2 years of monitoring followed by 2 years of no monitoring, etc.
+The analyses are not substantially more complicated and the cost and capacity savings can be quite impactful.
+However, all such decisions, again, really depend on the Park's motivating information objectives.
+
 I've also included some of the basic survey effort data visualizations that we've already developed but haven't spent time on the associated narrative.
@@
 * Issues to Resolve with content (narrative, figures, analyses)
-  + **Joel** needs to add commands to target_species.R to define spp & Common Name as **ordinal** variables following AOU sequencing (e.g., so all ducks show up together, all gulls show up together, etc.), so that spp panels are ordered 'as expected' in any figures.
-  + **Joel & Madeleine** will discuss survey_dates.R commands and how to modify them so that we get a sense of when surveys were & when target species was first (or last) seen each year.
+  + Use the TargetList index vectors in the 'data_wrangling.R' script to redo all the figures with species-specific panels so that they are sets of four figures, each of which should fit on a single page.
+This should order the panels by AOU species number and thus improve interpretation and use by avian biologists.
+  + decide on a 'frequency of occurrence' rule to winnow out the many species that don't appear with enough consistency to warrant wasting plotting space on their inclusion.
+In developing the rule, consider the implications of Audobon's Climate Watch List predictions for shifting species ranges and shifts in phenology.
+  + Modify survey_dates.R script to get a sense of when surveys were & when target species was first (or last) seen each year.
 E.g., separate out any filtering process stemming from survey effort levels and timing.
 (thinking some thing like showing open circles for survey events PRIOR to first sighting...)
-  + Need to think about way to winnow out target species that do not occurr w/ adequate frequency to report on.
-E.g., those only seen <3 or 5 years???
-Look at phenology graphs and determine threshold to apply.
-Consider some species N-ward and S-ward migration routes may differ.
-  + Joel will finish script uploading AOU data csv object & using to order Target Species
+  + Consider some species N-ward and S-ward migration routes may differ.
+  + clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.
   + What is up with 2008 surveys? why such distinct timing?
-  + Think about Covariate EDA, if any...change in observers?
   + Begin thinking about (Freq of Occurrence, Phenology, Abundance | target species) and then Richness - both Phenology (Day of Season) and across years.
-  + clean up structure of survey design section - consistent pattern in {Element Definition, Settings, Assessment}; as cleanly as possible separate space and time aspects of survey design (& effort) from spp-specific responses.
   + covariate distribution graphs (lots of univariate & bivariate things; around line 534);
@@
 The size and fill intensity of the circles denote the number of surveys that week (bigger circle and lighter fill denote more surveys).
-Vertical lines denote, approximately, the beginning of April, May, June (black line), July, August (black line), and September.",warning=FALSE, out.height="60%"}
+Vertical lines denote, approximately, the beginning of April, May, June (black line), July, August (black line), and September.
+In many years the survey intensity has been much more than the stated targeted frequency of 1/wk in the early season and 1/every other week in late season.",warning=FALSE, out.height="60%"}
 
@@
 
-> MADELEINE: FIRST graph - x axis is year, y axis is Day of Season, curve for each of the target species defined in next sentence; possibly consider facet by site if anything of interest is revealed when doing so.
-Curve should be made of points of the first day the species was observed each year; maybe do one with just raw points connected by line segments and one with some sort of smooth - use default until we get a chance to look at these and talk about smoothers.
-Different color for each of the target species.
-Might be too many curves (species), in which case define some grouping on species or facet by species (more important than site).
-
 ```{r SurveyDates, eval=TRUE, cache = FALSE, include = FALSE}
@@
 
-```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.cap="The first sighting of each species each year.",fig.height=20, fig.width=12, warning=FALSE}
-
+```{r Fig-SurveyDates1, echo=FALSE, eval=TRUE, cache=FALSE, fig.height=20,fig.width=12, fig.show= 'asis', fig.cap="The date of the first sighting of each species each year.
+The solid line connects the dates of the first survey each year.
+Note that many of the species are detected on the very first survey, thus eliminating any insight into shifts to earlier migratrations.
+THIS FIGURE NEEDS TO BE CUT INTO MULTIPLE SEPARATE FIGURES BASED ON FAMILIES.", warning=FALSE}
+# fig.asp=1.4, major candidate for cutting into multiple plots
 print(plot_timing_1)
@@
 
-```{r Fig-SurveyTiming1, echo=FALSE, eval=FALSE, cache=FALSE, fig.show= 'asis',fig.height=8, fig.width=12, fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
+```{r Fig-SurveyTiming1, echo=FALSE, eval=FALSE, cache=FALSE, fig.show= 'asis',fig.asp=1.4, fig.cap="The annual dates of the first and last survey have both varied by over a month across the years."}
 
 print(plot_alldates)
+```
+
+```{r SurveyEffort, eval=TRUE, cache = FALSE, include = FALSE}
+source("../scripts/survey_duration.R", echo=FALSE)
+```
+
+```{r Fig-SurveyEffort, echo=FALSE, eval=TRUE, cache=FALSE, fig.show= 'asis', fig.asp=1.4, fig.cap="The amount of time spent at each site each survey varied widely across each season, each year, and across sites.",warning= FALSE}
+# fig.asp=0.618,
+print(plot_duration)
 ```
@@
 Depending on the analysis, this might best be done by adjusting the response variable from 'counts' to 'counts per unit effort', weighting observations by survey effort, or considering how survey duration might be accounted for as a covariate.
-
-```{r SurveyEffort, eval=TRUE, cache = FALSE, include = FALSE}
-source("../scripts/survey_duration.R", echo=FALSE)
-```
-
-```{r Fig-SurveyEffort, echo=FALSE, eval=FALSE, cache=FALSE, fig.show= 'asis', fig.asp=0.618, fig.cap="The amount of time spent at each site each survey varied widely across each season, each year, and across sites.",warning= FALSE}
-
-print(plot_duration)
-```
 
@@
 
-```{r EDA_AllSpp_Frequency, eval=TRUE, cache = FALSE, include = FALSE, fig.height=8, fig.width=6}
+```{r EDA_AllSpp_Frequency, eval=TRUE, cache = FALSE, include = FALSE, warning=FALSE, echo=FALSE}
 
@@
 
-```{r Fig-AllSppFreq1, echo=FALSE, eval=TRUE, warning=FALSE, cache=FALSE, fig.show= 'asis', fig.height=20, fig.width=12, fig.cap="blah blah about consistency in target species occurrence.
-Maybe something about new spp showing up?"}
+```{r Fig-AllSppFreq1, echo=FALSE, eval=TRUE, warning=FALSE, cache=FALSE, fig.show= 'asis', fig.height=20, fig.width=12, fig.cap="The percentage of surveys each year during which a species was detected at any site.
+Lower percentages likely reflect migratory or rare species, higher frequencies (seasonal?) residents.
+[Ultimately review this for what it tells about which target species occur with adequate regularity to actually support monitoring change in the outcomes of interest.
+Maybe something about new spp showing up?]"}
 

### 2020-03-20 analysis/scripts/survey_duration.R
minor clean up and edits to overview

@@ -18,17 +18,17 @@
 #   ungroup()
 
 # group by year to get separate boxplot for each year
-plot_duration <-
-  ggplot(survey_duration %>%
-           # filter for negative (error) durations, NA Site IDs
-           filter(Duration >= 0, !is.na(GIS_Location_ID)),
-         aes(Year, Duration, group = Year)) +
+blah<-survey_duration %>%
+  # filter for negative (error) durations, NA Site IDs
+  filter(Duration > 0, !is.na(GIS_Location_ID))
+
+plot_duration <-ggplot(blah,aes(Year, Duration, group = Year)) +
   geom_boxplot() +
   # will result in vertical plots; flip
   coord_flip() +
   # JR: keeping x scale fixed to support easier comparison
   # across panels (sites).
-  facet_wrap(~GIS_Location_ID, scales = "fixed") +
+  facet_wrap(~GIS_Location_ID, scales = "fixed", ncol = 2) +
   # Resolve extreme observations by fixing scale limits at 180 min (3 hrs)
   # with vertical reference grid at 30 min intervals (rather than default 50)
   scale_y_continuous(limits=c(0,185),

### 2020-06-16 analysis/paper/CCRPmarkdownExample.rmd
Added file as an example for NPS-CCRP

--- 2020-03-20 analysis/paper/paper.Rmd
@@
 output:
-  bookdown::word_document2:
+  html_document:
+    df_print: paged
     fig_caption: yes
+    number_sections: yes
     toc: no
@@
     reference_docx: NRR_Author_Template_V4.2.docx
-  html_document:
-    df_print: paged
+  bookdown::word_document2:
     fig_caption: yes
-    number_sections: yes
     toc: no
@@
 Regional Biometrician pointed out that reducing the data collection effort requires further thought and familiarity with existing program's limitations and KLGO's 'hierarchy of objectives' for undertaking this monitoring effort and that, from a statistical perspective, skipping one year in an 15+ year monitoring effort was not problematic in terms of summary analyses of trends and changes.
-  + 2020 Feb/Mar (as time permitted) -
 
@@
 The timing should be clarified and consistently implemented, as much as possible, to reduce this source of uncertainty and problems in analyzing changes in phenology and other outcomes of interest.
-The clarification should be targeted as the priority information goals the Park has identified in the Problem Framing.
+The clarification should target the priority information goals the Park has identified in the Problem Framing.
 
@@
 
-```{r Fig-SpeciesTiming2, echo=FALSE, eval=TRUE, cache=TRUE, fig.show= 'asis', fig.height=25, fig.width=12, fig.cap="For each of the target species, the dates on which they were observed in any of the sample areas."}
+```{r Fig-SpeciesTiming2, echo=FALSE, eval=FALSE, cache=TRUE, fig.show= 'asis', fig.height=25, fig.width=12, fig.cap="For each of the target species, the dates on which they were observed in any of the sample areas."}
 

### 2020-06-23 analysis/paper/paper.Rmd
minor clean up and edits to overview
Testing - CCRP (AKD)
Making sure I have write access to this repo

@@
 ```
+#
